---
title: "zebra-force"
author: "Kyu Min Huh"
date: "2025-11-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Aims 

In this markdown, I aim to see the age effect on the leg force at take-off. 
Aims/Steps:

1. show morphological trend with age
- Graphs: a. body mass x age, b. wing area x age, c. wing chord length x age, d. tarsus length x age. 

2. See if there's any visible trend in raw data with age
- Graphs: a. max resultant force x age, with each individual marked by point and whiskers b. horizontal force at the time of max resultant force x age c. vertical force at the time of max resultant force x age d. impulse x age e. impulse ratio x age
f. take-off angle x age 

# Libraries
```{r libraries}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(ggeffects)

library(rptR)
library(ggpubr)
library(segmented)

library(nlme)
library(lme4)
library(lmerTest)
library(emmeans)
library(PerformanceAnalytics)
library(MuMIn)
library(lmtest)
library(lmerTest)
library(AICcmodavg)
library(devtools)
library(redres)
library(robustlmm)
library(car)
library(ggeffects)
library(gvlma)
library(performance)
library(sjPlot)
library(multcomp)
library(EnvStats)
```

# I. Raw data plots (mean+-SD)
## load the dataset 
### force_data
```{r load the datasets}

# force data
force_data <- read.delim(file = "C:\\Users\\jesu4036\\Documents\\flight-ontogeny\\data\\takeoff\\force_final.csv", sep = ",") %>% distinct()
#edited greenR-orangeL age40 Takeoff 3, 5 and pinkR-blueL 40 takeoff 1 to Takeoff 1 and 6. 

nas_filled <- read.delim(file = "C:\\Users\\jesu4036\\Documents\\flight-ontogeny\\data\\takeoff\\nas_filled.csv", sep = ",") %>% distinct()
nas_filled <- nas_filled[,-1]

dim(force_data)

force_data <- force_data %>%
  mutate(angle = as.numeric(angle)) %>%
  mutate(Fx_maxFtot = as.numeric(Fx_maxFtot)) %>%
  mutate(Fy_maxFtot = as.numeric(Fy_maxFtot)) %>%
  mutate(Fz_maxFtot = as.numeric(Fz_maxFtot)) 

nas <- force_data[is.na(force_data$Fx_maxFtot),]
nas

force_data <- force_data %>%
  filter(!FlightID %in% nas$FlightID)

dim(force_data)
```

```{r}
force_data <- rbind(force_data, nas_filled) # adding the added rows. 

dim(force_data)

# check if the nas are fixed ---> should return zero rows. 
force_data[is.na(force_data$Fx_maxFtot),]

force_data <- force_data %>%
  mutate(Fx_maxFtot_bw = abs(Fx_maxFtot)/bodyweight) %>%
  mutate(Fy_maxFtot_bw = abs(Fy_maxFtot)/bodyweight) %>%
  mutate(Fz_maxFtot_bw = abs(Fz_maxFtot)/bodyweight)
```

```{r}
# changing the unit to mN
# more suitable for birds that weigh in grams not kilograms
#force_data$max_Ftot = force_data$max_Ftot*1000

# newton to gram conversion
force_data$bodyweight_g = force_data$bodyweight * 101.9716

# make relative max res force
force_data$max_Ftot_bw = force_data$max_Ftot/force_data$bodyweight

# other relative force as well 





# 'Age': change 81->80 and 61->60 

force_data$Age[force_data$Age == 81] <- 80
force_data$Age[force_data$Age == 61] <- 60

#outlier that did not have flight data
force_data <- force_data %>%
  filter(!FlightID %in% c( 3702006, 2604005, 1404001, 2902805))
#3702006 outlier (0.8) in impulse that didn't have flight data
#2604005, 1404001, 2902805 => outliers in takeoff angle that don't have flight data and seemed to have taken off opposite directions. 
```

### wing area data
```{r}
wing_data <- read.delim(file = "C:\\Users\\jesu4036\\Documents\\flight-ontogeny\\data\\morphology\\wingdata_names_edited.csv", sep = ",") %>% distinct()

scale_weight <- read.delim(file = "C:\\Users\\jesu4036\\Documents\\flight-ontogeny\\data\\morphology\\bodyweight_measured.csv", sep = ",")


wing_data_sorted <- wing_data %>%
  # sort by Bird (alphabetical) and Age (increasing)
  arrange(Bird, Age) %>%
  # keep only birds with at least 3 age records
  group_by(Bird) %>%
  filter(n_distinct(Age) >= 3) %>%
  mutate(wingarea = wingarea/10000) %>% #converting mm^2 -> m^2
  ungroup() %>%
  left_join(scale_weight, by = c("Bird", "Age")) %>%
  mutate(bodyweight_measured = as.numeric(bodyweight_measured)) %>%
  mutate(bodyweight_measured = bodyweight_measured * 0.00980665) %>%
  mutate(wing_loading = wingarea/bodyweight_measured) %>%
  dplyr::select('Bird','Age','wingspan.calc','wingarea','bodyweight_measured','wing_loading')

dim(wing_data_sorted)

wing_data_final <- wing_data_sorted %>%
  group_by(Bird, Age) %>%
  summarise(
    wingspan = mean(wingspan.calc, na.rm = TRUE),
    wingarea = mean(wingarea, na.rm = TRUE),
    scale_weight = mean(bodyweight_measured, na.rm = TRUE),
    wing_loading = mean(wing_loading, na.rm = TRUE),
    .groups = "drop"
  ) %>% drop_na()

dim(wing_data_final) #192

```

### sex and brood measured bodyweight data 
```{r}
sex_data <- read.delim(file = "C:\\Users\\jesu4036\\Documents\\flight-ontogeny\\data\\morphology\\birds_sex_brood.csv", sep = ",") %>% distinct()

force_data <- merge(force_data, sex_data, by = "Bird")
wing_data_final <- merge(wing_data_final, sex_data, by = "Bird")


```

### flight trajectory data
```{r}
flights <- read.delim(file = "C:\\Users\\jesu4036\\Documents\\flight-ontogeny\\data\\takeoff\\flight_outcome_full.csv", sep = ",") %>% left_join(sex_data, by = "Bird") 

raw_flight <- read.delim(file = "C:\\Users\\jesu4036\\Documents\\flight-ontogeny\\data\\takeoff\\clean_filtered_flights.csv", sep =",")
colnames(flights)
colnames(raw_flight)
```
## one flight
```{r}
#cyanR-blackL, Age 36, trial 1
#test_flight1 <- raw_flight %>%
 #filter(FlightID == 2303601.0)

test_flight <- raw_flight %>%
  filter(FlightID == 2303601.0)
```
## RAW PLOTS
```{r}

library(ggplot2)
library(patchwork)
df <- test_flight

# Position plots
p1 <- ggplot(df, aes(cumul_time, X)) +
  geom_line(color = "blue") +
  geom_point(color = "darkblue") +
  ggtitle("Position X")

p2 <- ggplot(df, aes(cumul_time, Y)) +
  geom_line(color = "green") +
  geom_point(color = "darkgreen") +
  ggtitle("Position Y")

p3 <- ggplot(df, aes(cumul_time, Z)) +
  geom_line(color = "red") +
  geom_point(color = "darkred") +
  ggtitle("Position Z")

# Velocity plots
p4 <- ggplot(df, aes(cumul_time, inst_V_X)) +
  geom_line(color = "blue") +
  geom_point(color = "darkblue") +
  ggtitle("Velocity X")

p5 <- ggplot(df, aes(cumul_time, inst_V_Y)) +
  geom_line(color = "green") +
  geom_point(color = "darkgreen") +
  ggtitle("Velocity Y")

p6 <- ggplot(df, aes(cumul_time, inst_V_Z)) +
  geom_line(color = "red") +
  geom_point(color = "darkred") +
  ggtitle("Velocity Z")

# Acceleration plots
p7 <- ggplot(df, aes(cumul_time, inst_acc_X)) +
  geom_line(color = "blue") +
  geom_point(color = "darkblue") +
  ggtitle("Acceleration X")

p8 <- ggplot(df, aes(cumul_time, inst_acc_Y)) +
  geom_line(color = "green") +
  geom_point(color = "darkgreen") +
  ggtitle("Acceleration Y")

p9 <- ggplot(df, aes(cumul_time, inst_acc_Z)) +
  geom_line(color = "red") +
  geom_point(color = "darkred") +
  ggtitle("Acceleration Z")


# New 5 plots
p10 <- ggplot(df, aes(cumul_time, inst_V_res)) +
  geom_line(color="purple") + geom_point(color="purple") + ggtitle("Velocity Resultant")

p11 <- ggplot(df, aes(cumul_time, inst_acc_res)) +
  geom_line(color="orange") + geom_point(color="darkorange") + ggtitle("Tangential acceleration")

p12 <- ggplot(df, aes(cumul_time, inst_acc_mag)) +
  geom_line(color="orange") + geom_point(color="darkorange") + ggtitle("Acceleration magnitude")


p13 <- ggplot(df, aes(cumul_time, Turn_Rate_degrees)) +
  geom_line(color="brown") + geom_point(color="brown") + ggtitle("Turn Rate (degrees)")

p14 <- ggplot(df, aes(cumul_time, Centripetal_acc)) +
  geom_line(color="pink") + geom_point(color="deeppink") + ggtitle("Centripetal Acceleration")

# Combine all plots
final_plot2 <- (p1 | p2 | p3) /
              (p4 | p5 | p6) /
              (p7 | p8 | p9) /
              (p10 | p11 | p12)/
              (p13| p14)

final_plot2




```
#### function plot_age_var 
```{r}
plot_age_var <- function(variable, df, title) {
  
  # Summary by individual
  summary_df <- df %>%
    group_by(Bird, Age) %>%
    summarise(
      mean = mean({{variable}}, na.rm = TRUE),
      sd = sd({{variable}}, na.rm = TRUE),
      n = n(),
      .groups = "drop"
    )
  
  # Summary by age
  trend_df <- df %>%
    group_by(Age) %>%
    summarise(
      mean = mean({{variable}}, na.rm = TRUE),
      sd = sd({{variable}}, na.rm = TRUE),
      n = n(),
      .groups = "drop"
    )
  
  # Plot
  plot_age_performance<- ggplot(df, aes(x = Age, y = {{variable}}, color = Bird)) +
    geom_point(position = position_jitter(width = 1), alpha = 0.3) +
    geom_errorbar(
      data = summary_df,
      aes(x = Age, ymin = mean - sd, ymax = mean + sd, color = Bird, alpha = 0.3),
      width = 2, size = 0.8,
      inherit.aes = FALSE
    ) +
    geom_point(
      data = summary_df,
      aes(y = mean, x = Age, color = Bird, alpha = 0.9),
      size = 2, shape = 18,
      inherit.aes = FALSE
    ) +
    geom_line(
      data = summary_df,
      aes(x = Age, y = mean, color = Bird, group = Bird, alpha = 0.8),
      size = 0.7,
      inherit.aes = FALSE
    ) +
    geom_ribbon(
      data = trend_df,
      aes(x = Age, ymin = mean - sd, ymax = mean + sd),
      fill = "grey70", alpha = 0.5,
      inherit.aes = FALSE
    ) +
    geom_line(
      data = trend_df,
      aes(x = Age, y = mean),
      ,color = "grey50", alpha = 0.8, size = 1.2,
      inherit.aes = FALSE
    ) +
    scale_x_continuous(breaks = unique(df$Age)) +
    labs(x = "Age", y = title, title = "") +
    guides(alpha = "none", color = "none") +
    theme(legend.position = "none")+
    theme_minimal()
  
  return(plot_age_performance)
}
```
## A. Morphology
### a. wing area

```{r wing area per bird}

# Summarize mean and standard deviation per Bird & Age
summary_df <- wing_data_final %>%
  group_by(Bird, Age) %>%
  summarise(
    mean_wa = mean(wingarea),
    sd_wa = sd(wingarea),
    .groups = "drop"
  )

trend_df <- wing_data_final %>%
  group_by(Age) %>%
  summarise(
    mean_wa = mean(wingarea, na.rm = TRUE),
    sd_wa   = sd(wingarea, na.rm = TRUE),
    .groups = "drop"
  )

# Plot: trials + mean + SD
plot_wingarea <- ggplot(wing_data_final, aes(x = Age, y = wingarea, color = Bird)) +
  geom_point(position = position_jitter(width = 1), alpha = 0.3) +   # dots = trials
  geom_errorbar(
    data = summary_df,
    aes(x=Age, ymin = mean_wa - sd_wa, ymax = mean_wa + sd_wa, color = Bird, alpha = 0.6),
    width = 2, size = 0.8,
    inherit.aes = FALSE) +
  
  geom_point(
    data = summary_df,
    aes(x = Age, y = mean_wa,  color = Bird, alpha = 0.9),
    size = 3, shape = 18,
    inherit.aes = FALSE
  ) +
  
   geom_line(
    data = summary_df,
    aes(x = Age, y = mean_wa, color = Bird, group = Bird, alpha = 0.8),
    linewidth = 0.7,
    inherit.aes = FALSE) +
  
  geom_ribbon(
    data = trend_df,
    aes(x = Age, ymin = mean_wa - sd_wa, ymax = mean_wa + sd_wa),
    fill = "grey70", alpha = 0.5,
    inherit.aes = FALSE
  ) +
  
  geom_line(
    data = trend_df,
    aes(x = Age, y = mean_wa),
    color = "grey50", alpha = 0.8, size = 1.2,
    inherit.aes = FALSE) +
  
   scale_x_continuous(breaks = unique(wing_data_sorted$Age)) +
  labs(x = "Age", y = "Wing Area (m^2)", title = "") +
  ylim(0.5, 1.0)+
  guides(alpha = "none") +
  theme_minimal() #+
  #theme(legend.position = "none")

plot_wingarea



```
```{r}
plot_wingarea <- plot_age_var(wingarea, wing_data_final, "wing area (N/m^2)")
plot_wingarea
```


### b. calculated wingspan
```{r}
plot_wingspan <- plot_age_var(wingspan, wing_data_final, "wingspan (mm)")
plot_wingspan
```

### c. perching weight
```{r}
plot_perch_g <- plot_age_var(bodyweight_g, force_data, "perching weight(g)")
```

### d. scale weight
```{r}
plot_scale_g <- plot_age_var(scale_weight*101.97, wing_data_final, "scale weight (g)")
ggarrange(plot_perch_g, plot_scale_g, nrow = 1, ncol = 2)
```

### e. wing loading 
```{r}
plot_wingloading <- plot_age_var(wing_loading, wing_data_final, "wing loading (N/m^2)")
```

### PLOTS - wingspan, wing area, scale weight, wing loading 
```{r}

plots_morphology <- ggarrange(plot_wingspan, plot_wingarea, plot_scale_g, plot_wingloading, nrow = 2, ncol =2)

plots_morphology

```
## B. Performance

### LEGS
#### a. max resultant force
```{r}

summary_df <- force_data %>%
  group_by(Bird, Age) %>%
  summarise(
    mean_maxF = mean(max_Ftot),
    sd_maxF   = sd(max_Ftot),
    .groups = "drop"
  ) %>% as.data.frame()

# Reference means at Age 20 for each Bird
ref_summary <- summary_df %>%
  filter(Age == 20) %>%
  dplyr::select(Bird, mean_maxF, sd_maxF) %>%
  dplyr::rename(ref_mean_maxF = mean_maxF, 
         ref_sd_maxF = sd_maxF)

# Centered summary: subtract the bird-specific mean at Age 20
summary_df_init <- summary_df %>%
  left_join(ref_summary, by = "Bird") %>%
  mutate(
    mean_maxF_centered = mean_maxF - ref_mean_maxF,
    # Optionally keep sd as-is; centering doesn't change variability
    sd_maxF_centered   = sd_maxF #SD unchanged
  ) %>%
  arrange(Bird, Age)


trend_df <- force_data %>%
  group_by(Age) %>%
  summarise(
    mean_maxF = mean(max_Ftot, na.rm = TRUE),
    sd_maxF   = sd(max_Ftot, na.rm = TRUE),
    .groups = "drop"
  )


# Extract reference mean at Age 20
ref_mean <- trend_df %>%
  filter(Age == 20) %>%
  pull(mean_maxF)

# Center all means by subtracting the mean at Age 20
trend_df_init <- trend_df %>%
  mutate(
    mean_maxF_centered = mean_maxF - ref_mean,
    sd_maxF_centered   = sd_maxF  # SD stays the same
  )
```
##### relative
```{r}
plot_raw_Ftot <- plot_age_var(max_Ftot_bw, force_data, "force magnitude relative to perching weight")
plot_raw_Ftot
```
##### raw
```{r}
plot_age_maxFtot <- ggplot(force_data, aes(x = Age, y = max_Ftot, color = Bird)) +
  geom_point(position = position_jitter(width = 1), alpha = 0.3) +   # dots = trials
  geom_errorbar(
    data = summary_df,
    aes(x=Age, ymin = mean_maxF - sd_maxF, ymax = mean_maxF + sd_maxF, color = Bird, alpha = 0.6),
    width = 2, size = 0.8,
    inherit.aes = FALSE) +
  
  geom_point(
    data = summary_df,
    aes(y = mean_maxF, x = Age, color = Bird, alpha = 0.9),
    size = 3, shape = 18,
    inherit.aes = FALSE
  ) +
  
   geom_line(
    data = summary_df,
    aes(x = Age, y = mean_maxF, color = Bird, group = Bird, alpha = 0.8),
    size = 0.7,
    inherit.aes = FALSE) +
  
  geom_ribbon(
    data = trend_df,
    aes(x = Age, ymin = mean_maxF - sd_maxF, ymax = mean_maxF + sd_maxF),
    fill = "grey70", alpha = 0.5,
    inherit.aes = FALSE
  ) +
  
  geom_line(
    data = trend_df,
    aes(x = Age, y = mean_maxF),
    color = "grey50", alpha = 0.8, size = 1.2,
    inherit.aes = FALSE) +
  
   scale_x_continuous(breaks = unique(force_data$Age)) +
  labs(x = "Age", y = "max F (N)", title = "") +
  #ylim(10, 25)+
  guides(alpha = "none")+
  theme_minimal() + 
  theme(legend.position = "none")

plot_age_maxFtot
```
##### centrerd
```{r}
plot_age_maxFtot_centred <- ggplot() +
  
  # error bard for per-bird summary
  geom_errorbar(
    data = summary_df_init,
    aes(x = Age, 
        ymin = mean_maxF_centered - sd_maxF_centered, 
        ymax = mean_maxF_centered + sd_maxF_centered, 
        color = Bird, alpha = 0.6),
    width = 2, size = 0.8,
    inherit.aes = FALSE) +
  
  # Points for per-bird means
  geom_point(
    data = summary_df_init,
    aes(y = mean_maxF_centered, x = Age, color = Bird, alpha = 0.9),
    size = 3, shape = 18,
    inherit.aes = FALSE
  ) +
  # Lines for per-bird trends
   geom_line(
    data = summary_df_init,
    aes(x = Age, y = mean_maxF_centered, color = Bird, group = Bird, alpha = 0.8),
    size = 0.7,
    inherit.aes = FALSE) +
  
  # Ribbon for overall trend SD
  geom_ribbon(
    data = trend_df_init,
    aes(x = Age, ymin = mean_maxF_centered - sd_maxF_centered, ymax = mean_maxF_centered + sd_maxF_centered),
    fill = "grey70", alpha = 0.5,
    inherit.aes = FALSE
  ) +
  
  # Line for overall trend mean
  geom_line(
    data = trend_df_init,
    aes(x = Age, y = mean_maxF_centered),
    color = "grey50", alpha = 0.8, size = 1.2,
    inherit.aes = FALSE) +
  
   scale_x_continuous(breaks = unique(force_data$Age)) +
  labs(x = "Age", y = "max F centred (N)", title = "") +
  #ylim(10, 25)+
  guides(alpha = "none")+
  theme_minimal() + 
  theme(legend.position = "none")

plot_age_maxFtot_centred




```

#### b. bodyweight relative max Ftot
```{r}

summary_df <- force_data %>%
  group_by(Bird, Age) %>%
  summarise(
    mean_maxF_bw = mean(max_Ftot_bw),
    sd_maxF_bw   = sd(max_Ftot_bw),
    .groups = "drop"
  )

trend_df <- force_data %>%
  group_by(Age) %>%
  summarise(
    mean_maxF_bw = mean(max_Ftot_bw, na.rm = TRUE),
    sd_maxF_bw   = sd(max_Ftot_bw, na.rm = TRUE),
    .groups = "drop"
  )


plot_age_maxF_bw <- ggplot(force_data, aes(x = Age, y = max_Ftot_bw, color = Bird)) +
  geom_point(position = position_jitter(width = 1), alpha = 0.3) +   # dots = trials
  geom_errorbar(
    data = summary_df,
    aes(x=Age, ymin = mean_maxF_bw - sd_maxF_bw, ymax = mean_maxF_bw + sd_maxF_bw, color = Bird, alpha = 0.6),
    width = 2, size = 0.8,
    inherit.aes = FALSE) +
  
  geom_point(
    data = summary_df,
    aes(y = mean_maxF_bw, x = Age, color = Bird, alpha = 0.9),
    size = 3, shape = 18,
    inherit.aes = FALSE
  ) +
  
   geom_line(
    data = summary_df,
    aes(x = Age, y = mean_maxF_bw, color = Bird, group = Bird, alpha = 0.8),
    size = 0.7,
    inherit.aes = FALSE) +
  
  geom_ribbon(
    data = trend_df,
    aes(x = Age, ymin = mean_maxF_bw - sd_maxF_bw, ymax = mean_maxF_bw + sd_maxF_bw),
    fill = "grey70", alpha = 0.5,
    inherit.aes = FALSE
  ) +
  
  geom_line(
    data = trend_df,
    aes(x = Age, y = mean_maxF_bw),
    color = "grey50", alpha = 0.8, size = 1.2,
    inherit.aes = FALSE) +
  
   scale_x_continuous(breaks = unique(force_data$Age)) +
  labs(x = "Age", y = "max F / bw", title = "") +
  #ylim(10, 25)+
  guides(alpha = "none")+
  theme_minimal()+
  theme(legend.position = "none")

plot_age_maxF_bw
```
#### c. vertical force
##### absolute
```{r}

summary_df <- force_data %>%
  group_by(Bird, Age) %>%
  summarise(
    mean_Fx = mean(-Fx_maxFtot),
    sd_Fx = sd(-Fx_maxFtot),
    .groups = "drop"
  )

trend_df <- force_data %>%
  group_by(Age) %>%
  summarise(
    mean_Fx = mean(-Fx_maxFtot, na.rm = TRUE),
    sd_Fx   = sd(-Fx_maxFtot, na.rm = TRUE),
    .groups = "drop"
  )


plot_age_Fx <- ggplot(force_data, aes(x = Age, y = -Fx_maxFtot, color = Bird)) +
  geom_point(position = position_jitter(width = 1), alpha = 0.3) +   # dots = trials
  geom_errorbar(
    data = summary_df,
    aes(x=Age, ymin = mean_Fx - sd_Fx, ymax = mean_Fx + sd_Fx, color = Bird, alpha = 0.6),
    width = 2, size = 0.8,
    inherit.aes = FALSE) +
  
  geom_point(
    data = summary_df,
    aes(y = mean_Fx, x = Age, color = Bird, alpha = 0.9),
    size = 2, shape = 18,
    inherit.aes = FALSE
  ) +
  
   geom_line(
    data = summary_df,
    aes(x = Age, y = mean_Fx, color = Bird, group = Bird, alpha = 0.8),
    size = 0.7,
    inherit.aes = FALSE) +
  
  geom_ribbon(
    data = trend_df,
    aes(x = Age, ymin = mean_Fx - sd_Fx, ymax = mean_Fx + sd_Fx),
    fill = "grey70", alpha = 0.5,
    inherit.aes = FALSE
  ) +
  
  geom_line(
    data = trend_df,
    aes(x = Age, y = mean_Fx),
    color = "grey50", alpha = 0.8, size = 1.2,
    inherit.aes = FALSE) +
  
   scale_x_continuous(breaks = unique(force_data$Age)) +
  labs(x = "Age", y = "Fx", title = "") +
  #ylim(10, 25)+
  guides(alpha = "none")+
  theme_minimal() +
  theme(legend.position = "none")

plot_age_Fx


```
##### relative
```{r}

summary_df <- force_data %>%
  group_by(Bird, Age) %>%
  summarise(
    mean_Fx_bw = mean(Fx_maxFtot_bw),
    sd_Fx_bw = sd(Fx_maxFtot_bw),
    .groups = "drop"
  )

trend_df <- force_data %>%
  group_by(Age) %>%
  summarise(
    mean_Fx_bw = mean(Fx_maxFtot_bw, na.rm = TRUE),
    sd_Fx_bw   = sd(Fx_maxFtot_bw, na.rm = TRUE),
    .groups = "drop"
  )


plot_age_Fx_bw <- ggplot(force_data, aes(x = Age, y = Fx_maxFtot_bw, color = Bird)) +
  geom_point(position = position_jitter(width = 1), alpha = 0.3) +   # dots = trials
  geom_errorbar(
    data = summary_df,
    aes(x=Age, ymin = mean_Fx_bw - sd_Fx_bw, ymax = mean_Fx_bw + sd_Fx_bw, color = Bird, alpha = 0.6),
    width = 2, size = 0.8,
    inherit.aes = FALSE) +
  
  geom_point(
    data = summary_df,
    aes(y = mean_Fx_bw, x = Age, color = Bird, alpha = 0.9),
    size = 2, shape = 18,
    inherit.aes = FALSE
  ) +
  
   geom_line(
    data = summary_df,
    aes(x = Age, y = mean_Fx_bw, color = Bird, group = Bird, alpha = 0.8),
    size = 0.7,
    inherit.aes = FALSE) +
  
  geom_ribbon(
    data = trend_df,
    aes(x = Age, ymin = mean_Fx_bw - sd_Fx_bw, ymax = mean_Fx_bw + sd_Fx_bw),
    fill = "grey70", alpha = 0.5,
    inherit.aes = FALSE
  ) +
  
  geom_line(
    data = trend_df,
    aes(x = Age, y = mean_Fx_bw),
    color = "grey50", alpha = 0.8, size = 1.2,
    inherit.aes = FALSE) +
  
   scale_x_continuous(breaks = unique(force_data$Age)) +
  labs(x = "Age", y = "Fx / bw (N)", title = "") +
  #ylim(10, 25)+
  guides(alpha = "none")+
  theme_minimal() +
  theme(legend.position = "none")

plot_age_Fx_bw

```
#### d. horizontal force
##### absolute
```{r}
summary_df <- force_data %>%
  group_by(Bird, Age) %>%
  summarise(
    mean_Fy = mean(Fy_maxFtot),
    sd_Fy = sd(Fy_maxFtot),
    .groups = "drop"
  )

trend_df <- force_data %>%
  group_by(Age) %>%
  summarise(
    mean_Fy = mean(Fy_maxFtot, na.rm = TRUE),
    sd_Fy   = sd(Fy_maxFtot, na.rm = TRUE),
    .groups = "drop"
  )


plot_age_Fy <- ggplot(force_data, aes(x = Age, y = Fy_maxFtot, color = Bird)) +
  geom_point(position = position_jitter(width = 1), alpha = 0.3) +   # dots = trials
  geom_errorbar(
    data = summary_df,
    aes(x=Age, ymin = mean_Fy - sd_Fy, ymax = mean_Fy + sd_Fy, color = Bird, alpha = 0.6),
    width = 2, size = 0.8,
    inherit.aes = FALSE) +
  
  geom_point(
    data = summary_df,
    aes(y = mean_Fy, x = Age, color = Bird, alpha = 0.9),
    size = 2, shape = 18,
    inherit.aes = FALSE
  ) +
  
   geom_line(
    data = summary_df,
    aes(x = Age, y = mean_Fy, color = Bird, group = Bird, alpha = 0.8),
    size = 0.7,
    inherit.aes = FALSE) +
  
  geom_ribbon(
    data = trend_df,
    aes(x = Age, ymin = mean_Fy - sd_Fy, ymax = mean_Fy + sd_Fy),
    fill = "grey70", alpha = 0.5,
    inherit.aes = FALSE
  ) +
  
  geom_line(
    data = trend_df,
    aes(x = Age, y = mean_Fy),
    color = "grey50", alpha = 0.8,size = 1.2,
    inherit.aes = FALSE) +
  
   scale_x_continuous(breaks = unique(force_data$Age)) +
  labs(x = "Age", y = "Fy", title = "") +
  #ylim(10, 25)+
  guides(alpha = "none")+
  theme_minimal() +
  theme(legend.position = "none")

plot_age_Fy

```

##### relative
```{r}
summary_df <- force_data %>%
  group_by(Bird, Age) %>%
  summarise(
    mean_Fy_bw = mean(Fy_maxFtot_bw),
    sd_Fy_bw = sd(Fy_maxFtot_bw),
    .groups = "drop"
  )

trend_df <- force_data %>%
  group_by(Age) %>%
  summarise(
    mean_Fy_bw = mean(Fy_maxFtot_bw, na.rm = TRUE),
    sd_Fy_bw   = sd(Fy_maxFtot_bw, na.rm = TRUE),
    .groups = "drop"
  )


plot_age_Fy_bw <- ggplot(force_data, aes(x = Age, y = Fy_maxFtot_bw, color = Bird)) +
  geom_point(position = position_jitter(width = 1), alpha = 0.3) +   # dots = trials
  geom_errorbar(
    data = summary_df,
    aes(x=Age, ymin = mean_Fy_bw - sd_Fy_bw, ymax = mean_Fy_bw + sd_Fy_bw, color = Bird, alpha = 0.6),
    width = 2, size = 0.8,
    inherit.aes = FALSE) +
  
  geom_point(
    data = summary_df,
    aes(y = mean_Fy_bw, x = Age, color = Bird, alpha = 0.9),
    size = 2, shape = 18,
    inherit.aes = FALSE
  ) +
  
   geom_line(
    data = summary_df,
    aes(x = Age, y = mean_Fy_bw, color = Bird, group = Bird, alpha = 0.8),
    size = 0.7,
    inherit.aes = FALSE) +
  
  geom_ribbon(
    data = trend_df,
    aes(x = Age, ymin = mean_Fy_bw - sd_Fy_bw, ymax = mean_Fy_bw + sd_Fy_bw),
    fill = "grey70", alpha = 0.5,
    inherit.aes = FALSE
  ) +
  
  geom_line(
    data = trend_df,
    aes(x = Age, y = mean_Fy_bw),
    color = "grey50", alpha = 0.8, size = 1.2,
    inherit.aes = FALSE) +
  
   scale_x_continuous(breaks = unique(force_data$Age)) +
  labs(x = "Age", y = "Fy / bw (N)", title = "") +
  #ylim(10, 25)+
  guides(alpha = "none")+
  ylim(0,6) +
  theme_minimal() +
  theme(legend.position = "none")

plot_age_Fy_bw


```
#### e. longitudinal force
##### absolute
```{r}

plot_age_Fz <- plot_age_var(Fz_maxFtot, force_data, "max longitudinal force (N)")
```
##### relative
```{r}
plot_age_Fx_bw <- plot_age_var(Fz_maxFtot_bw, force_data, "Fx/bw (N)")
plot_age_Fx_bw

```


#### f. impulse
##### impulse 
```{r}

plot_age_impulse <- plot_age_var(impulse*1000, force_data, "impulse (mN*s)") +ylim(-10, 60) #1 row removed

plot_age_impulse
```

##### duration
```{r}
plot_age_duration <- plot_age_var(duration*1000, force_data, "impulse duration (ms)") 

plot_age_duration
```

##### impulse ratio
```{r}
plot_age_impulse_ratio <- plot_age_var(impulse_ratio, force_data, "impulse ratio ( left impulse / right impulse )")

plot_age_impulse_ratio
```


##### left impulse
```{r}
plot_left_impulse <- plot_age_var(left_impulse*1000, force_data, "left impulse (mN*s)") #1 outlier removed
plot_left_impulse
```

##### right impulse
```{r}
plot_right_impulse <- plot_age_var(right_impulse*1000, force_data, "right impulse (mN*s)") 
plot_right_impulse
```

##### left duration
```{r}
plot_left_duration <- plot_age_var(left_duration*1000, force_data, "left duration (ms)")
plot_left_duration
```

##### right duration
```{r}

plot_right_duration <- plot_age_var(right_duration*1000, force_data, "right duration (ms)") #+ ylim(0, 0.075)

plot_right_duration

```


#### f. takeoff angle 
```{r}

plot_age_angle <- plot_age_var(angle, force_data, "takeoff angle (degrees)")
plot_age_angle 
```


### PLOTS. altogether
```{r}
ggarrange(plot_age_maxFtot, plot_age_Fx, plot_age_Fy, plot_age_maxF_bw, plot_age_Fx_bw, plot_age_Fy_bw, nrow = 2, ncol=3)
```
### OUTLIERS detection

```{r}
test_Ftot <- rosnerTest(force_data$max_Ftot_bw, k= 10)
test_Ftot$all.stats #no outliers

test_Fx <- rosnerTest(force_data$Fx_maxFtot_bw, k = 10)
test_Fx$all.stats #no outliers

test_Fy <- rosnerTest(force_data$Fy_maxFtot_bw, k = 10)
test_Fy$all.stats

test_Fz <- rosnerTest(force_data$Fz_maxFtot_bw, k = 10)
test_Fz$all.stats
```

```{r}
test_impulse <- rosnerTest(force_data$impulse, k = 10)
test_impulse$all.stats

test_duration <- rosnerTest(force_data$duration, k = 20)
test_duration$all.stats

test_angle <- rosnerTest(force_data$angle, k = 10)
test_angle$all.stats
```

```{r}

test_impulse_ratio <- rosnerTest(force_data$impulse, k = 10)
test_impulse_ratio$all.stats

test_left_impulse <- rosnerTest(force_data$left_impulse, k= 10)
test_left_impulse$all.stats

test_right_impulse <- rosnerTest(force_data$right_impulse, k = 10)
test_right_impulse$all.stats

test_left_duration <- rosnerTest(force_data$left_duration, k = 20)
test_left_duration$all.stats

test_right_duration <- rosnerTest(force_data$right_duration, k = 10)
test_right_duration$all.stats

```
```{r}

out_impulse <- test_impulse$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)


out_duration <- test_duration$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)


out_angle <- test_angle$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)


out_impulse_ratio <- test_impulse_ratio$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)


out_left_impulse <- test_left_impulse$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)


out_left_duration <- test_left_duration$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)


out_right_duration <- test_right_duration$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)


outliers_force <-rbind(out_impulse, out_duration, out_angle, out_impulse_ratio, out_left_impulse, out_left_duration, out_right_duration)

dim(outliers_force) 
 #41 outliers

```
```{r}
dim(force_data) #988

force_clean <- force_data[-c(outliers_force$Obs.Num),]

dim(force_clean) #961
```

### FLIGHT SPEED

#### a. turn rate
##### mean
```{r}
plot_tr_mean <- plot_age_var(mean_turn_rate, flights, "mean turn rate (degrees/s)") +ylim(0,750)
plot_tr_mean

```

##### max
```{r}

plot_tr_max <- plot_age_var(max_turn_rate, flights, "max turn rate (degrees/s)") +ylim(-1250, 5000)
plot_tr_max
```


#### b. centripetal acceleration
##### mean
```{r}

plot_ctracc_mean <- plot_age_var(mean_centri_acc, flights, "mean centripetal acceleration (m/s^2)") +ylim(0, 30)
plot_ctracc_mean
```

##### max
```{r}

plot_ctracc_max <- plot_age_var(max_centri_acc, flights, "max centripetal acceleration (m/s^2)") +ylim(-50, 150)
plot_ctracc_max

```
#### c. resultant velocity
##### mean
```{r}
plot_vres_mean <- plot_age_var(mean_v_res, flights, "mean resultant velocity (m/s)")
plot_vres_mean
```
##### max
```{r}
plot_vres_max <- plot_age_var(variable = max_v_res, df = flights, title = "max resultant velocity (m/s)")
plot_vres_max

```

#### d. resultant acceleration
##### mean
```{r}

plot_accres_mean <- plot_age_var(mean_acc_res, flights, "mean resultant acceleration (m/s^2)")
plot_accres_mean
```
##### max
```{r}
plot_accres_max <- plot_age_var(max_acc_res, flights, "max resultant acceleration (m/s^2)") +ylim(-100, 150)
plot_accres_max

```
#### e. X vel & acc
##### mean v
```{r}
plot_vx_mean <- plot_age_var(mean_vx, flights, "mean X velocity (m/s)")
plot_vx_mean
```
##### max v
```{r}
plot_vx_max <- plot_age_var(max_vx, flights, "max X velocity (m/s)")
plot_vx_max
```
##### mean acc
```{r}
plot_accx_mean <- plot_age_var(mean_acc_x, flights, "mean X acceleration (m/s^2)") 
plot_accx_mean

```
##### max acc
```{r}
plot_accx_max <- plot_age_var(max_acc_x, flights, "max X acceleration (m/s^2)") 
plot_accx_max

```
#### f. Y vel & acc
##### mean v
```{r}
plot_vy_mean <- plot_age_var(mean_vy, flights, "mean Y velocity (m/s)") 
plot_vy_mean
```
##### max v
```{r}
plot_vy_max <- plot_age_var(max_vy, flights, "max Y velocity (m/s)")
plot_vy_max +ylim(0, 6)
```
##### mean acc
```{r}
plot_accy_mean <- plot_age_var(mean_acc_y, flights, "mean Y acceleration (m/s^2)") 
plot_accy_mean
```
##### max acc
```{r}
plot_accy_max <- plot_age_var(mean_acc_y, flights, "max Y acceleration (m/s^2)") 
plot_accy_max
```
#### g. Z vel & acc
##### mean v
```{r}
plot_vz_mean <- plot_age_var(mean_vz, flights, "mean Z velocity (m/s)")
plot_vz_mean
```
##### max v
```{r}
plot_vz_max <- plot_age_var(max_vz, flights, "max Z velocity (m/s)")
plot_vz_max

```
##### mean acc
```{r}
plot_accz_mean <- plot_age_var(mean_acc_z, flights, "mean Z acceleration (m/s^2)")
plot_accz_mean

```
##### max acc
```{r}
plot_accz_max <- plot_age_var(max_acc_z, flights, "max Z acceleration (m/s^2)") 
plot_accz_max
```
### WINGBEATS
#### frequency
```{r}
plot_wb_mean <- plot_age_var(wingbeat_frequency, flights, "wingbeat frequency (Hz)")
plot_wb_mean
```
#### upstroke
```{r}
plot_upstroke_acc <- plot_age_var(upstroke_amplitude_acc, flights, "upstroke_head_acc (m/s^2)")
plot_upstroke_acc

plot_upstroke_mm <- plot_age_var(1000*upstroke_amplitude_m, flights_wing, "upstroke head displacement (m)")
plot_upstroke_mm
```
#### downstroke
```{r}
plot_downstroke_acc <- plot_age_var(downstroke_amplitude_acc, flights, "downstroke_head_acc (m/s^2)")
plot_downstroke_acc

plot_downstroke_mm <- plot_age_var(1000*downstroke_amplitude_m, flights_wing, "downstroke head displacement (m)")
plot_downstroke_mm
```

### ggarrange Plots
```{r}
ggarrange(plot_vx_mean, plot_vy_mean, plot_vz_mean, 
          plot_accx_mean, plot_accy_mean, plot_accz_mean, nrow = 2, ncol = 3)

ggarrange(plot_vx_max, plot_vy_max, plot_vz_max,
          plot_accx_max, plot_accy_max, plot_accz_max, nrow = 2, ncol = 3)

ggarrange(plot_vres_mean, plot_accres_mean, plot_tr_mean, plot_ctracc_mean, nrow = 2, ncol =2)

ggarrange(plot_vres_max, plot_accres_max, plot_tr_max, plot_ctracc_max, nrow = 2, ncol =2 )

ggarrange(plot_wb_mean, plot_downstroke_m, plot_upstroke_m, nrow = 1, ncol = 3)
```

### OUTLIERS detection
```{r}
test_vx <- rosnerTest(flights$mean_vx, k= 10)
test_vx$all.stats #no outliers

test_vy <- rosnerTest(flights$mean_vy, k = 10)
test_vy$all.stats #no outliers

test_vz <- rosnerTest(flights$mean_vz, k = 10)
test_vz$all.stats

test_vres <- rosnerTest(flights$mean_v_res, k = 10)
test_vres$all.stats
```

```{r}
test_accx <- rosnerTest(flights$mean_acc_x, k = 10)
test_accx$all.stats

test_accy <- rosnerTest(flights$mean_acc_y, k = 10)
test_accy$all.stats

test_accz <- rosnerTest(flights$mean_acc_z, k = 10)
test_accz$all.stats

test_accres <- rosnerTest(flights$mean_acc_res, k = 10) #tangential acceleration
test_accres$all.stats
```
```{r}
test_tr <- rosnerTest(flights$mean_turn_rate, k = 20)
test_tr$all.stats

test_centri <- rosnerTest(flights$mean_centri_acc, k = 10)
test_centri$all.stats

test_wingbeat <- rosnerTest(flights$wingbeat_frequency, k = 10)
test_wingbeat$all.stats

test_upstroke <- rosnerTest(flights$upstroke_amplitude_m, k = 30)
test_upstroke$all.stats

test_downstroke <- rosnerTest(flights$downstroke_amplitude_m, k = 30)
test_downstroke$all.stats
```
```{r}

out_vres <- test_vres$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)


out_accx <- test_accx$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)


out_accy <- test_accy$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)


out_accz <- test_accz$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)


out_accres <- test_accres$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)


out_tr <- test_tr$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)


out_centri <- test_centri$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)

out_wingbeat <- test_wingbeat$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)

out_upstroke <- test_upstroke$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)

out_downstroke <- test_downstroke$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)


outliers_flights_speed <-rbind(out_vres, out_accx, out_accy, out_accz, out_accres) %>% distinct()

outliers_flights_tort <- rbind(out_tr, out_centri) %>% distinct()

outliers_flights_wing <- rbind(out_wingbeat, out_upstroke, out_downstroke) %>% distinct()

dim(outliers_flights_speed) # 14
dim(outliers_flights_tort) # 16
dim(outliers_flights_wing) # 43
```
```{r}
dim(flights) #937
#speed and acceleration variables

flights_speed  <- flights[-c(outliers_flights_speed$Obs.Num),]

# tortuosity variables 

flights_tort <- flights[-c(outliers_flights_tort$Obs.Num),]

# wingbeat and head displacement variables


flights_wing <- flights[-c(outliers_flights_wing$Obs.Num),]

dim(flights_speed) #937
dim(flights_tort) #923
dim(flights_wing) #894

```


# II. Model analysis 
## function detect_breakpoint()
```{r}

detect_breakpoint <- function(variable, df){
  
# Capture the symbol name passed (e.g., wing_loading)
  response <- deparse(substitute(variable))
  
  # Build a formula: response ~ Age*Sex + Brood
  form <- reformulate(termlabels = c("Age*Sex", "Brood"), response = response)
  
  # Fit linear model
  m_lin <- lm(form, data = df, na.action = na.omit)
  
  # Fit segmented model
  m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 28))  # initial guess at 36
  #summary(m_seg) # Age not significant. Larger brood, larger wing area
  plot(m_seg)  # quick diagnostic plot
  
  # Extract estimated breakpoint and slopes
  breakpoint <- m_seg$psi["psi1.Age", "Est."] #27.8
  breakpoint_se <- m_seg$psi["psi1.Age", "St.Err"]
  slope_early <- slope(m_seg)$Age[1, "Est."]
  slope_late  <- slope(m_seg)$Age[2, "Est."]
  model_summary <- summary(m_seg)
  
  # return both the model object and the numeric estimates
  out <- list(
    model_summary = model_summary,
    estimates = c(
      breakpoint = breakpoint,
      breakpoint_se = breakpoint_se,
      slope_early = slope_early, 
      slope_late = slope_late))
  
    return(out)
}

```

## function compare_models_aic()
```{r}
# Function to compare models using AIC, ΔAIC, and Akaike weights
compare_models_aic <- function(models) {
  # Validate input
  if (!is.list(models) || length(models) < 2) {
    stop("Please provide a list of at least two fitted model objects.")
  }
  
  # Extract model names
  model_names <- names(models)
  if (is.null(model_names) || any(model_names == "")) {
    model_names <- paste0("Model_", seq_along(models))
  }
  
  # Compute AIC for each model
  aic_values <- sapply(models, function(m) {
    tryCatch(AIC(m), error = function(e) NA)
  })
  
  # Check for NA values
  if (any(is.na(aic_values))) {
    stop("One or more models could not have their AIC computed.")
  }
  
  # Calculate ΔAIC
  delta_aic <- aic_values - min(aic_values)
  
  # Calculate Akaike weights
  rel_likelihood <- exp(-0.5 * delta_aic)
  akaike_weights <- rel_likelihood / sum(rel_likelihood)
  
  # Create results data frame
  results <- data.frame(
    Model = model_names,
    AIC = round(aic_values, 3),
    Delta_AIC = round(delta_aic, 3),
    Akaike_Weight = round(akaike_weights, 4),
    stringsAsFactors = FALSE
  )
  
  # Order by AIC
  results <- results[order(results$AIC), ]
  rownames(results) <- NULL
  
  return(results)
}
```

## function compare_models_pmin()
```{r}

compare_models_pmin <- function(variable, breakpoint, df){
  
  # 1) Hinge coding: 0 before breakpoint, positive after
  # using pmin() as hinge control with pmax() highly correlates with Age. 
  df$hinge <- pmin(df$Age - breakpoint, 0)

  # 2) Capture response name for formulas
  response <- deparse(substitute(variable))

  # Helpers
  form_lm   <- function(terms) reformulate(terms, response = response)
  # For lmer, we must build the full formula string because of random-effects syntax
  form_lmer <- function(rhs) as.formula(paste(response, "~", rhs))

  ## ----- Linear models (fixed effects) -----
  m1 <- lm(form_lm(c("Age", "Sex", "hinge", "Brood", "Bird")), data = df)
  m2 <- lm(form_lm(c("Age", "Sex", "hinge", "Bird")), data = df)
  m3 <- lm(form_lm(c("Age*Sex", "hinge", "Bird")), data = df)
  m4 <- lm(form_lm(c("Age*Sex", "hinge", "Brood", "Bird")), data = df)

  ## ----- Linear mixed models (random effects) -----
  # NOTE: If comparing mixed models by AIC across different fixed-effect structures,
  # consider REML = FALSE; uncomment the REML = FALSE versions if your
  # compare_models_aic relies on ML fits.
  m5  <- lme4::lmer(form_lmer("Age + Sex + Brood + hinge + (1|Bird)"), data = df)         # , REML = FALSE
  m6  <- lme4::lmer(form_lmer("Age + Sex + hinge + (1|Bird)"), data = df,)                  # , REML = FALSE
  m7  <- lme4::lmer(form_lmer("Age*Sex + Brood + hinge + (1|Bird)"), data = df)            # , REML = FALSE
  m8  <- lme4::lmer(form_lmer("Age*Sex + hinge + (1|Bird)"), data = df)                    # , REML = FALSE
  m9  <- lme4::lmer(form_lmer("Age*Sex + hinge + (1|Brood/Bird)"), data = df)              # , REML = FALSE
  m10 <- lme4::lmer(form_lmer("Age + Sex + hinge + (1|Brood/Bird)"), data = df)            # , REML = FALSE

  ## ----- Compare models -----
  all_models <- compare_models_aic(list(
    "m1-lm no interaction"                  = m1,
    "m2-lm no interaction, no brood"        = m2,
    "m3-lm interaction, no brood"           = m3,
    "m4-lm interaction"                     = m4,
    "m5-(1|Bird), no interaction"           = m5,
    "m6-(1|Bird), no interaction, no brood" = m6,
    "m7-(1|Bird), interaction"              = m7,
    "m8-(1|Bird), interaction, no brood"    = m8,
    "m9-(1|Brood/Bird), interaction"        = m9,
    "m10-(1|Brood/Bird), no interaction"    = m10
  ))

  just_lmer <- compare_models_aic(list(
    "m5-(1|Bird), no interaction"           = m5,
    "m6-(1|Bird), no interaction, no brood" = m6,
    "m7-(1|Bird), interaction"              = m7,
    "m8-(1|Bird), interaction, no brood"    = m8,
    "m9-(1|Brood/Bird), interaction"        = m9,
    "m10-(1|Brood/Bird), no interaction"    = m10
  ))
  
  bonus <- performance::compare_performance(m1, m2, m3, m4, m5, m6, m7, m8, m9, m10, rank = TRUE)
  
  # Return comparisons AND the fitted models
  fits <- list(m1=m1, m2=m2, m3=m3, m4=m4, m5=m5, m6=m6, m7=m7, m8=m8, m9=m9, m10=m10)


  return(list(all_models = all_models, just_lmer = just_lmer, bonus = bonus, fits))
}



```
## function compare_models_pmax()
```{r}

compare_models_pmax <- function(variable, breakpoint, df){
  
  # 1) Hinge coding: 0 before breakpoint, positive after
  # using pmin() as hinge control with pmax() highly correlates with Age. 
  df$hinge <- pmax(df$Age - breakpoint, 0)

  # 2) Capture response name for formulas
  response <- deparse(substitute(variable))

  # Helpers
  form_lm   <- function(terms) reformulate(terms, response = response)
  # For lmer, we must build the full formula string because of random-effects syntax
  form_lmer <- function(rhs) as.formula(paste(response, "~", rhs))

  ## ----- Linear models (fixed effects) -----
  m1 <- lm(form_lm(c("Age", "Sex", "hinge", "Brood", "Bird")), data = df)
  m2 <- lm(form_lm(c("Age", "Sex", "hinge", "Bird")), data = df)
  m3 <- lm(form_lm(c("Age*Sex", "hinge", "Bird")), data = df)
  m4 <- lm(form_lm(c("Age*Sex", "hinge", "Brood", "Bird")), data = df)

  ## ----- Linear mixed models (random effects) -----
  # NOTE: If comparing mixed models by AIC across different fixed-effect structures,
  # consider REML = FALSE; uncomment the REML = FALSE versions if your
  # compare_models_aic relies on ML fits.
  m5  <- lme4::lmer(form_lmer("Age + Sex + Brood + hinge + (1|Bird)"), data = df)         # , REML = FALSE
  m6  <- lme4::lmer(form_lmer("Age + Sex + hinge + (1|Bird)"), data = df,)                  # , REML = FALSE
  m7  <- lme4::lmer(form_lmer("Age*Sex + Brood + hinge + (1|Bird)"), data = df)            # , REML = FALSE
  m8  <- lme4::lmer(form_lmer("Age*Sex + hinge + (1|Bird)"), data = df)                    # , REML = FALSE
  m9  <- lme4::lmer(form_lmer("Age*Sex + hinge + (1|Brood/Bird)"), data = df)              # , REML = FALSE
  m10 <- lme4::lmer(form_lmer("Age + Sex + hinge + (1|Brood/Bird)"), data = df)            # , REML = FALSE

  ## ----- Compare models -----
  all_models <- compare_models_aic(list(
    "m1-lm no interaction"                  = m1,
    "m2-lm no interaction, no brood"        = m2,
    "m3-lm interaction, no brood"           = m3,
    "m4-lm interaction"                     = m4,
    "m5-(1|Bird), no interaction"           = m5,
    "m6-(1|Bird), no interaction, no brood" = m6,
    "m7-(1|Bird), interaction"              = m7,
    "m8-(1|Bird), interaction, no brood"    = m8,
    "m9-(1|Brood/Bird), interaction"        = m9,
    "m10-(1|Brood/Bird), no interaction"    = m10
  ))

  just_lmer <- compare_models_aic(list(
    "m5-(1|Bird), no interaction"           = m5,
    "m6-(1|Bird), no interaction, no brood" = m6,
    "m7-(1|Bird), interaction"              = m7,
    "m8-(1|Bird), interaction, no brood"    = m8,
    "m9-(1|Brood/Bird), interaction"        = m9,
    "m10-(1|Brood/Bird), no interaction"    = m10
  ))
  
  bonus <- performance::compare_performance(m1, m2, m3, m4, m5, m6, m7, m8, m9, m10, rank = TRUE)
  
  # Return comparisons AND the fitted models
  fits <- list(m1=m1, m2=m2, m3=m3, m4=m4, m5=m5, m6=m6, m7=m7, m8=m8, m9=m9, m10=m10)


  return(list(all_models = all_models, just_lmer = just_lmer, bonus = bonus, fits))
}




```
## function check_assumptions()
```{r}

check_assumptions <- function(model){
  
  # check collinearity
  check_collinearity(model)
  
  # Computes the default residuals (raw conditional)
  raw_cond_model <- redres::compute_redres(model)
  
  hist(raw_cond_model, main = "Normality of residuals") #normality in histogram
  p_hetero <- plot_redres(model) #heteroskedasticity
  p_resqq <- plot_resqq(model) #residual QQ
  p_ranef <- plot_ranef(model) #random effects QQ / density 
  
  print(p_hetero)
  print(p_resqq)
  print(p_ranef)
  
  invisible(NULL)
}
```

## function plot_final
```{r}

plot_final <- function(breakpoint, breakpoint_se,
                           model,            # lmer model
                           df,               # data used for model
                           base_plot,        # ggplot object
                           include_brood = TRUE,
                           brood_level = "Box5",
                           sex = "M"
                           ) {
  df <- as.data.frame(df)
  # 1) Hinge coding
  df$hinge <- pmin(df$Age - breakpoint, 0)

  # 2) Age sequence
  age_seq <- seq(min(df$Age, na.rm = TRUE), max(df$Age, na.rm = TRUE), by = 1)

  # 3) Ensure factors
  if ("Sex" %in% names(df) && !is.factor(df$Sex)) df$Sex <- factor(df$Sex)
  if ("Brood" %in% names(df) && !is.factor(df$Brood)) df$Brood <- factor(df$Brood)

  # 4) Base newdat
  newdat_out <- data.frame(
    Age = age_seq,
    Sex = factor(sex, levels = levels(df$Sex)),
    hinge = pmin(age_seq - breakpoint, 0)
  )

  # 5) Add Brood if requested
  if (include_brood) {
    if (!("Brood" %in% names(df))) {
      stop("'Brood' column not found in df.")
    }
    if (!(brood_level %in% levels(df$Brood))) {
      warning(sprintf("brood_level='%s' not found; using reference '%s'.",
                      brood_level, levels(df$Brood)[1]))
      brood_level <- levels(df$Brood)[1]
    }
    newdat_out$Brood <- factor(brood_level, levels = levels(df$Brood))
  }

  # 6) Predict (population-level)
  newdat_out$pred <- predict(model, newdat_out, re.form = NA)

  # 7) Compose plot
  yr <- ggplot_build(base_plot)$layout$panel_params[[1]]$y.range
  y_top <- max(yr)

  p <- base_plot +
    
    ggplot2::geom_line(
      data = newdat_out,
      ggplot2::aes(x = Age, y = pred),
      color = "#1f77b4", size = 1.5, inherit.aes = FALSE
    ) +
    ggplot2::geom_vline(xintercept = breakpoint, linetype = "dashed", color = "red") +
    ggplot2::annotate(
        "segment",
        x = breakpoint - breakpoint_se,
        xend = breakpoint + breakpoint_se,
        y    = y_top,
        yend = y_top,
        color = "#E2AFDE",
        linewidth = 3,
        alpha = 0.7) +
    ggplot2::labs(subtitle = sprintf("breakpoint = %.1f (± %.1f SE)", breakpoint, breakpoint_se))
  

  return(list(plot = p, newdat = newdat_out))
}


```
## A. Significant
### MORPHOLOGY
#### bodyweight - 28.1
##### breakpoint detection
```{r}
out <- detect_breakpoint(scale_weight, wing_data_final)
print(out)
breakpoint_sw <- round(out$estimates["breakpoint"], 1)
breakpoint_se_sw <- round(out$estimates["breakpoint_se"], 1)

```
##### aic comparison
```{r}
wing_data_final$scale_g <- wing_data_final$scale_weight*101.97

model_results <- compare_models_pmin(scale_g, breakpoint_sw, wing_data_final)

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_sw <- model_results[[4]]$m5
m_sw
```

##### model assumption checks
```{r}
check_collinearity(m_sw) #checking collinearity for the best linear model

# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(m_sw)
  
hist(raw_cond_model, main = "Normality of residuals") #normality in histogram
plot_redres(m_sw) #heteroskedasticity
plot_resqq(m_sw) #residual QQ
plot_ranef(m_sw)
```

##### test
```{r}
summary(m_sw)
tab_model(m_sw)
anova_sw <- Anova(m_sw, type =3, test.statistic = "F") %>% as.data.frame() %>%
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

#before slope = 0?
bs_sw <- linearHypothesis(m_sw, "Age+hinge = 0", test = "F") %>% as.data.frame()# Is the slope before the breakpoint flat? -> no

#after slope = 0?
as_sw <- linearHypothesis(m_sw, "Age = 0", test = "F") %>% as.data.frame() # Is the slope after the breakpoint flat? -> no

#same slope?
ss_sw <- linearHypothesis(m_sw, "Age = Age+hinge", test = "F") %>% as.data.frame()#Is the slope before and after the same? -> n

#before breakpoint slope estimate
est_bs_sw <- deltaMethod(m_sw, "Age + hinge") %>% as.data.frame()

#after breakpoint slope estimate
est_as_sw <- deltaMethod(m_sw, "Age") %>% as.data.frame()

anova_sw
bs_sw
as_sw
ss_sw
est_bs_sw
est_as_sw

```


```{r}

ageF <- anova_sw["Age",]$F
  ageDf <- anova_sw["Age",]$Df
  ageDf.res <- anova_sw["Age",]$Df.res
  ageP <- anova_sw["Age",]$p_FDR
  
  SexF <- anova_sw["Sex",]$F
  SexDf <- anova_sw["Sex",]$Df
  SexDf.res <- anova_sw["Sex",]$Df.res
  SexP <- anova_sw["Sex",]$p_FDR
  
  BroodF <- anova_sw["Brood",]$F
  BroodDf <- anova_sw["Brood",]$Df
  BroodDf.res <- anova_sw["Brood",]$Df.res
  BroodP <- anova_sw["Brood",]$p_FDR
  
  hingeF <- anova_sw["hinge",]$F
  hingeDf <- anova_sw["hinge",]$Df
  hingeDf.res <- anova_sw["hinge",]$Df.res
  hingeP <- anova_sw["hinge",]$p_FDR
  



```


```{r}
# Before slope
print(deltaMethod(model_sw, "Age + hinge"))

# After slope
print(deltaMethod(model_sw, "Age"))


```
##### plot
```{r}

output <- plot_final(breakpoint_sw, breakpoint_se_sw, 
                     model_sw, 
                     wing_data_final,
                     plot_scale_g,
                     include_brood = TRUE,
                     brood_level = "Box5",
                     sex = "M"
                     )
plot_scale_g_final <- output[1]
plot_scale_g_final

```


#### wingspan - 28
##### breakpoint detection
```{r}
out <- detect_breakpoint(wingspan, wing_data_final)
print(out)
breakpoint_ws <- round(out$estimates["breakpoint"], 1)
breakpoint_se_ws <- round(out$estimates["breakpoint_se"], 1)
```
##### aic comparison
```{r}

model_results <- compare_models_pmin(wingspan, breakpoint_ws, wing_data_final)

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_wingspan <- model_results[[4]]$m5
m_wingspan


```

##### model assumption check
```{r}
check_collinearity(m_wingspan)
  
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(m_wingspan)
  
hist(raw_cond_model, main = "Normality of residuals") #normality in histogram
plot_redres(m_wingspan) #heteroskedasticity
plot_resqq(m_wingspan) #residual QQ
plot_ranef(m_wingspan)
```
##### test
```{r}
# for residual variance / R^2
tab_model(m_wingspan)


# Wald F test with KR approximation & FDR
anova_wingspan <- Anova(m_wingspan, type =3, test.statistic = "F") %>% as.data.frame() %>%
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_wingspan

# Hypothesis testing
linearHypothesis(m_wingspan, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
linearHypothesis(m_wingspan, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
linearHypothesis(m_wingspan, "Age = Age+hinge", test = "F") #Is the slope before and after the same? -> n

# slope estimates, SE, CI 

#before slope
print(deltaMethod(m_wingspan, "Age + hinge"))
#after slope
print(deltaMethod(m_wingspan, "Age"))


```
##### plot
```{r}

output <- plot_final(breakpoint_ws, breakpoint_se_ws, 
                     m_wingspan, 
                     wing_data_final,
                     plot_wingspan,
                     include_brood = TRUE,
                     brood_level = "Box5",
                     sex = "M"
                     )
plot_wingspan_final <- output[1]
plot_wingspan_final

```

#### wing area - 28.5
##### breakpoint detection
```{r}
out <- detect_breakpoint(wingarea, wing_data_final)
print(out)
breakpoint_wa <- round(out$estimates["breakpoint"], 1)
breakpoint_se_wa <- round(out$estimates["breakpoint_se"], 1)
```
##### aic comparison
```{r}

model_results <- compare_models_pmin(wingarea, breakpoint_wa, wing_data_final)

# shows the results of model comparison
model_results[1:3]
```
```{r}
# saves the model to test 
m_wingarea <- model_results[[4]]$m6
m_wingarea


```

##### model assumption check
```{r}

check_collinearity(m_wingarea)
  
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(m_wingarea)
  
hist(raw_cond_model, main = "Normality of residuals") #normality in histogram
plot_redres(m_wingarea) #heteroskedasticity
plot_resqq(m_wingarea) #residual QQ
plot_ranef(m_wingarea)
```
##### test
```{r}


# for residual variance / R^2
tab_model(m_wingarea)

# Wald F test with KR approximation & FDR
anova_wingarea <- Anova(m_wingarea, type =3, test.statistic = "F") %>% as.data.frame() %>%
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_wingarea

# Hypothesis testing
linearHypothesis(m_wingarea, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
linearHypothesis(m_wingarea, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
linearHypothesis(m_wingspan, "Age = Age+hinge", test = "F")

# slope estimates, SE, CI 

#before slope
print(deltaMethod(m_wingspan, "Age + hinge"))
#after slope
print(deltaMethod(m_wingspan, "Age"))

```
##### plot
```{r}

output <- plot_final(breakpoint_wa, breakpoint_se_wa, 
                     m_wingarea, 
                     wing_data_final,
                     plot_wingarea,
                     include_brood = FALSE,
                     brood_level = "Box5",
                     sex = "M"
                     )
plot_wingarea_final <- output[1]
plot_wingarea_final

```
#### wing loading
##### breakpoint duration
```{r}

m_lin <- lm(wing_loading ~ Age*Sex + Brood, data = wing_data_final) #CHANGE HERE
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_wl <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
breakpoint_se_wl <- m_seg$psi["psi1.Age", "St.Err"] # breakpoint se.


breakpoint_wl #32.9
breakpoint_se_wl


```
##### aic comparison
```{r}

model_results <- compare_models_pmax(wing_loading, breakpoint_wl, wing_data_final) #CHANGE

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_wl <- model_results[[4]]$m10 # in theory could
m_wl

```
##### model assumptions
```{r}

model <- m_wl
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)



```
##### test
```{r}

model <- m_wl

tab_model(model)
anova_wl <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_wl # CHANGE THIS!!

linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
linearHypothesis(model, "Age = Age+hinge", test = "F")

#before slope
print(deltaMethod(model, "Age + hinge"))
#after slope
print(deltaMethod(model, "Age"))

```
##### plot
```{r}

output <- plot_final(breakpoint_wl, breakpoint_se_wl, #CHANGE HERE
                     m_wl, # CHANGE HERE
                     wing_data_final, # CHANGE HERE
                     plot_wingloading, # CHANGE HERE
                     include_brood = FALSE,
                     brood_level = "Box3",
                     sex = "M"
                     )
plot_wl_final <- output[1] #CHANGE HERE
plot_wl_final # CHANGE HERE


```

### MODEL OUTPUT
#### function fmt_term: F = xx.xx (Df, Df.res) (p-display)
```{r}
# Helper to format one term: F = xx.xx (Df, Df.res) (p-display)
fmt_term <- function(F, Df, Df_res, p, digits_F = 1, digits_Df = 1, p_thresh = 0.001) {
  if (anyNA(c(F, Df, Df_res, p))) return(NA_character_)
  F_str   <- format(round(F, digits_F), nsmall = digits_F, trim = TRUE)
  Df_str  <- format(Df, trim = TRUE)
  Res_str <- format(round(Df_res, digits_Df), nsmall = digits_Df, trim = TRUE)
  
  p_str <- if (is.na(p)) {
      "p = NA"
    } else if (p < p_thresh) {
      "p < 0.001"
    } else {
      paste0("p = ", format(round(p, 2), nsmall = 2))
    }

  sprintf("F<sub>%s, %s</sub> = %s, %s",  Df_str, Res_str, F_str, p_str)

  }
```

```{r}
library(knitr)

labels <- c("scaled_weight", "wingspan", "wingarea")
anova_tabs <- list(anova_sw, anova_wingspan, anova_wingarea) #anova dataframe
lmer_mods <- list(m_sw, m_wingspan, m_wingarea)             # fitted lmer models

#stop if the number of anova tabs != number of models. It should
stopifnot(length(labels) == length(anova_tabs), length(anova_tabs) == length(lmer_mods))

#storage for all rows
out_all <- list()

#hypothesis test output
hypothesis_tests <- list()


for (i in seq_along(models)){
  
  tab <- anova_tabs[[i]] # the ANOVA data.frame for term extraction
  mod <- lmer_mods[[i]] # the fitted model for summary metrics

  
# --- helper to extract a term safely from the current model ---
  get_term <- function(term, what) {
    # if the term is not present, return NA
    if (!term %in% rownames(tab)) return(NA_real_)
    val <- tab[term, ][[what]]
    if (is.null(val)) return(NA_real_)
    val
  }

  # pull values for each term from *this* model
  ageF      <- get_term("Age",   "F")
  ageDf     <- get_term("Age",   "Df")
  ageDf.res <- get_term("Age",   "Df.res")
  ageP      <- get_term("Age",   "p_FDR")

  SexF      <- get_term("Sex",   "F")
  SexDf     <- get_term("Sex",   "Df")
  SexDf.res <- get_term("Sex",   "Df.res")
  SexP      <- get_term("Sex",   "p_FDR")

  hingeF      <- get_term("hinge", "F")
  hingeDf     <- get_term("hinge", "Df")
  hingeDf.res <- get_term("hinge", "Df.res")
  hingeP      <- get_term("hinge", "p_FDR")
  
  # Brood is optional — if missing, all will be NA
  BroodF      <- get_term("Brood",   "F")
  BroodDf     <- get_term("Brood",   "Df")
  BroodDf.res <- get_term("Brood",   "Df.res")
  BroodP      <- get_term("Brood",   "p_FDR")
  
  #Optional interaction
  AgeSexF      <- get_term("Age:Sex",   "F")
  AgeSexDf     <- get_term("Age:Sex",   "Df")
  AgeSexDf.res <- get_term("Age:Sex",   "Df.res")
  AgeSexP      <- get_term("Age:Sex",   "p_FDR")
  
  # Build the single-row data frame in your requested format ---
  out_row <- data.frame(
    Age   = fmt_term(ageF,   ageDf,   ageDf.res,   ageP),
    Sex   = fmt_term(SexF,   SexDf,   SexDf.res,   SexP),
    Hinge = fmt_term(hingeF, hingeDf, hingeDf.res,  hingeP),# uses the parenthesis style
    Brood = fmt_term(BroodF, BroodDf, BroodDf.res, BroodP),
    AgeSex = fmt_term(AgeSexF, AgeSexDf, AgeSexDf.res, AgeSexP) # uses "p = xx" style
    
    )
  
  # model summary variables
  
  #vc <- VarCorr(model)
  sigma_fixed <- sigma(mod)
  sigma_Bird <- as.data.frame(VarCorr(mod))$sdcor[1]
  ICC <- icc(mod)$ICC_adjusted
  N_Bird <- ngrps(mod)
  Observations <- stats::nobs(mod)
  R2_marginal <- performance::r2(mod)$R2_marginal
  R2_conditional <- performance::r2(mod)$R2_conditional

  summary_cols <- data.frame(
    sigma_fixed   = round(sigma_fixed, 2),
    sigma_Bird    = round(sigma_Bird, 2),
    ICC           = round(ICC, 2),
    N_Bird        = N_Bird,
    Observations  = Observations,
    `Marginal R2 / Conditional R2` = paste0(round(R2_marginal,2), " / ", round(R2_conditional,2)),
    check.names = FALSE, 
    row.names = NULL
  )

  
  # Add the Variables column as the first column
  out_row <- cbind(Variables = labels[i], out_row, summary_cols)
  
  out_all[[i]] <- out_row
}

# Combine rows from all models into one data frame
out_all <- do.call(rbind, out_all)

kable(out_all, escape = FALSE)

```
### TEST OUTPUT
```{r}

library(knitr)

labels <- c("scaled_weight", "wingspan", "wingarea")

lmer_mods <- list(m_sw, m_wingspan, m_wingarea)             # fitted lmer models

out_all <- vector("list", length(lmer_mods))


for (i in seq_along(lmer_mods)){
  
  #tab <- anova_tabs[[i]] # the ANOVA data.frame for term extraction
  mod <- lmer_mods[[i]] # the fitted model for summary metrics

  #before slope = 0?
  before_slope <- linearHypothesis(mod, "Age+hinge = 0", test = "F") %>% as.data.frame()# Is the slope before the breakpoint flat? -> no
  
  #after slope = 0?
  after_slope <- linearHypothesis(mod, "Age = 0", test = "F") %>% as.data.frame() # Is the slope after the breakpoint flat? -> no
  
  #same slope?
  same_slope <- linearHypothesis(mod, "Age = Age+hinge", test = "F") %>% as.data.frame()#Is the slope before and after the same? -> n
  
  #before breakpoint slope estimate
  before_slope_estimate <- deltaMethod(mod, "Age + hinge") %>% as.data.frame()
  
  #after breakpoint slope estimate
  after_slope_estimate <- deltaMethod(mod, "Age") %>% as.data.frame()
  
  before_slope_mean <-  before_slope_estimate$Estimate 
  before_slope_se <-  before_slope_estimate$SE
  before_slope_lower <- before_slope_estimate$`2.5 %`
  before_slope_upper <- before_slope_estimate$`97.5 %`
  
  before_slope_F <- before_slope$F[2]
  before_slope_Df <- before_slope$Df[2]
  before_slope_ResDf <- before_slope$Res.Df[2]
  before_slope_P <- before_slope$`Pr(>F)`[2]
  
  after_slope_mean <-  after_slope_estimate$Estimate 
  after_slope_se <-  after_slope_estimate$SE
  after_slope_lower <- after_slope_estimate$`2.5 %`
  after_slope_upper <- after_slope_estimate$`97.5 %`
  
  after_slope_F <- after_slope$F[2]
  after_slope_Df <- after_slope$Df[2]
  after_slope_ResDf <- after_slope$Res.Df[2]
  after_slope_P <- after_slope$`Pr(>F)`[2]
  
  same_slope_F <- same_slope$F[2]
  same_slope_Df <- same_slope$Df[2]
  same_slope_ResDf <- same_slope$Res.Df[2]
  same_slope_P <- same_slope$`Pr(>F)`[2]
  
  
  before_string <- sprintf(
    "%.2f ± %.2f [%.2f, %.2f]",
    before_slope_mean,
    before_slope_se,
    before_slope_lower,
    before_slope_upper)
  
  after_string <- sprintf(
    "%.2f ± %.2f [%.2f, %.2f]",
    after_slope_mean,
    after_slope_se,
    after_slope_lower,
    after_slope_upper)
  
  
  # test summary variables

  test_cols <- data.frame(
    "slope before Mean ± SE [CI]" = before_string,
    "slope before = 0?" = fmt_term(before_slope_F, before_slope_Df, before_slope_ResDf, before_slope_P),
    "slope after Mean ± SE [CI]" = after_string,
    "slope after = 0?" = fmt_term(after_slope_F, after_slope_Df, after_slope_ResDf, after_slope_P),
    "slope before = after?" = fmt_term(same_slope_F, same_slope_Df, same_slope_ResDf, same_slope_P),
    check.names = FALSE, 
    row.names = NULL
  )

  
  # Add the Variables column as the first column
  out_row <- cbind(Variables = labels[i], test_cols)
  
  out_all[[i]] <- out_row
}

# Combine rows from all models into one data frame
out_all_test <- do.call(rbind, out_all)

kable(out_all_test, escape = FALSE)



```

### LEG


#### max Fz-> x(paper)/bw -40
##### breakpoint detection
```{r}
out <- detect_breakpoint(Fz_maxFtot_bw, force_clean)
print(out)
breakpoint_Fx <- round(out$estimates["breakpoint"], 1)
breakpoint_se_Fx <- round(out$estimates["breakpoint_se"], 1)
```
##### aic comparison
```{r}

model_results <- compare_models_pmin(Fz_maxFtot_bw, breakpoint_Fx, force_clean)

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
lm_Fx <- model_results[[4]]$m1
m_Fx <- model_results[[4]]$m6
#m_Fx_brood <- model_results[[4]]$m10


```

##### model assumption check
```{r}

check_collinearity(lm_Fx)
check_collinearity(m_Fx)



  
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(m_Fx)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(m_Fx) #heteroskedasticity
plot_resqq(m_Fx) #residual QQ
plot_ranef(m_Fx)

```
##### test
```{r}

#before slope
print(deltaMethod(m_Fx, "Age + hinge"))
#after slope
print(deltaMethod(m_Fx, "Age"))

#hypothesis test
linearHypothesis(m_Fx, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
linearHypothesis(m_Fx, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
linearHypothesis(m_Fx, "Age = Age+hinge", test = "F")

#term test
anova_Fx <- Anova(m_Fx, type =3, test.statistic = "F") %>% as.data.frame() %>%
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_Fx

# Random effects 
tab_model(m_Fx)



# slope estimates, SE, CI 



```
##### plot
```{r}

output <- plot_final(breakpoint_Fx, breakpoint_se_Fx, 
                     m_Fx, 
                     force_clean,
                     plot_age_Fx_bw,
                     include_brood = FALSE,
                     brood_level = "Box5",
                     sex = "M"
                     )
plot_Fx_final <- output[1]
plot_Fx_final

```
#### max Fy / bw - 33.8
##### breakpoint detection
```{r}
out <- detect_breakpoint(Fy_maxFtot_bw, force_clean)
print(out)
breakpoint_Fy <- round(out$estimates["breakpoint"], 1)
breakpoint_se_Fy <- round(out$estimates["breakpoint_se"], 1)
```
##### aic comparison
```{r}

model_results <- compare_models_pmin(Fy_maxFtot_bw, breakpoint_Fy, force_clean)

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_Fy <- model_results[[4]]$m6
m_Fy


```

##### model assumption check
```{r}
check_collinearity(m_Fy)
  
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(m_Fy)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(m_Fy) #heteroskedasticity
plot_resqq(m_Fy) #residual QQ
plot_ranef(m_Fy)
```
##### test
```{r}

tab_model(m_Fy)
anova_Fy <- Anova(m_Fy, type =3, test.statistic = "F") %>% as.data.frame() %>%
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_Fy

linearHypothesis(m_Fy, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
linearHypothesis(m_Fy, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
linearHypothesis(m_Fy, "Age = Age+hinge", test = "F")

#before slope
print(deltaMethod(m_Fy, "Age + hinge"))
#after slope
print(deltaMethod(m_Fy, "Age"))

```
##### plot
```{r}

output <- plot_final(breakpoint_Fy, breakpoint_se_Fy, 
                     m_Fy, 
                     force_clean,
                     plot_age_Fy_bw,
                     include_brood = FALSE,
                     brood_level = "Box5",
                     sex = "M"
                     )
plot_Fy_final <- output[1]
plot_Fy_final

```
#### max Ftot /bw - 37.1
##### breakpoint detection
```{r}
out <- detect_breakpoint(max_Ftot_bw, force_clean)
print(out)
breakpoint_Ftot <- round(out$estimates["breakpoint"], 1)
breakpoint_se_Ftot <- round(out$estimates["breakpoint_se"], 1)
```
##### aic comparison
```{r}

model_results <- compare_models_pmin(max_Ftot_bw, breakpoint_Ftot, force_clean)

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_Ftot <- model_results[[4]]$m6
m_Ftot


```

##### model assumption check
```{r}
check_collinearity(m_Ftot)
  
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(m_Ftot)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(m_Ftot) #heteroskedasticity
plot_resqq(m_Ftot) #residual QQ
plot_ranef(m_Ftot)
```
##### test
```{r}
model <- m_Ftot

#before slope
print(deltaMethod(model, "Age + hinge"))
#after slope
print(deltaMethod(model, "Age"))


linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
linearHypothesis(model, "Age = Age+hinge", test = "F")

anova_Ftot <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>%
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_Ftot

tab_model(model)


```
##### plot
```{r}

output <- plot_final(breakpoint_Ftot, breakpoint_se_Ftot, 
                     m_Ftot, 
                     force_data,
                     plot_raw_Ftot,
                     include_brood = FALSE,
                     brood_level = "Box5",
                     sex = "M"
                     )
plot_Ftot_final <- output[1]
plot_Ftot_final

```

#### take-off angle - 34.01
##### breakpoint detection
```{r}
out <- detect_breakpoint(angle, force_clean)
print(out)
breakpoint_angle <- round(out$estimates["breakpoint"], 1)
breakpoint_se_angle <- round(out$estimates["breakpoint_se"], 1)
```
##### aic comparison
```{r}

model_results <- compare_models_pmin(angle, breakpoint_angle, force_clean)

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_angle <- model_results[[4]]$m7
m_angle


```

##### model assumption check
```{r}

check_collinearity(m_angle)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(m_angle)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(m_angle) #heteroskedasticity
plot_resqq(m_angle) #residual QQ
plot_ranef(m_angle)
```
##### test
```{r}
model <- m_angle

tab_model(model)
anova_angle <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>%
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_angle

linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
linearHypothesis(model, "Age = Age+hinge", test = "F")

#before slope
print(deltaMethod(model, "Age + hinge"))
#after slope
print(deltaMethod(model, "Age"))
```
##### plot
```{r}

output_M <- plot_final(breakpoint_angle, breakpoint_se_angle, 
                     m_angle, 
                     force_data,
                     plot_age_angle,
                     include_brood = TRUE,
                     brood_level = "Box5",
                     sex = "M"
                     )
output_F <- plot_final(breakpoint_angle, breakpoint_se_angle, 
                     m_angle, 
                     force_data,
                     plot_age_angle,
                     include_brood = TRUE,
                     brood_level = "Box5",
                     sex = "F"
                     )


plot_angle_M <- output_M[1]
plot_angle_F <- output_F[1]
plot_angle_M
plot_angle_F
```
##### plot with brood & Sex
- salmon (F) and blue (M) are in the same brood and green (M) in a different brood. 
```{r}

# 1) Hinge coding
force_data$hinge <- pmin(force_data$Age - breakpoint_angle, 0)

# 2) Age sequence
age_seq <- seq(min(force_data$Age, na.rm = TRUE), max(force_data$Age, na.rm = TRUE), by = 1)

# 4) Base newdat
newdat_F_wpbr <- data.frame(
    Age = age_seq,
    Sex = "F",
    Brood = "WhitePinkR-BlueR",
    hinge = pmin(age_seq - breakpoint_angle, 0)
  )

newdat_M_wpbr <- data.frame(
    Age = age_seq,
    Sex = "M",
    Brood = "WhitePinkR-BlueR",
    hinge = pmin(age_seq - breakpoint_angle, 0)
  )

newdat_M_Box11 <- data.frame(
  Age = age_seq, 
  Sex = "M",
  Brood = "Box11",
  hinge = pmin(age_seq - breakpoint_angle, 0)
)

# 6) Predict (population-level)
newdat_F_wpbr$pred <- predict(m_angle, newdat_F_wpbr, re.form = NA)
newdat_M_wpbr$pred <- predict(m_angle, newdat_M_wpbr, re.form = NA)
newdat_M_Box11$pred <- predict(m_angle, newdat_M_Box11, re.form = NA)


# 7) Compose plot
yr <- ggplot_build(plot_age_angle)$layout$panel_params[[1]]$y.range
y_top <- max(yr)

p <- plot_age_angle +
    # prediction line no.1 
    ggplot2::geom_line(
      data = newdat_F_wpbr,
      ggplot2::aes(x = Age, y = pred),
      color = "#E08D79", size = 1.5, inherit.aes = FALSE
    ) +
  
   # prediction line no.2 
    ggplot2::geom_line(
      data = newdat_M_wpbr,
      ggplot2::aes(x = Age, y = pred),
      color = "#1f77b4", size = 1.5, inherit.aes = FALSE
    ) +
   # prediction line no. 3
    ggplot2::geom_line(
      data = newdat_M_Box11,
      ggplot2::aes(x = Age, y = pred),
      color = "#4D9078", size = 1.5, inherit.aes = FALSE
    ) +
  
    ggplot2::geom_vline(xintercept = breakpoint_angle, linetype = "dashed", color = "red") +
    ggplot2::annotate(
        "segment",
        x = breakpoint_angle - breakpoint_se_angle,
        xend = breakpoint_angle + breakpoint_se_angle,
        y    = y_top,
        yend = y_top,
        color = "#E2AFDE",
        linewidth = 3,
        alpha = 0.7) +
    ggplot2::labs(subtitle = sprintf("breakpoint = %.1f (± %.1f SE)", breakpoint_angle, breakpoint_se_angle))
  
p

```

#### impulse -30.2
##### breakpoint detection
```{r}
#getting rid of 3702006 because there is no connected flight data and seems like there were three push attempts 

out <- detect_breakpoint(impulse*1000, force_clean)
print(out)
breakpoint_impulse <- round(out$estimates["breakpoint"], 1)
breakpoint_se_impulse <- round(out$estimates["breakpoint_se"], 1)
```
##### aic comparison
```{r}

model_results <- compare_models_pmin(impulse*1000, breakpoint_impulse, force_clean)

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_impulse <- model_results[[4]]$m5
m_impulse


```
##### model assumption check
```{r}

check_collinearity(m_impulse)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(m_impulse)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(m_impulse) #heteroskedasticity
plot_resqq(m_impulse) #residual QQ
plot_ranef(m_impulse)
```
##### test
```{r}
model <- m_impulse

tab_model(model)
anova_impulse <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>%
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_impulse

linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
linearHypothesis(model, "Age = Age+hinge", test = "F")

#before slope
print(deltaMethod(model, "Age + hinge"))
#after slope
print(deltaMethod(model, "Age"))



```

##### plot
```{r}

output <- plot_final(breakpoint_impulse, breakpoint_se_impulse, 
                     m_impulse, 
                     force_clean,
                     plot_age_impulse,
                     include_brood = TRUE,
                     brood_level = "Box3",
                     sex = "M"
                     )
plot_impulse_final <- output[1]
plot_impulse_final


```

#### left impulse -32
##### breakpoint detection
```{r}
m_lin <- lm(left_impulse*1000 ~ Age*Sex + Brood, data = force_clean)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_left_impulse <- m_seg$psi["psi1.Age", "Est."]
breakpoint_se_left_impulse <- m_seg$psi["psi1.Age", "St.Err"]
slope_early_left_impulse <- slope(m_seg)$Age[1, "Est."]
slope_late_left_impulse  <- slope(m_seg)$Age[2, "Est."]

breakpoint_left_impulse
slope_early_left_impulse
slope_late_left_impulse
```
##### aic comparison
```{r}

model_results <- compare_models_pmin(left_impulse*1000, breakpoint_left_impulse, force_clean)

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_left_impulse <- model_results[[4]]$m5
m_left_impulse


```
##### model assumptions
```{r}
check_collinearity(m_left_impulse)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(m_left_impulse)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(m_left_impulse) #heteroskedasticity
plot_resqq(m_left_impulse) #residual QQ
plot_ranef(m_left_impulse)


```
##### test
```{r}

model <- m_left_impulse

tab_model(model)
anova_left_impulse <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>%
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_left_impulse

linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
linearHypothesis(model, "Age = Age+hinge", test = "F")

#before slope
print(deltaMethod(model, "Age + hinge"))
#after slope
print(deltaMethod(model, "Age"))


```
##### plot
```{r}

output <- plot_final(breakpoint_left_impulse, breakpoint_se_left_impulse, 
                     m_left_impulse, 
                     force_clean,
                     plot_left_impulse,
                     include_brood = TRUE,
                     brood_level = "Box10",
                     sex = "M"
                     )
plot_left_impulse_final <- output[1]
plot_left_impulse_final


```
#### right impulse - 24.9
##### breakpoint detection
```{r}
m_lin <- lm(right_impulse*1000 ~ Age*Sex + Brood, data = force_clean)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_right_impulse <- m_seg$psi["psi1.Age", "Est."] #36.5. # male max resultant force generally stronger than females.
breakpoint_se_right_impulse <- m_seg$psi["psi1.Age", "St.Err"]
slope_early <- slope(m_seg)$Age[1, "Est."]
slope_late  <- slope(m_seg)$Age[2, "Est."]

breakpoint_right_impulse
breakpoint_se_right_impulse
```
##### aic comparison
```{r}

model_results <- compare_models_pmin(right_impulse*1000, breakpoint_right_impulse, force_clean)

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_right_impulse <- model_results[[4]]$m5
m_right_impulse
```
##### model assumptions
```{r}
model <- m_right_impulse
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)
```
##### test
```{r}

model <- m_right_impulse

tab_model(model)
anova_right_impulse <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>%
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_right_impulse

linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
linearHypothesis(model, "Age = Age+hinge", test = "F")

#before slope
print(deltaMethod(model, "Age + hinge"))
#after slope
print(deltaMethod(model, "Age"))
```
##### plot
```{r}

output <- plot_final(breakpoint_right_impulse, breakpoint_se_right_impulse, 
                     m_right_impulse, 
                     force_clean,
                     plot_right_impulse,
                     include_brood = TRUE,
                     brood_level = "Box11",
                     sex = "M"
                     )
plot_right_impulse_final <- output[1]
plot_right_impulse_final


```
#### duration - 39.8
##### breakpoint detection
```{r}
m_lin <- lm(100*duration ~ Age*Sex + Brood, data = force_clean)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_duration <- m_seg$psi["psi1.Age", "Est."] #36.5. # male max resultant force generally stronger than females.
breakpoint_se_duration <- m_seg$psi["psi1.Age", "St.Err"] #36.5. # male max resultant force generally stronger than females.


breakpoint_duration
breakpoint_se_duration
```
##### aic comparison
```{r}

model_results <- compare_models_pmin(1000*duration, breakpoint_duration, force_clean)

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_duration <- model_results[[4]]$m5 # in theory could
m_duration


```
##### model assumptions
```{r}

model <- m_duration
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)



```
##### test
```{r}

model <- m_duration

tab_model(model)
anova_duration <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>%
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_duration

linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
linearHypothesis(model, "Age = Age+hinge", test = "F")

#before slope
print(deltaMethod(model, "Age + hinge"))
#after slope
print(deltaMethod(model, "Age"))

```
##### plot
```{r}

output <- plot_final(breakpoint_duration, breakpoint_se_duration, 
                     m_duration, 
                     force_clean,
                     plot_age_duration,
                     include_brood = TRUE,
                     brood_level = "orangeR-pinkR",
                     sex = "M"
                     )
plot_duration_final <- output[1]
plot_duration_final

```


#### left duration - 40
##### breakpoint detection
```{r}
m_lin <- lm(1000*left_duration ~ Age*Sex + Brood, data = force_clean)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_left_duration <- m_seg$psi["psi1.Age", "Est."] #36.5. # male max resultant force generally stronger than females.
breakpoint_se_left_duration <- m_seg$psi["psi1.Age", "St.Err"]
slope_early <- slope(m_seg)$Age[1, "Est."]
slope_late  <- slope(m_seg)$Age[2, "Est."]

breakpoint_left_duration
breakpoint_se_left_duration
slope_early
slope_late
```
##### aic comparison
```{r}

model_results <- compare_models_pmin(1000*left_duration, breakpoint_left_duration, force_clean)

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_left_duration <- model_results[[4]]$m5 # in theory could
m_left_duration
```
##### model assumption
```{r}

model <- m_left_duration
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)
```
##### test
```{r}

model <- m_left_duration

tab_model(model)
anova_left_duration <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_left_duration # CHANGE THIS!!

linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
linearHypothesis(model, "Age = Age+hinge", test = "F")

#before slope
print(deltaMethod(model, "Age + hinge"))
#after slope
print(deltaMethod(model, "Age"))

```
##### plot
```{r}

output <- plot_final(breakpoint_left_duration, breakpoint_se_left_duration, #CHANGE HERE
                     m_left_duration, # CHANGE HERE
                     force_clean,
                     plot_left_duration, # CHANGE HERE
                     include_brood = TRUE,
                     brood_level = "Box11",
                     sex = "M"
                     )
plot_left_duration_final <- output[1] #CHANGE HERE
plot_left_duration_final # CHANGE HERE

```
#### right duration -33.1
##### breakpoint detection
```{r}
m_lin <- lm(1000*right_duration ~ Age*Sex + Brood, data = force_clean) # CHANGE HERE
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_right_duration <- m_seg$psi["psi1.Age", "Est."] 
breakpoint_se_right_duration <- m_seg$psi["psi1.Age", "St.Err"] 

slope_early <- slope(m_seg)$Age[1, "Est."]
slope_late  <- slope(m_seg)$Age[2, "Est."]

breakpoint_right_duration
breakpoint_se_right_duration
slope_early
slope_late
```
##### aic comparison
```{r}

model_results <- compare_models_pmin(1000*right_duration, breakpoint_right_duration, force_clean)

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_right_duration <- model_results[[4]]$m5 # in theory could
m_right_duration

```

##### model assumptions
```{r}
model <- m_right_duration
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)

```

##### test
```{r}
model <- m_right_duration

tab_model(model)
anova_right_duration <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_right_duration # CHANGE THIS!!

linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
linearHypothesis(model, "Age = Age+hinge", test = "F")

#before slope
print(deltaMethod(model, "Age + hinge"))
#after slope
print(deltaMethod(model, "Age"))

```

##### plot
```{r}
output <- plot_final(breakpoint_right_duration, breakpoint_se_right_duration, #CHANGE HERE
                     m_right_duration, # CHANGE HERE
                     force_clean,
                     plot_right_duration, # CHANGE HERE
                     include_brood = TRUE,
                     brood_level = "Box5",
                     sex = "M"
                     )
plot_right_duration_final <- output[1] #CHANGE HERE
plot_right_duration_final # CHANGE HERE

```




### TRAJECTORY
#### Z velocity - 32.7
##### breakpoint duration
```{r}

m_lin <- lm(mean_vz ~ Age*Sex + Brood, data = flights_speed) #CHANGE HERE
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_vz <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
breakpoint_se_vz <- m_seg$psi["psi1.Age", "St.Err"] # breakpoint se.


breakpoint_vz #32.9
breakpoint_se_vz


```
##### aic comparison
```{r}

model_results <- compare_models_pmin(mean_vz, breakpoint_vz, flights_speed) #CHANGE

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_vz <- model_results[[4]]$m10 # in theory could
m_vz

```
##### model assumptions
```{r}

model <- m_vz
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)



```
##### test
```{r}

model <- m_vz

tab_model(model)
anova_vz <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_vz # CHANGE THIS!!

linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
linearHypothesis(model, "Age = Age+hinge", test = "F")

#before slope
print(deltaMethod(model, "Age + hinge"))
#after slope
print(deltaMethod(model, "Age"))

```
##### plot
```{r}

output <- plot_final(breakpoint_vz, breakpoint_se_vz, #CHANGE HERE
                     m_vz, # CHANGE HERE
                     flights_speed, # CHANGE HERE
                     plot_vz_mean, # CHANGE HERE
                     include_brood = FALSE,
                     brood_level = "Box5",
                     sex = "M"
                     )
plot_vz_final <- output[1] #CHANGE HERE
plot_vz_final # CHANGE HERE


```
#### Y velocity - 33.0
##### breakpoint duration
```{r}

m_lin <- lm(mean_vy ~ Age*Sex + Brood, data = flights_speed) #CHANGE HERE
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_vy<- m_seg$psi["psi1.Age", "Est."] 
breakpoint_se_vy <- m_seg$psi["psi1.Age", "St.Err"] 


breakpoint_vy #32.9
breakpoint_se_vy


```
##### aic comparison
```{r}

model_results <- compare_models_pmin(mean_vy, breakpoint_vy, flights_speed) #CHANGE

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_vy <- model_results[[4]]$m10 # in theory could
m_vy

```
##### model assumptions
```{r}

model <- m_vy
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)



```
##### test
```{r}

model <- m_vy

tab_model(model)
anova_vy <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_vy # CHANGE THIS!!

linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
linearHypothesis(model, "Age = Age+hinge", test = "F")

#before slope
print(deltaMethod(model, "Age + hinge"))
#after slope
print(deltaMethod(model, "Age"))

```
##### plot
```{r}

output <- plot_final(breakpoint_vy, breakpoint_se_vy, #CHANGE HERE
                     m_vy, # CHANGE HERE
                     flights_speed, # CHANGE HERE
                     plot_vy_mean, # CHANGE HERE
                     include_brood = FALSE,
                     brood_level = "Box5",
                     sex = "M"
                     )
plot_vy_final <- output[1] #CHANGE HERE
plot_vy_final # CHANGE HERE


```
#### X velocity - 60
##### breakpoint duration
```{r}

m_lin <- lm(mean_vx ~ Age*Sex + Brood, data = flights_speed) #CHANGE HERE
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_vx <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
breakpoint_se_vx <- m_seg$psi["psi1.Age", "St.Err"] # breakpoint se.


breakpoint_vx #32.9
breakpoint_se_vx


```
##### aic comparison
```{r}

model_results <- compare_models_pmin(mean_vx, breakpoint_vx, flights_speed) #CHANGE

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_vx <- model_results[[4]]$m10 # in theory could
m_vx

```
##### model assumptions
```{r}

model <- m_vx
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)



```
##### test
```{r}

model <- m_vx

tab_model(model)
anova_vx <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_vx # CHANGE THIS!!

linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
linearHypothesis(model, "Age = Age+hinge", test = "F")

#before slope
print(deltaMethod(model, "Age + hinge"))
#after slope
print(deltaMethod(model, "Age"))

```
##### plot
```{r}

output <- plot_final(breakpoint_vx, breakpoint_se_vx, #CHANGE HERE
                     m_vx, # CHANGE HERE
                     flights_speed, # CHANGE HERE
                     plot_vx_mean, # CHANGE HERE
                     include_brood = FALSE,
                     brood_level = "Box5",
                     sex = "M"
                     )
plot_vx_final <- output[1] #CHANGE HERE
plot_vx_final # CHANGE HERE


```
#### X acceleration -75.1
##### breakpoint duration
```{r}

m_lin <- lm(mean_acc_x ~ Age*Sex + Brood, data = flights_speed) #CHANGE HERE
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_accx <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
breakpoint_se_accx <- m_seg$psi["psi1.Age", "St.Err"] # breakpoint se.


breakpoint_accx #32.9
breakpoint_se_accx


```
##### aic comparison
```{r}

model_results <- compare_models_pmin(mean_acc_x, breakpoint_accx, flights_speed) #CHANGE

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_accx <- model_results[[4]]$m8 # in theory could
m_accx

```
##### model assumptions
```{r}

model <- m_accx
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)



```
##### test
```{r}

model <- m_accx

tab_model(model)
anova_accx <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_vx # CHANGE THIS!!

linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
linearHypothesis(model, "Age = Age+hinge", test = "F")

#before slope
print(deltaMethod(model, "Age + hinge"))
#after slope
print(deltaMethod(model, "Age"))

```
##### plot
```{r}

output <- plot_final(breakpoint_accx, breakpoint_se_accx, #CHANGE HERE
                     m_accx, # CHANGE HERE
                     flights_speed, # CHANGE HERE
                     plot_accx_mean, # CHANGE HERE
                     include_brood = FALSE,
                     brood_level = "Box5",
                     sex = "M"
                     )
plot_accx_final <- output[1] #CHANGE HERE
plot_accx_final # CHANGE HERE


```

#### Y acceleration -33.7
##### breakpoint duration
```{r}

m_lin <- lm(mean_acc_y ~ Age*Sex + Brood, data = flights_speed) #CHANGE HERE
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_accy <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
breakpoint_se_accy <- m_seg$psi["psi1.Age", "St.Err"] # breakpoint se.


breakpoint_accy #32.9
breakpoint_se_accy


```
##### aic comparison
```{r}

model_results <- compare_models_pmin(mean_acc_y, breakpoint_accy, flights_speed) #CHANGE

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_accy <- model_results[[4]]$m6 # in theory could
m_accy

```
##### model assumptions
```{r}

model <- m_accy
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)



```
##### test
```{r}

model <- m_accy

tab_model(model)
anova_accy <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_accy # CHANGE THIS!!

linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
linearHypothesis(model, "Age = Age+hinge", test = "F")

#before slope
print(deltaMethod(model, "Age + hinge"))
#after slope
print(deltaMethod(model, "Age"))

```
##### plot
```{r}

output <- plot_final(breakpoint_accy, breakpoint_se_accy, #CHANGE HERE
                     m_accy, # CHANGE HERE
                     flights_speed, # CHANGE HERE
                     plot_accy_mean, # CHANGE HERE
                     include_brood = FALSE,
                     brood_level = "Box5",
                     sex = "M"
                     )
plot_accy_final <- output[1] #CHANGE HERE
plot_accy_final # CHANGE HERE


```
#### Z acceleration -34.2
##### breakpoint duration
```{r}

m_lin <- lm(mean_acc_z ~ Age*Sex + Brood, data = flights_speed) #CHANGE HERE
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_accz <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
breakpoint_se_accz <- m_seg$psi["psi1.Age", "St.Err"] # breakpoint se.


breakpoint_accz #32.9
breakpoint_se_accz


```
##### aic comparison
```{r}

model_results <- compare_models_pmin(mean_acc_z, breakpoint_accz, flights_speed) #CHANGE

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_accz <- model_results[[4]]$m6 # in theory could
m_accz

```
##### model assumptions
```{r}

model <- m_accz
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)



```
##### test
```{r}

model <- m_accz

tab_model(model)
anova_accz <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_accz# CHANGE THIS!!

linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
linearHypothesis(model, "Age = Age+hinge", test = "F")

#before slope
print(deltaMethod(model, "Age + hinge"))
#after slope
print(deltaMethod(model, "Age"))

```
##### plot
```{r}

output <- plot_final(breakpoint_accz, breakpoint_se_accz, #CHANGE HERE
                     m_accz, # CHANGE HERE
                     flights_speed, # CHANGE HERE
                     plot_accz_mean, # CHANGE HERE
                     include_brood = FALSE,
                     brood_level = "Box5",
                     sex = "M"
                     )
plot_accz_final <- output[1] #CHANGE HERE
plot_accz_final # CHANGE HERE


```
#### tangential acceleration - 33.9
##### breakpoint duration
```{r}

m_lin <- lm(mean_acc_res ~ Age*Sex + Brood, data = flights_speed) #CHANGE HERE
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_tacc <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
breakpoint_se_tacc <- m_seg$psi["psi1.Age", "St.Err"] # breakpoint se.


breakpoint_tacc #32.9
breakpoint_se_tacc


```
##### aic comparison
```{r}

model_results <- compare_models_pmin(mean_acc_res, breakpoint_tacc, flights_speed) #CHANGE

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_tacc <- model_results[[4]]$m10 # in theory could
m_tacc

```
##### model assumptions
```{r}

model <- m_tacc
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)



```
##### test
```{r}

model <- m_tacc

tab_model(model)
anova_tacc <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_tacc # CHANGE THIS!!

linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
linearHypothesis(model, "Age = Age+hinge", test = "F")

#before slope
print(deltaMethod(model, "Age + hinge"))
#after slope
print(deltaMethod(model, "Age"))

```
##### plot
```{r}

output <- plot_final(breakpoint_tacc, breakpoint_se_tacc, #CHANGE HERE
                     m_tacc, # CHANGE HERE
                     flights_speed, # CHANGE HERE
                     plot_accres_mean, # CHANGE HERE
                     include_brood = FALSE,
                     brood_level = "Box5",
                     sex = "M"
                     )
plot_tacc_final <- output[1] #CHANGE HERE
plot_tacc_final # CHANGE HERE


```

#### velocity magnitude - 33.5
##### breakpoint duration
```{r}

m_lin <- lm(mean_v_res ~ Age*Sex + Brood, data = flights_speed) #CHANGE HERE
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_vres <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
breakpoint_se_vres <- m_seg$psi["psi1.Age", "St.Err"] # breakpoint se.


breakpoint_vres #32.9
breakpoint_se_vres


```
##### aic comparison
```{r}

model_results <- compare_models_pmin(mean_v_res, breakpoint_vres, flights_speed) #CHANGE

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_vres <- model_results[[4]]$m10 # in theory could
m_vres

```
##### model assumptions
```{r}

model <- m_vres
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)



```
##### test
```{r}

model <- m_vres

tab_model(model)
anova_vres <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_vres # CHANGE THIS!!

linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
linearHypothesis(model, "Age = Age+hinge", test = "F")

#before slope
print(deltaMethod(model, "Age + hinge"))
#after slope
print(deltaMethod(model, "Age"))

```
##### plot
```{r}

output <- plot_final(breakpoint_vres, breakpoint_se_vres, #CHANGE HERE
                     m_vres, # CHANGE HERE
                     flights_speed, # CHANGE HERE
                     plot_vres_mean, # CHANGE HERE
                     include_brood = FALSE,
                     brood_level = "Box5",
                     sex = "M"
                     )
plot_vres_final <- output[1] #CHANGE HERE
plot_vres_final # CHANGE HERE


```

#### turn rate - 40.0
##### breakpoint duration
```{r}

m_lin <- lm(mean_turn_rate ~ Age*Sex + Brood, data = flights_tort) #CHANGE HERE
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_tr <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
breakpoint_se_tr <- m_seg$psi["psi1.Age", "St.Err"] # breakpoint se.


breakpoint_tr #32.9
breakpoint_se_tr


```
##### aic comparison
```{r}

model_results <- compare_models_pmin(mean_turn_rate, breakpoint_tr, flights_speed) #CHANGE

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_tr <- model_results[[4]]$m7 # in theory could
m_tr

```
##### model assumptions
```{r}

model <- m_tr
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)



```
##### test
```{r}

model <- m_tr

tab_model(model)
anova_tr <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_tr # CHANGE THIS!!

linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
linearHypothesis(model, "Age = Age+hinge", test = "F")

#before slope
print(deltaMethod(model, "Age + hinge"))
#after slope
print(deltaMethod(model, "Age"))

```
##### plot
```{r}

output <- plot_final(breakpoint_tr, breakpoint_se_tr, #CHANGE HERE
                     m_tr, # CHANGE HERE
                     flights_speed, # CHANGE HERE
                     plot_tr_mean, # CHANGE HERE
                     include_brood = TRUE,
                     brood_level = "Box3",
                     sex = "M"
                     )
plot_tr_final <- output[1] #CHANGE HERE
plot_tr_final # CHANGE HERE


```




















#### impulse ratio - not significant!!
##### breakpoint detection
```{r}

m_lin <- lm(impulse_ratio ~ Age*Sex + Brood, data = force_clean) #CHANGE HERE
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_impulse_ratio <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
breakpoint_se_impulse_ratio <- m_seg$psi["psi1.Age", "St.Err"] # breakpoint se.


breakpoint_impulse_ratio #32.9
breakpoint_se_impulse_ratio


```
##### aic comparison
```{r}

model_results <- compare_models_pmin(impulse_ratio, breakpoint_impulse_ratio, force_clean) #CHANGE

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_impulse_ratio <- model_results[[4]]$m10 # in theory could
m_impulse_ratio

```
##### model assumptions
```{r}

model <- m_impulse_ratio
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)



```
##### test
```{r}

model <- m_impulse_ratio

tab_model(model)
anova_impulse_ratio <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_impulse_ratio # CHANGE THIS!!

linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
linearHypothesis(model, "Age = Age+hinge", test = "F")

#before slope
print(deltaMethod(model, "Age + hinge"))
#after slope
print(deltaMethod(model, "Age"))

```
##### plot
```{r}

output <- plot_final(breakpoint_impulse_ratio, breakpoint_se_impulse_ratio, #CHANGE HERE
                     m_impulse_ratio, # CHANGE HERE
                     force_clean, # CHANGE HERE
                     plot_age_impulse_ratio, # CHANGE HERE
                     include_brood = TRUE,
                     brood_level = "Box3",
                     sex = "M"
                     )
plot_impulse_ratio_final <- output[1] #CHANGE HERE
plot_impulse_ratio_final # CHANGE HERE


```

#### mean centripetal acceleration
##### breakpoint duration
```{r}

m_lin <- lm(mean_centri_acc ~ Age*Sex + Brood, data = flights_tort) #CHANGE HERE
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_centri <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
breakpoint_se_centri <- m_seg$psi["psi1.Age", "St.Err"] # breakpoint se.


breakpoint_centri #32.9
breakpoint_se_centri


```
##### aic comparison
```{r}

model_results <- compare_models_pmin(mean_centri_acc, breakpoint_centri, flights_tort) #CHANGE

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_centri <- model_results[[4]]$m5 # in theory could
m_centri

```
##### model assumptions
```{r}

model <- m_centri
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)



```
##### test
```{r}

model <- m_centri

tab_model(model)
anova_centri <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_centri # CHANGE THIS!!

linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
linearHypothesis(model, "Age = Age+hinge", test = "F")

#before slope
print(deltaMethod(model, "Age + hinge"))
#after slope
print(deltaMethod(model, "Age"))

```
##### plot
```{r}

output <- plot_final(breakpoint_centri, breakpoint_se_centri, #CHANGE HERE
                     m_centri, # CHANGE HERE
                     flights_speed, # CHANGE HERE
                     plot_ctracc_mean, # CHANGE HERE
                     include_brood = TRUE,
                     brood_level = "Box11",
                     sex = "M"
                     )
plot_centri_final <- output[1] #CHANGE HERE
plot_centri_final # CHANGE HERE


```
#### wingbeat frequency 
##### breakpoint detection = 40.0
```{r}

m_lin <- lm(wingbeat_frequency ~ Age*Sex + Brood, data = flights_wing) #CHANGE HERE
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_wb <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
breakpoint_se_wb <- m_seg$psi["psi1.Age", "St.Err"] # breakpoint se.


breakpoint_tr #32.9
breakpoint_se_wb


```
##### aic comparison
```{r}

model_results <- compare_models_pmax(wingbeat_frequency, breakpoint_wb, flights_wing) #CHANGE

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_wb <- model_results[[4]]$m5 # in theory could
m_wb

```
##### model assumptions
```{r}

model <- m_wb
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)



```
##### test
```{r}

model <- m_wb

tab_model(model)
anova_wb <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_wb # CHANGE THIS!!

linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
linearHypothesis(model, "Age = Age+hinge", test = "F")

#before slope
print(deltaMethod(model, "Age + hinge"))
#after slope
print(deltaMethod(model, "Age"))

```
##### plot
```{r}

output <- plot_final(breakpoint_wb, breakpoint_se_wb, #CHANGE HERE
                     m_wb, # CHANGE HERE
                     flights_wing, # CHANGE HERE
                     plot_wb_mean, # CHANGE HERE
                     include_brood = TRUE,
                     brood_level = "WhitePinkR-BlueR",
                     sex = "M"
                     )
plot_wb_final <- output[1] #CHANGE HERE
plot_wb_final # CHANGE HERE


```
#### downstroke
##### breakpoint detection = 40.0
```{r}

m_lin <- lm(1000*downstroke_amplitude_m ~ Age*Sex + Brood, data = flights_wing) #CHANGE HERE
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_downstroke <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
breakpoint_se_downstroke <- m_seg$psi["psi1.Age", "St.Err"] # breakpoint se.


breakpoint_downstroke #32.9
breakpoint_se_downstroke


```
##### aic comparison
```{r}

model_results <- compare_models_pmin(1000*downstroke_amplitude_m, breakpoint_downstroke, flights_wing) #CHANGE

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_downstroke <- model_results[[4]]$m5 # in theory could
m_downstroke

```
##### model assumptions
```{r}

model <- m_downstroke
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)



```
##### test
```{r}

model <- m_downstroke

tab_model(model)
anova_downstroke <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_downstroke # CHANGE THIS!!

linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
linearHypothesis(model, "Age = Age+hinge", test = "F")

#before slope
print(deltaMethod(model, "Age + hinge"))
#after slope
print(deltaMethod(model, "Age"))

```
##### plot
```{r}

output <- plot_final(breakpoint_downstroke, breakpoint_se_downstroke, #CHANGE HERE
                     m_downstroke, # CHANGE HERE
                     flights_wing, # CHANGE HERE
                     plot_downstroke_mm, # CHANGE HERE
                     include_brood = TRUE,
                     brood_level = "Box5",
                     sex = "M"
                     )
plot_downstroke_final <- output[1] #CHANGE HERE
plot_downstroke_final # CHANGE HERE


```
#### upstroke - 60 
##### breakpoint detection = 32.0
```{r}

m_lin <- lm(1000*upstroke_amplitude_m ~ Age*Sex + Brood, data = flights_wing) #CHANGE HERE
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_upstroke <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
breakpoint_se_upstroke <- m_seg$psi["psi1.Age", "St.Err"] # breakpoint se.


breakpoint_upstroke #32.9
breakpoint_se_upstroke


```
##### aic comparison
```{r}

model_results <- compare_models_pmin(1000*upstroke_amplitude_m, breakpoint_upstroke, flights_wing) #CHANGE

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_upstroke <- model_results[[4]]$m5 # in theory could
m_upstroke

```
##### model assumptions
```{r}

model <- m_upstroke
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)



```
##### test
```{r}

model <- m_upstroke

tab_model(model)
anova_upstroke <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_upstroke # CHANGE THIS!!

linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
linearHypothesis(model, "Age = Age+hinge", test = "F")

#before slope
print(deltaMethod(model, "Age + hinge"))
#after slope
print(deltaMethod(model, "Age"))

```
##### plot
```{r}

output <- plot_final(breakpoint_upstroke, breakpoint_se_upstroke, #CHANGE HERE
                     m_upstroke, # CHANGE HERE
                     flights_wing, # CHANGE HERE
                     plot_upstroke_mm, # CHANGE HERE
                     include_brood = TRUE,
                     brood_level = "Box5",
                     sex = "M"
                     )
plot_upstroke_final <- output[1] #CHANGE HERE
plot_upstroke_final # CHANGE HERE


```
### MODEL OUTPUT
#### function fmt_term: F = xx.xx (Df, Df.res) (p-display)
```{r}
# Helper to format one term: F = xx.xx (Df, Df.res) (p-display)
fmt_term <- function(F, Df, Df_res, p, digits_F = 1, digits_Df = 1, p_thresh = 0.001) {
  if (anyNA(c(F, Df, Df_res, p))) return(NA_character_)
  F_str   <- format(round(F, digits_F), nsmall = digits_F, trim = TRUE)
  Df_str  <- format(Df, trim = TRUE)
  Res_str <- format(round(Df_res, digits_Df), nsmall = digits_Df, trim = TRUE)
  
  p_str <- if (is.na(p)) {
      "p = NA"
    } else if (p < p_thresh) {
      "p < 0.001"
    } else {
      paste0("p = ", format(round(p, 2), nsmall = 2))
    }

  sprintf("F<sub>%s, %s</sub> = %s, %s",  Df_str, Res_str, F_str, p_str)

  }
```

```{r}
library(knitr)

labels <- c("scaled_weight", "wingspan", "wingarea", "Fx/bw (Fz in data)", "Fy/bw", "Ftot/bw", "angle", "impulse", "left impulse", "right impulse", "duration", "left_duration", "right_duration")
anova_tabs <- list(anova_sw, anova_wingspan, anova_wingarea, anova_Fx, anova_Fy, anova_Ftot, anova_angle, anova_impulse, anova_left_impulse, anova_right_impulse, anova_duration, anova_left_duration, anova_right_duration) #anova dataframe
lmer_mods <- list(m_sw, m_wingspan, m_wingarea, m_Fx, m_Fy, m_Ftot, m_angle, m_impulse, m_left_impulse, m_right_impulse, m_duration, m_left_duration, m_right_duration)             # fitted lmer models

#stop if the number of anova tabs != number of models. It should
stopifnot(length(labels) == length(anova_tabs), length(anova_tabs) == length(lmer_mods))

#storage for all rows
out_all <- list()


for (i in seq_along(lmer_mods)){
  
  tab <- anova_tabs[[i]] # the ANOVA data.frame for term extraction
  mod <- lmer_mods[[i]] # the fitted model for summary metrics

  
# --- helper to extract a term safely from the current model ---
  get_term <- function(term, what) {
    # if the term is not present, return NA
    if (!term %in% rownames(tab)) return(NA_real_)
    val <- tab[term, ][[what]]
    if (is.null(val)) return(NA_real_)
    val
  }

  # pull values for each term from *this* model
  ageF      <- get_term("Age",   "F")
  ageDf     <- get_term("Age",   "Df")
  ageDf.res <- get_term("Age",   "Df.res")
  ageP      <- get_term("Age",   "p_FDR")

  SexF      <- get_term("Sex",   "F")
  SexDf     <- get_term("Sex",   "Df")
  SexDf.res <- get_term("Sex",   "Df.res")
  SexP      <- get_term("Sex",   "p_FDR")

  hingeF      <- get_term("hinge", "F")
  hingeDf     <- get_term("hinge", "Df")
  hingeDf.res <- get_term("hinge", "Df.res")
  hingeP      <- get_term("hinge", "p_FDR")
  
  # Brood is optional — if missing, all will be NA
  BroodF      <- get_term("Brood",   "F")
  BroodDf     <- get_term("Brood",   "Df")
  BroodDf.res <- get_term("Brood",   "Df.res")
  BroodP      <- get_term("Brood",   "p_FDR")
  
  #Optional interaction
  AgeSexF      <- get_term("Age:Sex",   "F")
  AgeSexDf     <- get_term("Age:Sex",   "Df")
  AgeSexDf.res <- get_term("Age:Sex",   "Df.res")
  AgeSexP      <- get_term("Age:Sex",   "p_FDR")
  
  # Build the single-row data frame in your requested format ---
  out_row <- data.frame(
    Age   = fmt_term(ageF,   ageDf,   ageDf.res,   ageP),
    Sex   = fmt_term(SexF,   SexDf,   SexDf.res,   SexP),
    Hinge = fmt_term(hingeF, hingeDf, hingeDf.res,  hingeP),# uses the parenthesis style
    Brood = fmt_term(BroodF, BroodDf, BroodDf.res, BroodP),
    AgeSex = fmt_term(AgeSexF, AgeSexDf, AgeSexDf.res, AgeSexP) # uses "p = xx" style
    
    )
  
  # model summary variables
  
  var <- as.data.frame(VarCorr(mod))
  
  if (dim(var)[1] > 2){
    sd_bird <- var$sdcor[1] 
    sd_brood <- var$sdcor[2]
    sd_fixed <- var$sdcor[3]
    ICC <-  (var$sdcor[1] + var$sdcor[2]) / (var$sdcor[1] + var$sdcor[2] +var$sdcor[3])  # how much % of variation is due to individual variation nested in brood?
  }else{
    sd_bird <- var$sdcor[1]
    sd_brood <- NA
    sd_fixed <- var$sdcor[2]
    ICC <- var$sdcor[1] / (var$sdcor[1] + var$sdcor[2]) # how much % of variation is due to individual variation?
  }
  
  
  N_Bird <- ngrps(mod)[1]
  Observations <- stats::nobs(mod)
  R2_marginal <- performance::r2(mod, tolerance = 0 )$R2_marginal
  R2_conditional <- performance::r2(mod, tolerance = 0 )$R2_conditional

  summary_cols <- data.frame(
    sd_fixed   = round(sd_fixed, 2),
    sd_bird    = round(sd_bird, 2),
    sd_brood = ifelse(is.na(sd_brood), NA, round(sd_brood,2 )),
    ICC           = round(ICC, 2),
    N_Bird        = N_Bird,
    Observations  = Observations,
    `Marginal R2 / Conditional R2` = paste0(round(R2_marginal,3), " / ", round(R2_conditional,3)),
    check.names = FALSE, 
    row.names = NULL
  )

  
  # Add the Variables column as the first column
  out_row <- cbind(Variables = labels[i], out_row, summary_cols)
  
  out_all[[i]] <- out_row
}

# Combine rows from all models into one data frame
out_all <- do.call(rbind, out_all)

kable(out_all, escape = FALSE)

```
### TEST OUTPUT
```{r}

library(knitr)

labels <- c("scaled_weight", "wingspan", "wingarea", "Fx/bw (Fz in data)", "Fy/bw", "Ftot/bw", "angle", "impulse", "left impulse", "right impulse", "duration", "left_duration", "right_duration")

lmer_mods <- list(m_sw, m_wingspan, m_wingarea, m_Fx, m_Fy, m_Ftot, m_angle, m_impulse, m_left_impulse, m_right_impulse, m_duration, m_left_duration, m_right_duration) 

out_all <- vector("list", length(lmer_mods))


for (i in seq_along(lmer_mods)){
  
  #tab <- anova_tabs[[i]] # the ANOVA data.frame for term extraction
  mod <- lmer_mods[[i]] # the fitted model for summary metrics

  #before slope = 0?
  before_slope <- linearHypothesis(mod, "Age+hinge = 0", test = "F") %>% as.data.frame()# Is the slope before the breakpoint flat? -> no
  
  #after slope = 0?
  after_slope <- linearHypothesis(mod, "Age = 0", test = "F") %>% as.data.frame() # Is the slope after the breakpoint flat? -> no
  
  #same slope?
  same_slope <- linearHypothesis(mod, "Age = Age+hinge", test = "F") %>% as.data.frame()#Is the slope before and after the same? -> n
  
  #before breakpoint slope estimate
  before_slope_estimate <- deltaMethod(mod, "Age + hinge") %>% as.data.frame()
  
  #after breakpoint slope estimate
  after_slope_estimate <- deltaMethod(mod, "Age") %>% as.data.frame()
  
  before_slope_mean <-  before_slope_estimate$Estimate 
  before_slope_se <-  before_slope_estimate$SE
  before_slope_lower <- before_slope_estimate$`2.5 %`
  before_slope_upper <- before_slope_estimate$`97.5 %`
  
  before_slope_F <- before_slope$F[2]
  before_slope_Df <- before_slope$Df[2]
  before_slope_ResDf <- before_slope$Res.Df[2]
  before_slope_P <- before_slope$`Pr(>F)`[2]
  
  after_slope_mean <-  after_slope_estimate$Estimate 
  after_slope_se <-  after_slope_estimate$SE
  after_slope_lower <- after_slope_estimate$`2.5 %`
  after_slope_upper <- after_slope_estimate$`97.5 %`
  
  after_slope_F <- after_slope$F[2]
  after_slope_Df <- after_slope$Df[2]
  after_slope_ResDf <- after_slope$Res.Df[2]
  after_slope_P <- after_slope$`Pr(>F)`[2]
  
  same_slope_F <- same_slope$F[2]
  same_slope_Df <- same_slope$Df[2]
  same_slope_ResDf <- same_slope$Res.Df[2]
  same_slope_P <- same_slope$`Pr(>F)`[2]
  
  
  before_string <- sprintf(
    "%.2f ± %.2f [%.2f, %.2f]",
    before_slope_mean,
    before_slope_se,
    before_slope_lower,
    before_slope_upper)
  
  after_string <- sprintf(
    "%.2f ± %.2f [%.2f, %.2f]",
    after_slope_mean,
    after_slope_se,
    after_slope_lower,
    after_slope_upper)
  
  
  # test summary variables

  test_cols <- data.frame(
    "slope before Mean ± SE [CI]" = before_string,
    "slope before = 0?" = fmt_term(before_slope_F, before_slope_Df, before_slope_ResDf, before_slope_P),
    "slope after Mean ± SE [CI]" = after_string,
    "slope after = 0?" = fmt_term(after_slope_F, after_slope_Df, after_slope_ResDf, after_slope_P),
    "slope before = after?" = fmt_term(same_slope_F, same_slope_Df, same_slope_ResDf, same_slope_P),
    check.names = FALSE, 
    row.names = NULL
  )

  
  # Add the Variables column as the first column
  out_row <- cbind(Variables = labels[i], test_cols)
  
  out_all[[i]] <- out_row
}

# Combine rows from all models into one data frame
out_all_test <- do.call(rbind, out_all)

kable(out_all_test, escape = FALSE)
```
