---
title: "zebra-force"
author: "Kyu Min Huh"
date: "2025-11-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Aims 

In this markdown, I aim to see the age effect on the leg force at take-off. 
Aims/Steps:

1. show morphological trend with age
- Graphs: a. body mass x age, b. wing area x age, c. wing chord length x age, d. tarsus length x age. 

2. See if there's any visible trend in raw data with age
- Graphs: a. max resultant force x age, with each individual marked by point and whiskers b. horizontal force at the time of max resultant force x age c. vertical force at the time of max resultant force x age d. impulse x age e. impulse ratio x age
f. take-off angle x age 

# Libraries
```{r libraries}
library(dplyr)
library(tidyverse)
library(ggplot2)

library(rptR)
library(ggpubr)
library(segmented)

library(nlme)
library(lme4)
library(lmerTest)
library(emmeans)
library(PerformanceAnalytics)
library(MuMIn)
library(lmtest)
library(lmerTest)
library(AICcmodavg)
library(devtools)
library(redres)
library(robustlmm)
library(car)
library(ggeffects)

```

# I. Raw data plots (mean+-SD)
## load the dataset 
### force_data
```{r load the datasets}

# force data
force_data <- read.delim(file = "C:\\Users\\jesu4036\\Documents\\zebra-finches-fly\\data\\takeoff\\force_final.csv", sep = ",") %>% distinct()

dim(force_data)


# changing the unit to mN
# more suitable for birds that weigh in grams not kilograms
#force_data$max_Ftot = force_data$max_Ftot*1000

# newton to gram conversion
force_data$bodyweight_g = force_data$bodyweight * 101.9716

# make relative max res force
force_data$max_Ftot_bw = force_data$max_Ftot/force_data$bodyweight

# other relative force as well 

force_data <- force_data %>%
  mutate(angle = as.numeric(angle)) %>%
  mutate(Fx_maxFtot = as.numeric(Fx_maxFtot)) %>%
  mutate(Fy_maxFtot = as.numeric(Fy_maxFtot)) %>%
  mutate(Fz_maxFtot = as.numeric(Fz_maxFtot)) %>%
  mutate(Fx_maxFtot_bw = abs(Fx_maxFtot)/bodyweight) %>%
  mutate(Fy_maxFtot_bw = abs(Fy_maxFtot)/bodyweight) %>%
  mutate(Fz_maxFtot_bw = abs(Fz_maxFtot)/bodyweight)



# 'Age': change 81->80 and 61->60 

force_data$Age[force_data$Age == 81] <- 80
force_data$Age[force_data$Age == 61] <- 60
```

### wing area data
```{r}
wing_data <- read.delim(file = "C:\\Users\\jesu4036\\Documents\\flight-ontogeny\\data\\morphology\\wingdata_names_edited.csv", sep = ",") %>% distinct()

scale_weight <- read.delim(file = "C:\\Users\\jesu4036\\Documents\\flight-ontogeny\\data\\morphology\\bodyweight_measured.csv", sep = ",")


wing_data_sorted <- wing_data %>%
  # sort by Bird (alphabetical) and Age (increasing)
  arrange(Bird, Age) %>%
  # keep only birds with at least 3 age records
  group_by(Bird) %>%
  filter(n_distinct(Age) >= 3) %>%
  mutate(wingarea = wingarea/10000) %>% #converting mm^2 -> m^2
  ungroup() %>%
  left_join(scale_weight, by = c("Bird", "Age")) %>%
  mutate(bodyweight_measured = as.numeric(bodyweight_measured)) %>%
  mutate(bodyweight_measured = bodyweight_measured * 0.00980665) %>%
  mutate(wing_loading = wingarea/bodyweight_measured) %>%
  dplyr::select('Bird','Age','wingspan.calc','wingarea','bodyweight_measured','wing_loading')

dim(wing_data_sorted)

wing_data_final <- wing_data_sorted %>%
  group_by(Bird, Age) %>%
  summarise(
    wingspan = mean(wingspan.calc, na.rm = TRUE),
    wingarea = mean(wingarea, na.rm = TRUE),
    scale_weight = mean(bodyweight_measured, na.rm = TRUE),
    wing_loading = mean(wing_loading, na.rm = TRUE),
    .groups = "drop"
  ) %>% drop_na()

dim(wing_data_final) #192

```

### sex and brood measured bodyweight data 
```{r}
sex_data <- read.delim(file = "C:\\Users\\jesu4036\\Documents\\flight-ontogeny\\data\\morphology\\birds_sex_brood.csv", sep = ",") %>% distinct()

force_data <- merge(force_data, sex_data, by = "Bird")
wing_data_final <- merge(wing_data_final, sex_data, by = "Bird")


```

### flight trajectory data
```{r}
flights <- read.delim(file = "C:\\Users\\jesu4036\\Documents\\flight-ontogeny\\data\\takeoff\\flight_outcome_full.csv", sep = ",") %>% left_join(sex_data, by = "Bird")

raw_flight <- read.delim(file = "C:\\Users\\jesu4036\\Documents\\flight-ontogeny\\data\\takeoff\\clean_filtered_flights.csv", sep =",")
colnames(flights)
colnames(raw_flight)
```
## one flight
```{r}
#cyanR-blackL, Age 36, trial 1
#test_flight1 <- raw_flight %>%
 #filter(FlightID == 2303601.0)

test_flight <- raw_flight %>%
  filter(FlightID == 2303601.0)
```
## RAW PLOTS
```{r}

library(ggplot2)
library(patchwork)

df <- test_flight2

# Position plots
p1 <- ggplot(df, aes(cumul_time, X)) +
  geom_line(color = "blue") +
  geom_point(color = "darkblue") +
  ggtitle("Position X")

p2 <- ggplot(df, aes(cumul_time, Y)) +
  geom_line(color = "green") +
  geom_point(color = "darkgreen") +
  ggtitle("Position Y")

p3 <- ggplot(df, aes(cumul_time, Z)) +
  geom_line(color = "red") +
  geom_point(color = "darkred") +
  ggtitle("Position Z")

# Velocity plots
p4 <- ggplot(df, aes(cumul_time, inst_V_X)) +
  geom_line(color = "blue") +
  geom_point(color = "darkblue") +
  ggtitle("Velocity X")

p5 <- ggplot(df, aes(cumul_time, inst_V_Y)) +
  geom_line(color = "green") +
  geom_point(color = "darkgreen") +
  ggtitle("Velocity Y")

p6 <- ggplot(df, aes(cumul_time, inst_V_Z)) +
  geom_line(color = "red") +
  geom_point(color = "darkred") +
  ggtitle("Velocity Z")

# Acceleration plots
p7 <- ggplot(df, aes(cumul_time, inst_acc_X)) +
  geom_line(color = "blue") +
  geom_point(color = "darkblue") +
  ggtitle("Acceleration X")

p8 <- ggplot(df, aes(cumul_time, inst_acc_Y)) +
  geom_line(color = "green") +
  geom_point(color = "darkgreen") +
  ggtitle("Acceleration Y")

p9 <- ggplot(df, aes(cumul_time, inst_acc_Z)) +
  geom_line(color = "red") +
  geom_point(color = "darkred") +
  ggtitle("Acceleration Z")


# New 5 plots
p10 <- ggplot(df, aes(cumul_time, inst_V_res)) +
  geom_line(color="purple") + geom_point(color="purple") + ggtitle("Velocity Resultant")

p11 <- ggplot(df, aes(cumul_time, inst_acc_res)) +
  geom_line(color="orange") + geom_point(color="darkorange") + ggtitle("Tangential acceleration")

p12 <- ggplot(df, aes(cumul_time, inst_acc_mag)) +
  geom_line(color="orange") + geom_point(color="darkorange") + ggtitle("Acceleration magnitude")


p13 <- ggplot(df, aes(cumul_time, Turn_Rate_degrees)) +
  geom_line(color="brown") + geom_point(color="brown") + ggtitle("Turn Rate (degrees)")

p14 <- ggplot(df, aes(cumul_time, Centripetal_acc)) +
  geom_line(color="pink") + geom_point(color="deeppink") + ggtitle("Centripetal Acceleration")

# Combine all plots
final_plot2 <- (p1 | p2 | p3) /
              (p4 | p5 | p6) /
              (p7 | p8 | p9) /
              (p10 | p11 | p12)/
              (p13| p14)

final_plot2




```
#### function plot_age_var 
```{r}
plot_age_var <- function(variable, df, title) {
  
  # Summary by individual
  summary_df <- df %>%
    group_by(Bird, Age) %>%
    summarise(
      mean = mean({{variable}}, na.rm = TRUE),
      sd = sd({{variable}}, na.rm = TRUE),
      n = n(),
      .groups = "drop"
    )
  
  # Summary by age
  trend_df <- df %>%
    group_by(Age) %>%
    summarise(
      mean = mean({{variable}}, na.rm = TRUE),
      sd = sd({{variable}}, na.rm = TRUE),
      n = n(),
      .groups = "drop"
    )
  
  # Plot
  plot_age_performance<- ggplot(df, aes(x = Age, y = {{variable}}, color = Bird)) +
    geom_point(position = position_jitter(width = 1), alpha = 0.3) +
    geom_errorbar(
      data = summary_df,
      aes(x = Age, ymin = mean - sd, ymax = mean + sd, color = Bird, alpha = 0.6),
      width = 2, size = 0.8,
      inherit.aes = FALSE
    ) +
    geom_point(
      data = summary_df,
      aes(y = mean, x = Age, color = Bird, alpha = 0.9),
      size = 2, shape = 18,
      inherit.aes = FALSE
    ) +
    geom_line(
      data = summary_df,
      aes(x = Age, y = mean, color = Bird, group = Bird, alpha = 0.8),
      size = 0.7,
      inherit.aes = FALSE
    ) +
    geom_ribbon(
      data = trend_df,
      aes(x = Age, ymin = mean - sd, ymax = mean + sd),
      fill = "grey70", alpha = 0.5,
      inherit.aes = FALSE
    ) +
    geom_line(
      data = trend_df,
      aes(x = Age, y = mean),
      ,color = "grey50", alpha = 0.8, size = 1.2,
      inherit.aes = FALSE
    ) +
    scale_x_continuous(breaks = unique(df$Age)) +
    labs(x = "Age", y = title, title = "") +
    guides(alpha = "none", color = "none") +
    theme(legend.position = "none")+
    theme_minimal()
  
  return(plot_age_performance)
}
```
## A. Morphology
### a. wing area

```{r wing area per bird}

# Summarize mean and standard deviation per Bird & Age
summary_df <- wing_data_final %>%
  group_by(Bird, Age) %>%
  summarise(
    mean_wa = mean(wingarea),
    sd_wa = sd(wingarea),
    .groups = "drop"
  )

trend_df <- wing_data_final %>%
  group_by(Age) %>%
  summarise(
    mean_wa = mean(wingarea, na.rm = TRUE),
    sd_wa   = sd(wingarea, na.rm = TRUE),
    .groups = "drop"
  )

# Plot: trials + mean + SD
plot_wingarea <- ggplot(wing_data_final, aes(x = Age, y = wingarea, color = Bird)) +
  geom_point(position = position_jitter(width = 1), alpha = 0.3) +   # dots = trials
  geom_errorbar(
    data = summary_df,
    aes(x=Age, ymin = mean_wa - sd_wa, ymax = mean_wa + sd_wa, color = Bird, alpha = 0.6),
    width = 2, size = 0.8,
    inherit.aes = FALSE) +
  
  geom_point(
    data = summary_df,
    aes(x = Age, y = mean_wa,  color = Bird, alpha = 0.9),
    size = 3, shape = 18,
    inherit.aes = FALSE
  ) +
  
   geom_line(
    data = summary_df,
    aes(x = Age, y = mean_wa, color = Bird, group = Bird, alpha = 0.8),
    linewidth = 0.7,
    inherit.aes = FALSE) +
  
  geom_ribbon(
    data = trend_df,
    aes(x = Age, ymin = mean_wa - sd_wa, ymax = mean_wa + sd_wa),
    fill = "grey70", alpha = 0.5,
    inherit.aes = FALSE
  ) +
  
  geom_line(
    data = trend_df,
    aes(x = Age, y = mean_wa),
    color = "grey50", alpha = 0.8, size = 1.2,
    inherit.aes = FALSE) +
  
   scale_x_continuous(breaks = unique(wing_data_sorted$Age)) +
  labs(x = "Age", y = "Wing Area (m^2)", title = "") +
  ylim(0.5, 1.0)+
  guides(alpha = "none") +
  theme_minimal() #+
  #theme(legend.position = "none")

plot_wa_age



```
```{r}
plot_wingarea <- plot_age_var(wingarea, wing_data_final, "wing area (N/m^2)")

```


### b. calculated wingspan
```{r}
plot_wingspan <- plot_age_var(wingspan, wing_data_final, "wingspan (mm)")
```

### c. perching weight
```{r}
plot_perch_g <- plot_age_var(bodyweight_g, force_data, "perching weight(g)")
```

### d. scale weight
```{r}
plot_scale_g <- plot_age_performance(scale_weight*101.97, wing_data_final, "scale weight (g)")
ggarrange(plot_perch_g, plot_scale_g, nrow = 1, ncol = 2)
```

### e. wing loading 
```{r}
plot_wingloading <- plot_age_var(wing_loading, wing_data_final, "wing loading (N/m^2)")
```

### PLOTS - wingspan, wing area, scale weight, wing loading 
```{r}

plots_morphology <- ggarrange(plot_wingspan, plot_wingarea, plot_scale_g, plot_wingloading, nrow = 2, ncol =2)

plots_morphology

```
## B. Performance

### LEGS
#### a. max resultant force
```{r}

summary_df <- force_data %>%
  group_by(Bird, Age) %>%
  summarise(
    mean_maxF = mean(max_Ftot),
    sd_maxF   = sd(max_Ftot),
    .groups = "drop"
  ) %>% as.data.frame()

# Reference means at Age 20 for each Bird
ref_summary <- summary_df %>%
  filter(Age == 20) %>%
  dplyr::select(Bird, mean_maxF, sd_maxF) %>%
  dplyr::rename(ref_mean_maxF = mean_maxF, 
         ref_sd_maxF = sd_maxF)

# Centered summary: subtract the bird-specific mean at Age 20
summary_df_init <- summary_df %>%
  left_join(ref_summary, by = "Bird") %>%
  mutate(
    mean_maxF_centered = mean_maxF - ref_mean_maxF,
    # Optionally keep sd as-is; centering doesn't change variability
    sd_maxF_centered   = sd_maxF #SD unchanged
  ) %>%
  arrange(Bird, Age)


trend_df <- force_data %>%
  group_by(Age) %>%
  summarise(
    mean_maxF = mean(max_Ftot, na.rm = TRUE),
    sd_maxF   = sd(max_Ftot, na.rm = TRUE),
    .groups = "drop"
  )


# Extract reference mean at Age 20
ref_mean <- trend_df %>%
  filter(Age == 20) %>%
  pull(mean_maxF)

# Center all means by subtracting the mean at Age 20
trend_df_init <- trend_df %>%
  mutate(
    mean_maxF_centered = mean_maxF - ref_mean,
    sd_maxF_centered   = sd_maxF  # SD stays the same
  )
```
##### raw
```{r}
plot_age_maxFtot <- ggplot(force_data, aes(x = Age, y = max_Ftot, color = Bird)) +
  geom_point(position = position_jitter(width = 1), alpha = 0.3) +   # dots = trials
  geom_errorbar(
    data = summary_df,
    aes(x=Age, ymin = mean_maxF - sd_maxF, ymax = mean_maxF + sd_maxF, color = Bird, alpha = 0.6),
    width = 2, size = 0.8,
    inherit.aes = FALSE) +
  
  geom_point(
    data = summary_df,
    aes(y = mean_maxF, x = Age, color = Bird, alpha = 0.9),
    size = 3, shape = 18,
    inherit.aes = FALSE
  ) +
  
   geom_line(
    data = summary_df,
    aes(x = Age, y = mean_maxF, color = Bird, group = Bird, alpha = 0.8),
    size = 0.7,
    inherit.aes = FALSE) +
  
  geom_ribbon(
    data = trend_df,
    aes(x = Age, ymin = mean_maxF - sd_maxF, ymax = mean_maxF + sd_maxF),
    fill = "grey70", alpha = 0.5,
    inherit.aes = FALSE
  ) +
  
  geom_line(
    data = trend_df,
    aes(x = Age, y = mean_maxF),
    color = "grey50", alpha = 0.8, size = 1.2,
    inherit.aes = FALSE) +
  
   scale_x_continuous(breaks = unique(force_data$Age)) +
  labs(x = "Age", y = "max F (N)", title = "") +
  #ylim(10, 25)+
  guides(alpha = "none")+
  theme_minimal() + 
  theme(legend.position = "none")

plot_age_maxFtot
```
##### centrerd
```{r}
plot_age_maxFtot_centred <- ggplot() +
  
  # error bard for per-bird summary
  geom_errorbar(
    data = summary_df_init,
    aes(x = Age, 
        ymin = mean_maxF_centered - sd_maxF_centered, 
        ymax = mean_maxF_centered + sd_maxF_centered, 
        color = Bird, alpha = 0.6),
    width = 2, size = 0.8,
    inherit.aes = FALSE) +
  
  # Points for per-bird means
  geom_point(
    data = summary_df_init,
    aes(y = mean_maxF_centered, x = Age, color = Bird, alpha = 0.9),
    size = 3, shape = 18,
    inherit.aes = FALSE
  ) +
  # Lines for per-bird trends
   geom_line(
    data = summary_df_init,
    aes(x = Age, y = mean_maxF_centered, color = Bird, group = Bird, alpha = 0.8),
    size = 0.7,
    inherit.aes = FALSE) +
  
  # Ribbon for overall trend SD
  geom_ribbon(
    data = trend_df_init,
    aes(x = Age, ymin = mean_maxF_centered - sd_maxF_centered, ymax = mean_maxF_centered + sd_maxF_centered),
    fill = "grey70", alpha = 0.5,
    inherit.aes = FALSE
  ) +
  
  # Line for overall trend mean
  geom_line(
    data = trend_df_init,
    aes(x = Age, y = mean_maxF_centered),
    color = "grey50", alpha = 0.8, size = 1.2,
    inherit.aes = FALSE) +
  
   scale_x_continuous(breaks = unique(force_data$Age)) +
  labs(x = "Age", y = "max F centred (N)", title = "") +
  #ylim(10, 25)+
  guides(alpha = "none")+
  theme_minimal() + 
  theme(legend.position = "none")

plot_age_maxFtot_centred




```

#### b. bodyweight relative max Ftot
```{r}

summary_df <- force_data %>%
  group_by(Bird, Age) %>%
  summarise(
    mean_maxF_bw = mean(max_Ftot_bw),
    sd_maxF_bw   = sd(max_Ftot_bw),
    .groups = "drop"
  )

trend_df <- force_data %>%
  group_by(Age) %>%
  summarise(
    mean_maxF_bw = mean(max_Ftot_bw, na.rm = TRUE),
    sd_maxF_bw   = sd(max_Ftot_bw, na.rm = TRUE),
    .groups = "drop"
  )


plot_age_maxF_bw <- ggplot(force_data, aes(x = Age, y = max_Ftot_bw, color = Bird)) +
  geom_point(position = position_jitter(width = 1), alpha = 0.3) +   # dots = trials
  geom_errorbar(
    data = summary_df,
    aes(x=Age, ymin = mean_maxF_bw - sd_maxF_bw, ymax = mean_maxF_bw + sd_maxF_bw, color = Bird, alpha = 0.6),
    width = 2, size = 0.8,
    inherit.aes = FALSE) +
  
  geom_point(
    data = summary_df,
    aes(y = mean_maxF_bw, x = Age, color = Bird, alpha = 0.9),
    size = 3, shape = 18,
    inherit.aes = FALSE
  ) +
  
   geom_line(
    data = summary_df,
    aes(x = Age, y = mean_maxF_bw, color = Bird, group = Bird, alpha = 0.8),
    size = 0.7,
    inherit.aes = FALSE) +
  
  geom_ribbon(
    data = trend_df,
    aes(x = Age, ymin = mean_maxF_bw - sd_maxF_bw, ymax = mean_maxF_bw + sd_maxF_bw),
    fill = "grey70", alpha = 0.5,
    inherit.aes = FALSE
  ) +
  
  geom_line(
    data = trend_df,
    aes(x = Age, y = mean_maxF_bw),
    color = "grey50", alpha = 0.8, size = 1.2,
    inherit.aes = FALSE) +
  
   scale_x_continuous(breaks = unique(force_data$Age)) +
  labs(x = "Age", y = "max F / bw", title = "") +
  #ylim(10, 25)+
  guides(alpha = "none")+
  theme_minimal()+
  theme(legend.position = "none")

plot_age_maxF_bw
```
#### c. vertical force
##### absolute
```{r}

summary_df <- force_data %>%
  group_by(Bird, Age) %>%
  summarise(
    mean_Fx = mean(-Fx_maxFtot),
    sd_Fx = sd(-Fx_maxFtot),
    .groups = "drop"
  )

trend_df <- force_data %>%
  group_by(Age) %>%
  summarise(
    mean_Fx = mean(-Fx_maxFtot, na.rm = TRUE),
    sd_Fx   = sd(-Fx_maxFtot, na.rm = TRUE),
    .groups = "drop"
  )


plot_age_Fx <- ggplot(force_data, aes(x = Age, y = -Fx_maxFtot, color = Bird)) +
  geom_point(position = position_jitter(width = 1), alpha = 0.3) +   # dots = trials
  geom_errorbar(
    data = summary_df,
    aes(x=Age, ymin = mean_Fx - sd_Fx, ymax = mean_Fx + sd_Fx, color = Bird, alpha = 0.6),
    width = 2, size = 0.8,
    inherit.aes = FALSE) +
  
  geom_point(
    data = summary_df,
    aes(y = mean_Fx, x = Age, color = Bird, alpha = 0.9),
    size = 2, shape = 18,
    inherit.aes = FALSE
  ) +
  
   geom_line(
    data = summary_df,
    aes(x = Age, y = mean_Fx, color = Bird, group = Bird, alpha = 0.8),
    size = 0.7,
    inherit.aes = FALSE) +
  
  geom_ribbon(
    data = trend_df,
    aes(x = Age, ymin = mean_Fx - sd_Fx, ymax = mean_Fx + sd_Fx),
    fill = "grey70", alpha = 0.5,
    inherit.aes = FALSE
  ) +
  
  geom_line(
    data = trend_df,
    aes(x = Age, y = mean_Fx),
    color = "grey50", alpha = 0.8, size = 1.2,
    inherit.aes = FALSE) +
  
   scale_x_continuous(breaks = unique(force_data$Age)) +
  labs(x = "Age", y = "Fx", title = "") +
  #ylim(10, 25)+
  guides(alpha = "none")+
  theme_minimal() +
  theme(legend.position = "none")

plot_age_Fx


```
##### relative
```{r}

summary_df <- force_data %>%
  group_by(Bird, Age) %>%
  summarise(
    mean_Fx_bw = mean(Fx_maxFtot_bw),
    sd_Fx_bw = sd(Fx_maxFtot_bw),
    .groups = "drop"
  )

trend_df <- force_data %>%
  group_by(Age) %>%
  summarise(
    mean_Fx_bw = mean(Fx_maxFtot_bw, na.rm = TRUE),
    sd_Fx_bw   = sd(Fx_maxFtot_bw, na.rm = TRUE),
    .groups = "drop"
  )


plot_age_Fx_bw <- ggplot(force_data, aes(x = Age, y = Fx_maxFtot_bw, color = Bird)) +
  geom_point(position = position_jitter(width = 1), alpha = 0.3) +   # dots = trials
  geom_errorbar(
    data = summary_df,
    aes(x=Age, ymin = mean_Fx_bw - sd_Fx_bw, ymax = mean_Fx_bw + sd_Fx_bw, color = Bird, alpha = 0.6),
    width = 2, size = 0.8,
    inherit.aes = FALSE) +
  
  geom_point(
    data = summary_df,
    aes(y = mean_Fx_bw, x = Age, color = Bird, alpha = 0.9),
    size = 2, shape = 18,
    inherit.aes = FALSE
  ) +
  
   geom_line(
    data = summary_df,
    aes(x = Age, y = mean_Fx_bw, color = Bird, group = Bird, alpha = 0.8),
    size = 0.7,
    inherit.aes = FALSE) +
  
  geom_ribbon(
    data = trend_df,
    aes(x = Age, ymin = mean_Fx_bw - sd_Fx_bw, ymax = mean_Fx_bw + sd_Fx_bw),
    fill = "grey70", alpha = 0.5,
    inherit.aes = FALSE
  ) +
  
  geom_line(
    data = trend_df,
    aes(x = Age, y = mean_Fx_bw),
    color = "grey50", alpha = 0.8, size = 1.2,
    inherit.aes = FALSE) +
  
   scale_x_continuous(breaks = unique(force_data$Age)) +
  labs(x = "Age", y = "Fx / bw (N)", title = "") +
  #ylim(10, 25)+
  guides(alpha = "none")+
  theme_minimal() +
  theme(legend.position = "none")

plot_age_Fx_bw

```
#### d. horizontal force
##### absolute
```{r}
summary_df <- force_data %>%
  group_by(Bird, Age) %>%
  summarise(
    mean_Fy = mean(Fy_maxFtot),
    sd_Fy = sd(Fy_maxFtot),
    .groups = "drop"
  )

trend_df <- force_data %>%
  group_by(Age) %>%
  summarise(
    mean_Fy = mean(Fy_maxFtot, na.rm = TRUE),
    sd_Fy   = sd(Fy_maxFtot, na.rm = TRUE),
    .groups = "drop"
  )


plot_age_Fy <- ggplot(force_data, aes(x = Age, y = Fy_maxFtot, color = Bird)) +
  geom_point(position = position_jitter(width = 1), alpha = 0.3) +   # dots = trials
  geom_errorbar(
    data = summary_df,
    aes(x=Age, ymin = mean_Fy - sd_Fy, ymax = mean_Fy + sd_Fy, color = Bird, alpha = 0.6),
    width = 2, size = 0.8,
    inherit.aes = FALSE) +
  
  geom_point(
    data = summary_df,
    aes(y = mean_Fy, x = Age, color = Bird, alpha = 0.9),
    size = 2, shape = 18,
    inherit.aes = FALSE
  ) +
  
   geom_line(
    data = summary_df,
    aes(x = Age, y = mean_Fy, color = Bird, group = Bird, alpha = 0.8),
    size = 0.7,
    inherit.aes = FALSE) +
  
  geom_ribbon(
    data = trend_df,
    aes(x = Age, ymin = mean_Fy - sd_Fy, ymax = mean_Fy + sd_Fy),
    fill = "grey70", alpha = 0.5,
    inherit.aes = FALSE
  ) +
  
  geom_line(
    data = trend_df,
    aes(x = Age, y = mean_Fy),
    color = "grey50", alpha = 0.8,size = 1.2,
    inherit.aes = FALSE) +
  
   scale_x_continuous(breaks = unique(force_data$Age)) +
  labs(x = "Age", y = "Fy", title = "") +
  #ylim(10, 25)+
  guides(alpha = "none")+
  theme_minimal() +
  theme(legend.position = "none")

plot_age_Fy

```

##### relative
```{r}
summary_df <- force_data %>%
  group_by(Bird, Age) %>%
  summarise(
    mean_Fy_bw = mean(Fy_maxFtot_bw),
    sd_Fy_bw = sd(Fy_maxFtot_bw),
    .groups = "drop"
  )

trend_df <- force_data %>%
  group_by(Age) %>%
  summarise(
    mean_Fy_bw = mean(Fy_maxFtot_bw, na.rm = TRUE),
    sd_Fy_bw   = sd(Fy_maxFtot_bw, na.rm = TRUE),
    .groups = "drop"
  )


plot_age_Fy_bw <- ggplot(force_data, aes(x = Age, y = Fy_maxFtot_bw, color = Bird)) +
  geom_point(position = position_jitter(width = 1), alpha = 0.3) +   # dots = trials
  geom_errorbar(
    data = summary_df,
    aes(x=Age, ymin = mean_Fy_bw - sd_Fy_bw, ymax = mean_Fy_bw + sd_Fy_bw, color = Bird, alpha = 0.6),
    width = 2, size = 0.8,
    inherit.aes = FALSE) +
  
  geom_point(
    data = summary_df,
    aes(y = mean_Fy_bw, x = Age, color = Bird, alpha = 0.9),
    size = 2, shape = 18,
    inherit.aes = FALSE
  ) +
  
   geom_line(
    data = summary_df,
    aes(x = Age, y = mean_Fy_bw, color = Bird, group = Bird, alpha = 0.8),
    size = 0.7,
    inherit.aes = FALSE) +
  
  geom_ribbon(
    data = trend_df,
    aes(x = Age, ymin = mean_Fy_bw - sd_Fy_bw, ymax = mean_Fy_bw + sd_Fy_bw),
    fill = "grey70", alpha = 0.5,
    inherit.aes = FALSE
  ) +
  
  geom_line(
    data = trend_df,
    aes(x = Age, y = mean_Fy_bw),
    color = "grey50", alpha = 0.8, size = 1.2,
    inherit.aes = FALSE) +
  
   scale_x_continuous(breaks = unique(force_data$Age)) +
  labs(x = "Age", y = "Fy / bw (N)", title = "") +
  #ylim(10, 25)+
  guides(alpha = "none")+
  ylim(0,6) +
  theme_minimal() +
  theme(legend.position = "none")

plot_age_Fy_bw


```
#### e. longitudinal force
##### absolute
```{r}

plot_age_performance(Fz_maxFtot, force_data, "max longitudinal force (N)")
```
##### relative
```{r}
plot_age_performance(Fz_maxFtot_bw, force_data, "max longitudinal force (N)")

```


#### f. impulse
##### impulse 
```{r}

plot_age_impulse <- plot_age_performance(impulse*1000, force_data, "impulse (mN*s)") +ylim(-10, 60) #1 row removed

plot_age_impulse
```

##### duration
```{r}
plot_age_duration <- plot_age_performance(duration, force_data, "impulse duration (s)") + ylim(0, 0.20)

plot_age_duration
```

##### impulse ratio
```{r}
plot_age_impulse_ratio <- plot_age_performance(impulse_ratio, force_data, "impulse ratio ( left impulse / right impulse )")

plot_age_impulse_ratio
```


##### left impulse
```{r}
plot_left_impulse <- plot_age_performance(left_impulse*1000, force_data, "left impulse (mN*s)") #1 outlier removed
plot_left_impulse
```

##### right impulse
```{r}
plot_right_impulse <- plot_age_performance(right_impulse*1000, force_data, "right impulse (mN*s)") 
plot_right_impulse
```

##### left duration
```{r}
plot_left_duration <- plot_age_performance(left_duration, force_data, "left duration (s)")
plot_left_duration
```

##### right duration
```{r}

plot_right_duration <- plot_age_performance(right_duration, force_data, "right duration (s)") + ylim(0, 0.075)

plot_right_duration

```


#### f. takeoff angle 
```{r}

plot_age_angle <- plot_age_performance(angle, force_data, "takeoff angle (degrees)") 
plot_age_angle
```


### PLOTS. altogether
```{r}
ggarrange(plot_age_maxFtot, plot_age_Fx, plot_age_Fy, plot_age_maxF_bw, plot_age_Fx_bw, plot_age_Fy_bw, nrow = 2, ncol=3)
```

### FLIGHT SPEED

#### a. turn rate
##### mean
```{r}
plot_tr_mean <- plot_age_performance(mean_turn_rate, flights, "mean turn rate (degrees/s)") +ylim(0,750)
plot_tr_mean

```

##### max
```{r}

plot_tr_max <- plot_age_performance(max_turn_rate, flights, "max turn rate (degrees/s)") +ylim(-1250, 5000)
plot_tr_max
```


#### b. centripetal acceleration
##### mean
```{r}

plot_ctracc_mean <- plot_age_performance(mean_centri_acc, flights, "mean centripetal acceleration (m/s^2)") +ylim(0, 30)
plot_ctracc_mean
```

##### max
```{r}

plot_ctracc_max <- plot_age_performance(max_centri_acc, flights, "max centripetal acceleration (m/s^2)") +ylim(-50, 150)
plot_ctracc_max

```
#### c. resultant velocity
##### mean
```{r}
plot_vres_mean <- plot_age_performance(mean_v_res, flights, "mean resultant velocity (m/s)")
plot_vres_mean
```
##### max
```{r}
plot_vres_max <- plot_age_performance(variable = max_v_res, df = flights, title = "max resultant velocity (m/s)")
plot_vres_max

```

#### d. resultant acceleration
##### mean
```{r}

plot_accres_mean <- plot_age_performance(mean_acc_res, flights, "mean resultant acceleration (m/s^2)")
plot_accres_mean
```
##### max
```{r}
plot_accres_max <- plot_age_performance(max_acc_res, flights, "max resultant acceleration (m/s^2)") +ylim(-100, 150)
plot_accres_max

```
#### e. X vel & acc
##### mean v
```{r}
plot_vx_mean <- plot_age_performance(mean_vx, flights, "mean X velocity (m/s)")
plot_vx_mean
```
##### max v
```{r}
plot_vx_max <- plot_age_performance(max_vx, flights, "max X velocity (m/s)")
plot_vx_max
```
##### mean acc
```{r}
plot_accx_mean <- plot_age_performance(mean_acc_x, flights, "mean X acceleration (m/s^2)") 
plot_accx_mean

```
##### max acc
```{r}
plot_accx_max <- plot_age_performance(max_acc_x, flights, "max X acceleration (m/s^2)") 
plot_accx_max

```
#### f. Y vel & acc
##### mean v
```{r}
plot_vy_mean <- plot_age_performance(mean_vy, flights, "mean Y velocity (m/s)") 
plot_vy_mean
```
##### max v
```{r}
plot_vy_max <- plot_age_performance(max_vy, flights, "max Y velocity (m/s)")
plot_vy_max +ylim(0, 6)
```
##### mean acc
```{r}
plot_accy_mean <- plot_age_performance(mean_acc_y, flights, "mean Y acceleration (m/s^2)") 
plot_accy_mean
```
##### max acc
```{r}
plot_accy_max <- plot_age_performance(mean_acc_y, flights, "max Y acceleration (m/s^2)") 
plot_accy_max
```
#### g. Z vel & acc
##### mean v
```{r}
plot_vz_mean <- plot_age_performance(mean_vz, flights, "mean Z velocity (m/s)")
plot_vz_mean
```
##### max v
```{r}
plot_vz_max <- plot_age_performance(max_vz, flights, "max Z velocity (m/s)")
plot_vz_max

```
##### mean acc
```{r}
plot_accz_mean <- plot_age_performance(mean_acc_z, flights, "mean Z acceleration (m/s^2)")
plot_accz_mean

```
##### max acc
```{r}
plot_accz_max <- plot_age_performance(max_acc_z, flights, "max Z acceleration (m/s^2)") 
plot_accz_max
```
### WINGBEATS
#### frequency
```{r}
plot_wb_mean <- plot_age_performance(wingbeat_frequency, flights, "wingbeat frequency (Hz)")
plot_wb_mean
```
#### upstroke
```{r}
plot_upstroke_acc <- plot_age_performance(upstroke_amplitude_acc, flights, "upstroke_head_acc (m/s^2)")
plot_upstroke_acc

plot_upstroke_m <- plot_age_performance(upstroke_amplitude_m, flights, "upstroke head displacement (m)")
plot_upstroke_m
```
#### downstroke
```{r}
plot_downstroke_acc <- plot_age_performance(downstroke_amplitude_acc, flights, "downstroke_head_acc (m/s^2)")
plot_downstroke_acc

plot_downstroke_m <- plot_age_performance(downstroke_amplitude_m, flights, "downstroke head displacement (m)")
plot_downstroke_m
```

### ggarrange Plots
```{r}
ggarrange(plot_vx_mean, plot_vy_mean, plot_vz_mean, 
          plot_accx_mean, plot_accy_mean, plot_accz_mean, nrow = 2, ncol = 3)

ggarrange(plot_vx_max, plot_vy_max, plot_vz_max,
          plot_accx_max, plot_accy_max, plot_accz_max, nrow = 2, ncol = 3)

ggarrange(plot_vres_mean, plot_accres_mean, plot_tr_mean, plot_ctracc_mean, nrow = 2, ncol =2)

ggarrange(plot_vres_max, plot_accres_max, plot_tr_max, plot_ctracc_max, nrow = 2, ncol =2 )

ggarrange(plot_wb_mean, plot_downstroke_m, plot_upstroke_m, nrow = 1, ncol = 3)
```

# II. Finding the knot - Broken Stick Model
## A. Significant
### MORPHOLOGY
#### bodyweight - 28.1
```{r}
m_lin <- lm(scale_weight ~ Age * Sex +Brood, data = wing_data_final)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg) # Male 0.8g lower and Brood 0.7g higher. 
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint <- m_seg$psi["psi1.Age", "Est."] #breakpoint: 28.46. 
slope_early <- slope(m_seg)$Age[1, "Est."]
slope_late  <- slope(m_seg)$Age[2, "Est."]

print(breakpoint)
print(slope_early)
print(slope_late)

```
#### wingspan - 28
```{r}
m_lin <- lm(wingspan ~ Age * Sex +Brood, data = wing_data_final)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 28))  # initial guess at 36
summary(m_seg) # Male 0.8g lower and Brood 0.7g higher. 
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint <- m_seg$psi["psi1.Age", "Est."] #breakpoint: 28.46. 
slope_early <- slope(m_seg)$Age[1, "Est."]
slope_late  <- slope(m_seg)$Age[2, "Est."]

print(breakpoint)
print(slope_early)
print(slope_late)

```
#### wing area - 28.5
```{r}
m_lin <- lm(wingarea ~ Age*Sex + Brood, data = wing_data_final)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 28))  # initial guess at 36
summary(m_seg) # Age not significant. Larger brood, larger wing area
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint <- m_seg$psi["psi1.Age", "Est."] #27.8
slope_early <- slope(m_seg)$Age[1, "Est."]
slope_late  <- slope(m_seg)$Age[2, "Est."]

print(breakpoint)
print(slope_early)
print(slope_late)
```


### LEG
#### max Fz-> x(paper)/bw -40
```{r}

m_lin <- lm(Fz_maxFtot_bw ~ Age*Sex +Brood, data = force_data)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 40))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.46. Age significant, male grows quicker. 
slope_early <- slope(m_seg)$Age[1, "Est."]
slope_late  <- slope(m_seg)$Age[2, "Est."]

print(breakpoint)
print(slope_early)
print(slope_late)

```
#### max Fy / bw - 34.5
```{r}


m_lin <- lm(Fy_maxFtot_bw ~ Age*Sex, data = force_data)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.46. Age significant, male grows quicker. 
slope_early <- slope(m_seg)$age[1, "Est."]
slope_late  <- slope(m_seg)$age[2, "Est."]




```
#### max Ftot /bw - 36.5
```{r}
m_lin <- lm(max_Ftot_bw ~ Age*Sex, data = force_data)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint <- m_seg$psi["psi1.Age", "Est."] #36.5. # male max resultant force generally stronger than females.
slope_early <- slope(m_seg)$age[1, "Est."]
slope_late  <- slope(m_seg)$age[2, "Est."]

```

#### take-off angle - 34.01
```{r}
m_lin <- lm(angle ~ Age*Sex + Brood, data = force_data)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint <- m_seg$psi["psi1.Age", "Est."] #36.5. # male max resultant force generally stronger than females.
slope_early <- slope(m_seg)$Age[1, "Est."]
slope_late  <- slope(m_seg)$Age[2, "Est."]

breakpoint
slope_early
slope_late

```
#### impulse -30.2
```{r}
m_lin <- lm(impulse*1000 ~ Age*Sex + Brood, data = force_data)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint <- m_seg$psi["psi1.Age", "Est."] #36.5. # male max resultant force generally stronger than females.
slope_early <- slope(m_seg)$Age[1, "Est."]
slope_late  <- slope(m_seg)$Age[2, "Est."]

breakpoint
slope_early
slope_late
```
#### left impulse -32
```{r}
m_lin <- lm(left_impulse*1000 ~ Age*Sex + Brood, data = force_data)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint <- m_seg$psi["psi1.Age", "Est."] #36.5. # male max resultant force generally stronger than females.
slope_early <- slope(m_seg)$Age[1, "Est."]
slope_late  <- slope(m_seg)$Age[2, "Est."]

breakpoint
slope_early
slope_late
```
#### right impulse - 24.9
```{r}
m_lin <- lm(right_impulse*1000 ~ Age*Sex + Brood, data = force_data)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint <- m_seg$psi["psi1.Age", "Est."] #36.5. # male max resultant force generally stronger than females.
slope_early <- slope(m_seg)$Age[1, "Est."]
slope_late  <- slope(m_seg)$Age[2, "Est."]

breakpoint
slope_early
slope_late
```
#### duration - 38.4
```{r}
m_lin <- lm(duration ~ Age*Sex + Brood, data = force_data)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint <- m_seg$psi["psi1.Age", "Est."] #36.5. # male max resultant force generally stronger than females.
slope_early <- slope(m_seg)$Age[1, "Est."]
slope_late  <- slope(m_seg)$Age[2, "Est."]

breakpoint
slope_early
slope_late
```

#### left duration - 40
```{r}
m_lin <- lm(left_duration ~ Age*Sex + Brood, data = force_data)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint <- m_seg$psi["psi1.Age", "Est."] #36.5. # male max resultant force generally stronger than females.
slope_early <- slope(m_seg)$Age[1, "Est."]
slope_late  <- slope(m_seg)$Age[2, "Est."]

breakpoint
slope_early
slope_late
```

#### right duration -33.1
```{r}
m_lin <- lm(right_duration ~ Age*Sex + Brood, data = force_data)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint <- m_seg$psi["psi1.Age", "Est."] #36.5. # male max resultant force generally stronger than females.
slope_early <- slope(m_seg)$Age[1, "Est."]
slope_late  <- slope(m_seg)$Age[2, "Est."]

breakpoint
slope_early
slope_late
```

### TRAJECTORY
#### Z velocity - 32.9
```{r}

m_lin <- lm(mean_vz ~ Age*Sex + Brood, data = flights)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
slope_early <- slope(m_seg)$Age[1, "Est."]
slope_late  <- slope(m_seg)$Age[2, "Est."]

breakpoint #32.9
slope_early #-0.04
slope_late #0.0013 

```
#### Y velocity - 33.0
```{r}

m_lin <- lm(mean_vy ~ Age*Sex + Brood, data = flights)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
slope_early <- slope(m_seg)$Age[1, "Est."]
slope_late  <- slope(m_seg)$Age[2, "Est."]

breakpoint
slope_early
slope_late

```

#### Y acceleration -33.7
```{r}
m_lin <- lm(mean_acc_y ~ Age*Sex + Brood, data = flights)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
slope_early <- slope(m_seg)$Age[1, "Est."]
slope_late  <- slope(m_seg)$Age[2, "Est."]

breakpoint # 33.7
slope_early # 0.09 m/s^2 per day
slope_late # 0.0088 m/s^2 per day 

```
#### Z acceleration -33.7
```{r}
m_lin <- lm(mean_acc_z ~ Age*Sex + Brood, data = flights)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
slope_early <- slope(m_seg)$Age[1, "Est."]
slope_late  <- slope(m_seg)$Age[2, "Est."]

breakpoint # 48.3
slope_early # 0.09 m/s^2 per day until 48.3 
slope_late # 0.009 m/s^2 per day 
```
#### tangential acceleration - 33.8
```{r}

m_lin <- lm(mean_acc_res ~ Age*Sex + Brood, data = flights)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
slope_early <- slope(m_seg)$Age[1, "Est."]
slope_late  <- slope(m_seg)$Age[2, "Est."]

breakpoint # 33.8 
slope_early # +0.09 m/s^2 per day
slope_late # +0.0076 m/s^2 per day 


```

#### tangential velocity - 34.3
```{r}

m_lin <- lm(mean_v_res ~ Age*Sex + Brood, data = flights)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
slope_early <- slope(m_seg)$Age[1, "Est."]
slope_late  <- slope(m_seg)$Age[2, "Est."]

breakpoint
slope_early
slope_late


```

#### turn rate - 41.7
```{r}
m_lin <- lm(mean_turn_rate ~ Age*Sex + Brood, data = flights)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
slope_early <- slope(m_seg)$Age[1, "Est."]
slope_late  <- slope(m_seg)$Age[2, "Est."]

breakpoint #41.7
slope_early #-2.57 (birds turn less/day until 40 days)
slope_late  #1.29 (birds turn more/day after 40 days)

```

#### X velocity - 60
```{r}

m_lin <- lm(mean_vx ~ Age*Sex + Brood, data = flights)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 60))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
slope_early <- slope(m_seg)$Age[1, "Est."]
slope_late  <- slope(m_seg)$Age[2, "Est."]

breakpoint #60
slope_early # -0.015
slope_late # 0.001

```
#### X acceleration -74.6
```{r}
m_lin <- lm(mean_acc_x ~ Age*Sex + Brood, data = flights)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 80))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
slope_early <- slope(m_seg)$Age[1, "Est."]
slope_late  <- slope(m_seg)$Age[2, "Est."]

breakpoint # 74.6 
slope_early # -0.033 m/s^2 per day
slope_late # +0.029 m/s^2 per day 
```













## B. Not significant

### wing loading
```{r}

m_lin <- lm(wing_loading ~ Age*Sex + Brood, data = wing_data_final)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 28))  # initial guess at 36
summary(m_seg) # Age not significant. Larger brood, larger wing area
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint <- m_seg$psi["psi1.Age", "Est."] #27.8
slope_early <- slope(m_seg)$Age[1, "Est."]
slope_late  <- slope(m_seg)$Age[2, "Est."]

print(breakpoint)
print(slope_early)
print(slope_late)



```

### max Fx / bw
- not significant
```{r}
m_lin <- lm(Fx_maxFtot_bw ~ Age*Sex + Brood, data = force_data)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint <- m_seg$psi["psi1.Age", "Est."] # no breakpoint but higher Fx force in males
slope_early <- slope(m_seg)$age[1, "Est."]
slope_late  <- slope(m_seg)$age[2, "Est."]
```

### impulse ratio - not significant!!
```{r}
m_lin <- lm(impulse_ratio ~ Age*Sex + Brood, data = force_data)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint <- m_seg$psi["psi1.Age", "Est."] #36.5. # male max resultant force generally stronger than females.
slope_early <- slope(m_seg)$Age[1, "Est."]
slope_late  <- slope(m_seg)$Age[2, "Est."]

breakpoint
slope_early
slope_late
```

### mean centripetal acceleration
```{r}

m_lin <- lm(mean_centri_acc ~ Age*Sex + Brood, data = flights)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
slope_early <- slope(m_seg)$Age[1, "Est."]
slope_late  <- slope(m_seg)$Age[2, "Est."]

breakpoint 
slope_early 
slope_late 

```
### max centripetal acceleration
```{r}

m_lin <- lm(max_centri_acc ~ Age*Sex + Brood, data = flights)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
slope_early <- slope(m_seg)$Age[1, "Est."]
slope_late  <- slope(m_seg)$Age[2, "Est."]

breakpoint 
slope_early 
slope_late 

```
### upstroke 
```{r}

m_lin <- lm(upstroke_amplitude_m*1000 ~ Age*Sex + Brood, data = flights)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 32))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
slope_early <- slope(m_seg)$Age[1, "Est."]
slope_late  <- slope(m_seg)$Age[2, "Est."]

breakpoint #31
slope_early #0.18 mm (more head displacement/day until 80 days)
slope_late  #0.01 mm(less "" /day after 80 days)

```
### downstroke
```{r}

m_lin <- lm(downstroke_amplitude_m*1000 ~ Age*Sex + Brood, data = flights)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 32))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
slope_early <- slope(m_seg)$Age[1, "Est."]
slope_late  <- slope(m_seg)$Age[2, "Est."]

breakpoint #31
slope_early #0.18 mm (more head displacement/day until 80 days)
slope_late  #0.01 mm(less "" /day after 80 days)

```
### wingbeat frequency - 60 
```{r}
m_lin <- lm(wingbeat_frequency ~ Age*Sex + Brood, data = flights)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 60))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
slope_early <- slope(m_seg)$Age[1, "Est."]
slope_late  <- slope(m_seg)$Age[2, "Est."]

breakpoint #60
slope_early #0.0031966 (less wingbeat Hz/day until 42.9 days)
slope_late  #0.020644 (more wingbeat Hz/day after 42.9 days)

```

# III. Model Analysis
## example
```{r}
plot_age_performance <- function(variable, df, title) {
  
  # Summary by individual
  summary_df <- df %>%
    group_by(Bird, Age) %>%
    summarise(
      mean = mean({{variable}}, na.rm = TRUE),
      sd = sd({{variable}}, na.rm = TRUE),
      n = n(),
      .groups = "drop"
    )
  
  # Summary by age
  trend_df <- df %>%
    group_by(Age) %>%
    summarise(
      mean = mean({{variable}}, na.rm = TRUE),
      sd = sd({{variable}}, na.rm = TRUE),
      n = n(),
      .groups = "drop"
    )
  
  # Plot
  plot_age_performance<- ggplot(df, aes(x = Age, y = {{variable}}, color = Bird)) +
    geom_point(position = position_jitter(width = 1), alpha = 0.3) +
    geom_errorbar(
      data = summary_df,
      aes(x = Age, ymin = mean - sd, ymax = mean + sd, color = Bird, alpha = 0.6),
      width = 2, size = 0.8,
      inherit.aes = FALSE
    ) +
    geom_point(
      data = summary_df,
      aes(y = mean, x = Age, color = Bird, alpha = 0.9),
      size = 2, shape = 18,
      inherit.aes = FALSE
    ) +
    geom_line(
      data = summary_df,
      aes(x = Age, y = mean, color = Bird, group = Bird, alpha = 0.8),
      size = 0.7,
      inherit.aes = FALSE
    ) +
    geom_ribbon(
      data = trend_df,
      aes(x = Age, ymin = mean - sd, ymax = mean + sd),
      fill = "grey70", alpha = 0.5,
      inherit.aes = FALSE
    ) +
    geom_line(
      data = trend_df,
      aes(x = Age, y = mean),
      ,color = "grey50", alpha = 0.8, size = 1.2,
      inherit.aes = FALSE
    ) +
    scale_x_continuous(breaks = unique(df$Age)) +
    labs(x = "Age", y = title, title = "") +
    guides(alpha = "none", color = "none") +
    theme(legend.position = "none")+
    theme_minimal()
  
  return(plot_age_performance)
}

```
## function plot_lmer_stats
```{r}

plot_lmer_stats <- function(var, df, breakpoint) {
# Use precise breakpoint in hinge term
  
  options(contrasts = c("contr.sum", "contr.poly"))
  
  df$hinge <- pmax(df$Age - breakpoint, 0)
  
  lmer_{var} <- lmer(var ~ Age*Sex + hinge + (1|Brood/Bird), data = force_data)
  
  # computing residuals
  raw_cond_{var} <- compute_redres(lmer_{var})
  
  hist(raw_cond_{var}) #normality in histogram
  plot_redres(lmer_{var}) #heteroskedasticity
  plot_resqq(lmer_{var}) #normality of residuals
  plot_ranef(lmer_{var}) 
  ## Oops not so well behaving
  summary(lmer_{var})
  Anova(lmer_{var}, type =3, test.statistic = "Chisq") # Does age affect bodyweight overall?
  
  
  linearHypothesis(lmer_{var}, "Age = 0") # Is the slope before the breakpoint flat? -> no
  linearHypothesis(lmer_{var}, "Age+hinge = 0") # Is the slope after the breakpoint flat? -> no
  
  pred_{var} <- ggpredict(lmer_{var}, terms = c("Age [all]", "hinge"))
  
  # Create hinge variable
  
  df$hinge <- pmax(df$Age - breakpoint, 0)
  
  # Get fitted values across Age
  newdat <- data.frame(
    Age = seq(min(df$Age), max(df$Age), by = 1),
    Sex = "F",        # or average over Sex
    Brood = mean(df$Brood),
    hinge = pmax(seq(min(df$Age), max(df$Age), by = 1) - breakpoint, 0)
  )
  
  newdat$pred <- predict(lmer_{var}, newdat, re.form = NA)
  
  
  plot_{{var}}_final <-
    plot_{{var}}_age +
    # thick black line underneath
    geom_line(
      data = newdat,
      aes(x = Age, y = pred),
      color = "#1f77b4", size = 1.5, inherit.aes = FALSE, #linetype = "dashed"
    ) +
    #geom_line(
    #  data = newdat,
    #  aes(x = Age, y = pred),color = "red",
    #  size = 1.2, inherit.aes = FALSE
    #) +
    geom_vline(xintercept = breakpoint, linetype = "dashed", color = "red")
  
  return(plot_{{var}}_final)
  
}



```
## MORPHOLOGY
### bodyweight
#### lmer
-Age and Hinge significant
- Slope before and after breakpoint both significant
```{r}
# Use precise breakpoint in hinge term
force_data$hinge <- pmax(force_data$Age - 28.5, 0)

lmer_bw <- lmer(bodyweight_g ~ Age*Sex + Brood + hinge + (1|Bird), data = force_data)

# computing residuals
raw_cond_bw <- compute_redres(lmer_bw)

hist(raw_cond_bw) #normality in histogram
plot_redres(lmer_bw) #heteroskedasticity
plot_resqq(lmer_bw) #normality of residuals
plot_ranef(lmer_bw) 
## Oops not so well behaving
summary(lmer_bw)
Anova(lmer_bw, type =3, test.statistic = "Chisq") # Does age affect bodyweight overall?


linearHypothesis(lmer_bw, "Age = 0") # Is the slope before the breakpoint flat? -> no
linearHypothesis(lmer_bw, "Age+hinge = 0") # Is the slope after the breakpoint flat? -> no


```
#### function compare_models_aic()
```{r}
# Function to compare models using AIC, ΔAIC, and Akaike weights
compare_models_aic <- function(models) {
  # Validate input
  if (!is.list(models) || length(models) < 2) {
    stop("Please provide a list of at least two fitted model objects.")
  }
  
  # Extract model names
  model_names <- names(models)
  if (is.null(model_names) || any(model_names == "")) {
    model_names <- paste0("Model_", seq_along(models))
  }
  
  # Compute AIC for each model
  aic_values <- sapply(models, function(m) {
    tryCatch(AIC(m), error = function(e) NA)
  })
  
  # Check for NA values
  if (any(is.na(aic_values))) {
    stop("One or more models could not have their AIC computed.")
  }
  
  # Calculate ΔAIC
  delta_aic <- aic_values - min(aic_values)
  
  # Calculate Akaike weights
  rel_likelihood <- exp(-0.5 * delta_aic)
  akaike_weights <- rel_likelihood / sum(rel_likelihood)
  
  # Create results data frame
  results <- data.frame(
    Model = model_names,
    AIC = round(aic_values, 3),
    Delta_AIC = round(delta_aic, 3),
    Akaike_Weight = round(akaike_weights, 4),
    stringsAsFactors = FALSE
  )
  
  # Order by AIC
  results <- results[order(results$AIC), ]
  rownames(results) <- NULL
  
  return(results)
}
```
#### aic comparison
```{r}

wing_data_final$hinge <- pmax(wing_data_final$Age - 28.1, 0)
#model 1. Brood and Bird as random effect
m1 <- lm(scale_weight ~ Age*Sex + hinge + Brood + Bird, data = wing_data_final)
m2 <- lmer(scale_weight ~ Age*Sex + hinge + (1|Brood/Bird), data = wing_data_final)
# model 2. Brood as a fixed term 
m3 <- lmer(scale_weight ~ Age*Sex + Brood + hinge + (1|Bird), data = wing_data_final)

# Compare models
compare_models_aic(list(
  "linear model" = m1,
  "(1|Brood/Bird)" = m2,
  "(1|Bird)" = m3
))


```



#### Plot
```{r}

pred_bw <- ggpredict(lmer_bw, terms = c("Age [all]", "hinge"))

# Create hinge variable
breakpoint <- 28.5
force_data$hinge <- pmax(force_data$Age - breakpoint, 0)

# Get fitted values across Age
newdat <- data.frame(
  Age = seq(min(force_data$Age), max(force_data$Age), by = 1),
  Sex = "F",        # or average over Sex
  Brood = mean(force_data$Brood),
  hinge = pmax(seq(min(force_data$Age), max(force_data$Age), by = 1) - breakpoint, 0)
)

newdat$pred <- predict(lmer_bw, newdat, re.form = NA)


plot_bw_final <-
  plot_bw_age +
  # thick black line underneath
  geom_line(
    data = newdat,
    aes(x = Age, y = pred),
    color = "#1f77b4", size = 1.5, inherit.aes = FALSE, #linetype = "dashed"
  ) +
  #geom_line(
  #  data = newdat,
  #  aes(x = Age, y = pred),color = "red",
  #  size = 1.2, inherit.aes = FALSE
  #) +
  geom_vline(xintercept = 28.5, linetype = "dashed", color = "red")

plot_bw_final

```

### wingspan
### bodyweight
#### lmer
-Age and Hinge significant
- Slope before and after breakpoint both significant
```{r}
# Use precise breakpoint in hinge term
force_data$hinge <- pmax(force_data$Age - 28.3, 0)

lmer_bw <- lmer(wingspan_calc~ Age*Sex + Brood + hinge + (1|Bird), data = win)

# computing residuals
raw_cond_bw <- compute_redres(lmer_bw)

hist(raw_cond_bw) #normality in histogram
plot_redres(lmer_bw) #heteroskedasticity
plot_resqq(lmer_bw) #normality of residuals
plot_ranef(lmer_bw) 
## Oops not so well behaving
summary(lmer_bw)
Anova(lmer_bw, type =3, test.statistic = "Chisq") # Does age affect bodyweight overall?


linearHypothesis(lmer_bw, "Age = 0") # Is the slope before the breakpoint flat? -> no
linearHypothesis(lmer_bw, "Age+hinge = 0") # Is the slope after the breakpoint flat? -> no


```
## LEGS
### max Ftot / bw
-Age and Hinge signifcant 
-before and after breakpoint significant
#### lmer
```{r}
# Use precise breakpoint in hinge term
force_data$hinge <- pmax(force_data$Age - 36.5, 0)

lmer_maxF_bw <- lmer(max_Ftot_bw ~ Age*Sex + Brood + hinge + (1|Bird), data = force_data)

# computing residuals
raw_cond_maxF <- compute_redres(lmer_maxF_bw)

hist(raw_cond_maxF) #normality in histogram
plot_redres(lmer_maxF_bw) #heteroskedasticity
plot_resqq(lmer_maxF_bw) #normality of residuals
plot_ranef(lmer_maxF_bw) 
## Oops not so well behaving
summary(lmer_maxF_bw)
Anova(lmer_maxF_bw, type =3, test.statistic = "Chisq") # Does age affect bodyweight overall?


linearHypothesis(lmer_maxF_bw, "Age = 0") # Is the slope before the breakpoint flat? -> no
linearHypothesis(lmer_maxF_bw, "Age+hinge = 0") # Is the slope after the breakpoint flat? -> no


```
#### Plot
```{r}
pred_bw <- ggpredict(lmer_maxF_bw, terms = c("Age [all]", "hinge"))

# Create hinge variable
breakpoint <- 36.5
force_data$hinge <- pmax(force_data$Age - breakpoint, 0)

# Get fitted values across Age
newdat <- data.frame(
  Age = seq(min(force_data$Age), max(force_data$Age), by = 1),
  Sex = "F",        # or average over Sex
  Brood = mean(force_data$Brood),
  hinge = pmax(seq(min(force_data$Age), max(force_data$Age), by = 1) - breakpoint, 0)
)

newdat$pred <- predict(lmer_maxF_bw, newdat, re.form = NA)


plot_maxF_final <-
  plot_age_maxF_bw +
  geom_line(
    data = newdat,
    aes(x = Age, y = pred),
    color = "#1f77b4", size = 1.5, inherit.aes = FALSE, #linetype = "dashed"
  ) +
  #geom_line(
  #  data = newdat,
  #  aes(x = Age, y = pred),color = "red",
  #  size = 1.2, inherit.aes = FALSE
  #) +
  geom_vline(xintercept = 36.5, linetype = "dashed", color = "red")

plot_maxF_final

```

### max Fy / bw
- Age and hinge significant
- before and after breakpoint significant.
#### lmer
```{r}
# Use precise breakpoint in hinge term
force_data$hinge <- pmax(force_data$Age - 34.5, 0)

lmer_Fy_bw <- lmer(Fy_maxFtot_bw ~ Age*Sex + Brood + hinge + (1|Bird), data = force_data)

# computing residuals
raw_cond_Fy_bw <- compute_redres(lmer_Fy_bw)

hist(raw_cond_Fy_bw) #normality in histogram
plot_redres(lmer_Fy_bw) #heteroskedasticity
plot_resqq(lmer_Fy_bw) #normality of residuals
plot_ranef(lmer_Fy_bw) 
## Oops not so well behaving
summary(lmer_Fy_bw)
Anova(lmer_Fy_bw, type =3, test.statistic = "Chisq") # Does age affect bodyweight overall?


linearHypothesis(lmer_Fy_bw, "Age = 0") # Is the slope before the breakpoint flat? -> no
linearHypothesis(lmer_Fy_bw, "Age+hinge = 0") # Is the slope after the breakpoint flat? -> no


```
#### Plot
```{r}
pred_bw <- ggpredict(lmer_Fy_bw, terms = c("Age [all]", "hinge"))

# Create hinge variable
breakpoint <- 34.5
force_data$hinge <- pmax(force_data$Age - breakpoint, 0)

# Get fitted values across Age
newdat <- data.frame(
  Age = seq(min(force_data$Age), max(force_data$Age), by = 1),
  Sex = "F",        # or average over Sex
  Brood = mean(force_data$Brood),
  hinge = pmax(seq(min(force_data$Age), max(force_data$Age), by = 1) - breakpoint, 0)
)

newdat$pred <- predict(lmer_Fy_bw, newdat, re.form = NA)


plot_Fy_final <-
  plot_age_Fy_bw +
  geom_line(
    data = newdat,
    aes(x = Age, y = pred),
    color = "#1f77b4", size = 1.5, inherit.aes = FALSE, #linetype = "dashed"
  ) +
  #geom_line(
  #  data = newdat,
  #  aes(x = Age, y = pred),color = "red",
  #  size = 1.2, inherit.aes = FALSE
  #) +
  geom_vline(xintercept = 34.5, linetype = "dashed", color = "red")

plot_Fy_final

```
### takeoff angle
#### lmer
```{r}
# Use precise breakpoint in hinge term
force_data$hinge <- pmax(force_data$Age - 34.0, 0)

lmer_angle <- lmer(angle ~ Age*Sex + Brood + hinge + (1|Bird), data = force_data)

# computing residuals
raw_cond_angle <- compute_redres(lmer_angle)

hist(raw_cond_angle) #normality in histogram
plot_redres(lmer_angle) #heteroskedasticity
plot_resqq(lmer_angle) #normality of residuals
plot_ranef(lmer_angle) 
## Oops not so well behaving
summary(lmer_angle)
Anova(lmer_angle, type =3, test.statistic = "Chisq") # Does age affect bodyweight overall?


linearHypothesis(lmer_angle, "Age = 0") # Is the slope before the breakpoint flat? -> no
linearHypothesis(lmer_angle, "Age+hinge = 0") # Is the slope after the breakpoint flat? -> no


```
#### Plot
```{r}
pred_bw <- ggpredict(lmer_angle, terms = c("Age [all]", "hinge"))

# Create hinge variable
breakpoint <- 34.0
force_data$hinge <- pmax(force_data$Age - breakpoint, 0)

# Get fitted values across Age
newdat <- data.frame(
  Age = seq(min(force_data$Age), max(force_data$Age), by = 1),
  Sex = "F",        # or average over Sex
  Brood = mean(force_data$Brood),
  hinge = pmax(seq(min(force_data$Age), max(force_data$Age), by = 1) - breakpoint, 0)
)

newdat$pred <- predict(lmer_angle, newdat, re.form = NA)


plot_angle_final <-
  plot_age_angle +
  geom_line(
    data = newdat,
    aes(x = Age, y = pred),
    color = "#1f77b4", size = 1.5, inherit.aes = FALSE, #linetype = "dashed"
  ) +
  #geom_line(
  #  data = newdat,
  #  aes(x = Age, y = pred),color = "red",
  #  size = 1.2, inherit.aes = FALSE
  #) +
  geom_vline(xintercept = 34.0, linetype = "dashed", color = "red")

plot_angle_final

```
### impulse
#### lmer
- Age and hinge significant
- before and after breakpoint significant
```{r}
# Use precise breakpoint in hinge term
force_data$hinge <- pmax(force_data$Age - 30.2, 0)

lmer_impulse <- lmer(impulse*1000 ~ Age*Sex + Brood + hinge + (1|Brood/Bird), data = force_data)

# computing residuals
raw_cond_impulse <- compute_redres(lmer_impulse)

hist(raw_cond_impulse) #normality in histogram
plot_redres(lmer_impulse) #heteroskedasticity
plot_resqq(lmer_impulse) #normality of residuals
plot_ranef(lmer_impulse) 
## Oops not so well behaving
summary(lmer_impulse)
Anova(lmer_impulse, type =3, test.statistic = "Chisq") # Does age affect bodyweight overall?


linearHypothesis(lmer_impulse, "Age = 0") # Is the slope before the breakpoint flat? -> no
linearHypothesis(lmer_impulse, "Age+hinge = 0") # Is the slope after the breakpoint flat? -> no


```
#### Plot
```{r}
#pred_bw <- ggpredict(lmer_impulse, terms = c("Age [all]", "hinge"))

# Create hinge variable
breakpoint <- 30.2
force_data$hinge <- pmax(force_data$Age - breakpoint, 0)

# Get fitted values across Age
newdat <- data.frame(
  Age = seq(min(force_data$Age), max(force_data$Age), by = 1),
  Sex = "F",        # or average over Sex
  Brood = mean(force_data$Brood),
  hinge = pmax(seq(min(force_data$Age), max(force_data$Age), by = 1) - breakpoint, 0)
)

newdat$pred <- predict(lmer_impulse, newdat, re.form = NA)


plot_impulse_final <-
  plot_age_impulse +
  geom_line(
    data = newdat,
    aes(x = Age, y = pred),
    color = "#1f77b4", size = 1.5, inherit.aes = FALSE, #linetype = "dashed"
  ) +
  #geom_line(
  #  data = newdat,
  #  aes(x = Age, y = pred),color = "red",
  #  size = 1.2, inherit.aes = FALSE
  #) +
  geom_vline(xintercept = 30.2, linetype = "dashed", color = "red")

plot_impulse_final

```
### left impulse
#### lmer
- Age and hinge significant
- before and after breakpoint significant
```{r}
# Use precise breakpoint in hinge term
force_data$hinge <- pmax(force_data$Age - 32, 0)

lmer_left_impulse <- lmer(left_impulse*1000 ~ Age*Sex + Brood + hinge + (1|Brood/Bird), data = force_data)

# computing residuals
raw_cond_left_impulse <- compute_redres(lmer_left_impulse)

hist(raw_cond_left_impulse) #normality in histogram
plot_redres(lmer_left_impulse) #heteroskedasticity
plot_resqq(lmer_left_impulse) #normality of residuals
plot_ranef(lmer_left_impulse) 
## Oops not so well behaving
summary(lmer_left_impulse)
Anova(lmer_left_impulse, type =3, test.statistic = "Chisq") # Does age affect bodyweight overall?


linearHypothesis(lmer_left_impulse, "Age = 0") # Is the slope before the breakpoint flat? -> no
linearHypothesis(lmer_left_impulse, "Age+hinge = 0") # Is the slope after the breakpoint flat? -> no


```
#### Plot
```{r}
#pred_bw <- ggpredict(lmer_left_impulse, terms = c("Age [all]", "hinge"))

# Create hinge variable
breakpoint <- 32
force_data$hinge <- pmax(force_data$Age - breakpoint, 0)

# Get fitted values across Age
newdat <- data.frame(
  Age = seq(min(force_data$Age), max(force_data$Age), by = 1),
  Sex = "F",        # or average over Sex
  Brood = mean(force_data$Brood),
  hinge = pmax(seq(min(force_data$Age), max(force_data$Age), by = 1) - breakpoint, 0)
)

newdat$pred <- predict(lmer_left_impulse, newdat, re.form = NA)


plot_left_impulse_final <-
  plot_left_impulse +
  geom_line(
    data = newdat,
    aes(x = Age, y = pred),
    color = "#1f77b4", size = 1.5, inherit.aes = FALSE, #linetype = "dashed"
  ) +
  #geom_line(
  #  data = newdat,
  #  aes(x = Age, y = pred),color = "red",
  #  size = 1.2, inherit.aes = FALSE
  #) +
  geom_vline(xintercept = 32, linetype = "dashed", color = "red")

plot_left_impulse_final

```

### right impulse
#### lmer
- Age and hinge significant
- before and after breakpoint significant
```{r}
# Use precise breakpoint in hinge term
breakpoint <- 24.9
force_data$hinge <- pmax(force_data$Age - breakpoint, 0)

lmer_right_impulse <- lmer(right_impulse*1000 ~ Age*Sex + Brood + hinge + (1|Brood/Bird), data = force_data)

# computing residuals
raw_cond_right_impulse <- compute_redres(lmer_right_impulse)

hist(raw_cond_right_impulse) #normality in histogram
plot_redres(lmer_right_impulse) #heteroskedasticity
plot_resqq(lmer_right_impulse) #normality of residuals
plot_ranef(lmer_right_impulse) 
## Oops not so well behaving
summary(lmer_right_impulse)
Anova(lmer_right_impulse, type =3, test.statistic = "Chisq") # Does age affect bodyweight overall?


linearHypothesis(lmer_right_impulse, "Age = 0") # Is the slope before the breakpoint flat? -> no
linearHypothesis(lmer_right_impulse, "Age+hinge = 0") # Is the slope after the breakpoint flat? -> no

# Brood singular 
```
#### Plot
```{r}

# Create hinge variable
breakpoint <- 24.9
force_data$hinge <- pmax(force_data$Age - breakpoint, 0)

# Get fitted values across Age
newdat <- data.frame(
  Age = seq(min(force_data$Age), max(force_data$Age), by = 1),
  Sex = "F",        # or average over Sex
  Brood = mean(force_data$Brood),
  hinge = pmax(seq(min(force_data$Age), max(force_data$Age), by = 1) - breakpoint, 0)
)

newdat$pred <- predict(lmer_right_impulse, newdat, re.form = NA)


plot_right_impulse_final <-
  plot_right_impulse +
  geom_line(
    data = newdat,
    aes(x = Age, y = pred),
    color = "#1f77b4", size = 1.5, inherit.aes = FALSE, #linetype = "dashed"
  ) +
  #geom_line(
  #  data = newdat,
  #  aes(x = Age, y = pred),color = "red",
  #  size = 1.2, inherit.aes = FALSE
  #) +
  geom_vline(xintercept = breakpoint, linetype = "dashed", color = "red")

plot_right_impulse_final

```
### duration -38.4
#### lmer
- Age and hinge significant
- before and after breakpoint significant
```{r}
# Use precise breakpoint in hinge term
breakpoint <- 38.4
force_data$hinge <- pmax(force_data$Age - breakpoint, 0)

lmer_duration <- lmer(duration ~ Age*Sex + Brood + hinge + (1|Brood/Bird), data = force_data)

# computing residuals
raw_cond_duration <- compute_redres(lmer_duration)

hist(raw_cond_duration) #normality in histogram
plot_redres(lmer_duration) #heteroskedasticity
plot_resqq(lmer_duration) #normality of residuals
plot_ranef(lmer_duration) 
## Oops not so well behaving
summary(lmer_duration)
Anova(lmer_duration, type =3, test.statistic = "Chisq") # Does age affect bodyweight overall?


linearHypothesis(lmer_duration, "Age = 0") # Is the slope before the breakpoint flat? -> no
linearHypothesis(lmer_duration, "Age+hinge = 0") # Is the slope after the breakpoint flat? -> no

# Brood singular 
```
#### Plot
```{r}

# Create hinge variable
breakpoint <- 38.4
force_data$hinge <- pmax(force_data$Age - breakpoint, 0)

# Get fitted values across Age
newdat <- data.frame(
  Age = seq(min(force_data$Age), max(force_data$Age), by = 1),
  Sex = "F",        # or average over Sex
  Brood = mean(force_data$Brood),
  hinge = pmax(seq(min(force_data$Age), max(force_data$Age), by = 1) - breakpoint, 0)
)

newdat$pred <- predict(lmer_duration, newdat, re.form = NA)


plot_duration_final <-
  plot_age_duration +
  geom_line(
    data = newdat,
    aes(x = Age, y = pred),
    color = "#1f77b4", size = 1.5, inherit.aes = FALSE, #linetype = "dashed"
  ) +
  #geom_line(
  #  data = newdat,
  #  aes(x = Age, y = pred),color = "red",
  #  size = 1.2, inherit.aes = FALSE
  #) +
  geom_vline(xintercept = breakpoint, linetype = "dashed", color = "red")

plot_duration_final

```

### left duration
#### lmer
- Age and hinge significant
- before and after breakpoint significant
```{r}
# Use precise breakpoint in hinge term
breakpoint <- 40
force_data$hinge <- pmax(force_data$Age - breakpoint, 0)

lmer_left_duration <- lmer(left_duration ~ Age*Sex + Brood + hinge + (1|Brood/Bird), data = force_data)

# computing residuals
raw_cond_left_duration <- compute_redres(lmer_left_duration)

hist(raw_cond_left_duration) #normality in histogram
plot_redres(lmer_left_duration) #heteroskedasticity
plot_resqq(lmer_left_duration) #normality of residuals
plot_ranef(lmer_left_duration) 
## Oops not so well behaving
summary(lmer_left_duration)
Anova(lmer_left_duration, type =3, test.statistic = "Chisq") # Does age affect bodyweight overall?


linearHypothesis(lmer_left_duration, "Age = 0") # Is the slope before the breakpoint flat? -> no
linearHypothesis(lmer_left_duration, "Age+hinge = 0") # Is the slope after the breakpoint flat? -> no

# Brood singular 
```
#### Plot
```{r}

# Create hinge variable
breakpoint <- 40
force_data$hinge <- pmax(force_data$Age - breakpoint, 0)

# Get fitted values across Age
newdat <- data.frame(
  Age = seq(min(force_data$Age), max(force_data$Age), by = 1),
  Sex = "F",        # or average over Sex
  Brood = mean(force_data$Brood),
  hinge = pmax(seq(min(force_data$Age), max(force_data$Age), by = 1) - breakpoint, 0)
)

newdat$pred <- predict(lmer_left_duration, newdat, re.form = NA)


plot_left_duration_final <-
  plot_left_duration +
  geom_line(
    data = newdat,
    aes(x = Age, y = pred),
    color = "#1f77b4", size = 1.5, inherit.aes = FALSE, #linetype = "dashed"
  ) +
  #geom_line(
  #  data = newdat,
  #  aes(x = Age, y = pred),color = "red",
  #  size = 1.2, inherit.aes = FALSE
  #) +
  geom_vline(xintercept = breakpoint, linetype = "dashed", color = "red")

plot_left_duration_final

```
### right duration - 33.1
#### lmer
- Age and hinge significant
- after breakpoint not significant
```{r}
# Use precise breakpoint in hinge term
breakpoint <- 33.1
force_data$hinge <- pmax(force_data$Age - breakpoint, 0)

lmer_right_duration <- lmer(right_duration ~ Age*Sex + Brood + hinge + (1|Brood/Bird), data = force_data)

# computing residuals
raw_cond_right_duration <- compute_redres(lmer_right_duration)

hist(raw_cond_right_duration) #normality in histogram
plot_redres(lmer_right_duration) #heteroskedasticity
plot_resqq(lmer_right_duration) #normality of residuals
plot_ranef(lmer_right_duration) 
## Oops not so well behaving
summary(lmer_right_duration)
Anova(lmer_right_duration, type =3, test.statistic = "Chisq") # Does age affect bodyweight overall?


linearHypothesis(lmer_right_duration, "Age = 0") # Is the slope before the breakpoint flat? -> no
linearHypothesis(lmer_right_duration, "Age+hinge = 0") # Is the slope after the breakpoint flat? -> no

# Brood singular 
```
#### Plot
```{r}

# Create hinge variable
breakpoint <- 33.1
force_data$hinge <- pmax(force_data$Age - breakpoint, 0)

# Get fitted values across Age
newdat <- data.frame(
  Age = seq(min(force_data$Age), max(force_data$Age), by = 1),
  Sex = "F",        # or average over Sex
  Brood = mean(force_data$Brood),
  hinge = pmax(seq(min(force_data$Age), max(force_data$Age), by = 1) - breakpoint, 0)
)

newdat$pred <- predict(lmer_right_duration, newdat, re.form = NA)


plot_right_duration_final <-
  plot_right_duration +
  geom_line(
    data = newdat,
    aes(x = Age, y = pred),
    color = "#1f77b4", size = 1.5, inherit.aes = FALSE, #linetype = "dashed"
  ) +
  #geom_line(
  #  data = newdat,
  #  aes(x = Age, y = pred),color = "red",
  #  size = 1.2, inherit.aes = FALSE
  #) +
  geom_vline(xintercept = breakpoint, linetype = "dashed", color = "red")

plot_right_duration_final

```


### Final plots
```{r}

plots_body_legs_significants <-
  ggarrange(plot_bw_final, plot_maxF_final, 
            plot_Fy_final, plot_angle_final,
            ncol = 2, nrow=2
            )
plots_body_legs_significants
```


## TRAJECTORY
### Z velocity
- - - > + by age 
- Age, Sex, Hinge, and Age&Sex interaction significant. 
- Male faster but gets slower by age. 
- before and after breakpoint significant.
#### lmer
```{r}
# Use precise breakpoint in hinge term
flights$hinge <- pmax(flights$Age - 32.9, 0)

lmer_vz <- lmer(mean_vz ~ Age*Sex + hinge+ (1|Brood/Bird), data = flights)

# computing residuals
raw_cond_vz <- compute_redres(lmer_vz)

hist(raw_cond_vz) #normality in histogram
plot_redres(lmer_vz) #heteroskedasticity
plot_resqq(lmer_vz) #normality of residuals
plot_ranef(lmer_vz) 
## Oops not so well behaving
summary(lmer_vz)
Anova(lmer_vz, type =3, test.statistic = "Chisq") # Does age affect bodyweight overall?


linearHypothesis(lmer_vz, "Age = 0") # Is the slope before the breakpoint flat? -> no
linearHypothesis(lmer_vz, "Age+hinge = 0") # Is the slope after the breakpoint flat? -> no


```
#### Plot
```{r}
pred_vz <- ggpredict(lmer_vz, terms = c("Age [all]", "hinge"))

# Create hinge variable
breakpoint <- 32.9
flights$hinge <- pmax(flights$Age - breakpoint, 0)

# Get fitted values across Age
newdat <- data.frame(
  Age = seq(min(flights$Age), max(flights$Age), by = 1),
  Sex = "F",        # or average over Sex
  Brood = mean(flights$Brood),
  hinge = pmax(seq(min(flights$Age), max(flights$Age), by = 1) - breakpoint, 0)
)

newdat$pred <- predict(lmer_vz, newdat, re.form = NA)


plot_vz_final <-
  plot_vz_mean +
  geom_line(
    data = newdat,
    aes(x = Age, y = pred),
    color = "#1f77b4", size = 1.5, inherit.aes = FALSE, #linetype = "dashed"
  ) +
  #geom_line(
  #  data = newdat,
  #  aes(x = Age, y = pred),color = "red",
  #  size = 1.2, inherit.aes = FALSE
  #) +
  geom_vline(xintercept = 32.9, linetype = "dashed", color = "red")

plot_vz_final

```

### Y velocity
- increase by age then plateau
- Age and hinge significant 
- before and after breakpoint significant
#### lmer
```{r}
# Use precise breakpoint in hinge term
flights$hinge <- pmax(flights$Age - 33.0, 0)

lmer_vy <- lmer(mean_vy ~ Age*Sex + hinge+ (1|Brood/Bird), data = flights)

# computing residuals
raw_cond_vy <- compute_redres(lmer_vy)

hist(raw_cond_vy) #normality in histogram
plot_redres(lmer_vy) #heteroskedasticity
plot_resqq(lmer_vy) #normality of residuals
plot_ranef(lmer_vy) 
## Oops not so well behaving
summary(lmer_vy)
Anova(lmer_vy, type =3, test.statistic = "Chisq") # Does age affect bodyweight overall?


linearHypothesis(lmer_vy, "Age = 0") # Is the slope before the breakpoint flat? -> no
linearHypothesis(lmer_vy, "Age+hinge = 0") # Is the slope after the breakpoint flat? -> no

# Brood singular 
```

#### Plot
```{r}
pred_vy <- ggpredict(lmer_vy, terms = c("Age [all]", "hinge"))

# Create hinge variable
breakpoint <- 33.0
flights$hinge <- pmax(flights$Age - breakpoint, 0)

# Get fitted values across Age
newdat <- data.frame(
  Age = seq(min(flights$Age), max(flights$Age), by = 1),
  Sex = "F",        # or average over Sex
  Brood = mean(flights$Brood),
  hinge = pmax(seq(min(flights$Age), max(flights$Age), by = 1) - breakpoint, 0)
)

newdat$pred <- predict(lmer_vy, newdat, re.form = NA)


plot_vy_final <-
  plot_vy_mean +
  geom_line(
    data = newdat,
    aes(x = Age, y = pred),
    color = "#1f77b4", size = 1.5, inherit.aes = FALSE, #linetype = "dashed"
  ) +
  #geom_line(
  #  data = newdat,
  #  aes(x = Age, y = pred),color = "red",
  #  size = 1.2, inherit.aes = FALSE
  #) +
  geom_vline(xintercept = 33.0, linetype = "dashed", color = "red")

plot_vy_final

```

### Y acceleration
- decreases by age then plateau
- Age and Hinge significant, Age:Sex (p = 0.099)
- slope before and after significant. not 0. 
#### lmer
```{r}
# Use precise breakpoint in hinge term
flights$hinge <- pmax(flights$Age - 33.7, 0)

lmer_accy <- lmer(mean_acc_y ~ Age*Sex + hinge+ (1|Brood/Bird), data = flights)

# computing residuals
raw_cond_accy <- compute_redres(lmer_accy)

hist(raw_cond_accy) #normality in histogram
plot_redres(lmer_accy) #heteroskedasticity
plot_resqq(lmer_accy) #normality of residuals
plot_ranef(lmer_accy) 
## Oops not so well behaving
summary(lmer_accy)
Anova(lmer_accy, type =3, test.statistic = "Chisq") # Does age affect bodyweight overall?


linearHypothesis(lmer_accy, "Age = 0") # Is the slope before the breakpoint flat? -> no
linearHypothesis(lmer_accy, "Age+hinge = 0") # Is the slope after the breakpoint flat? -> no

# Brood singular 
```
#### Plot
```{r}
pred_accy <- ggpredict(lmer_accy, terms = c("Age [all]", "hinge"))

# Create hinge variable
breakpoint <- 33.7
flights$hinge <- pmax(flights$Age - breakpoint, 0)

# Get fitted values across Age
newdat <- data.frame(
  Age = seq(min(flights$Age), max(flights$Age), by = 1),
  Sex = "F",        # or average over Sex
  Brood = mean(flights$Brood),
  hinge = pmax(seq(min(flights$Age), max(flights$Age), by = 1) - breakpoint, 0)
)

newdat$pred <- predict(lmer_accy, newdat, re.form = NA)


plot_accy_final <-
  plot_accy_mean +
  geom_line(
    data = newdat,
    aes(x = Age, y = pred),
    color = "#1f77b4", size = 1.5, inherit.aes = FALSE, #linetype = "dashed"
  ) +
  #geom_line(
  #  data = newdat,
  #  aes(x = Age, y = pred),color = "red",
  #  size = 1.2, inherit.aes = FALSE
  #) +
  geom_vline(xintercept = 33.7, linetype = "dashed", color = "red")

plot_accy_final
```

### tangential acceleration
- decreases by age then plateau
- Age and Hinge significant
- slope before and after significant. not 0. 
#### lmer
```{r}
# Use precise breakpoint in hinge term
flights$hinge <- pmax(flights$Age - 33.8, 0)

lmer_accres <- lmer(mean_acc_res ~ Age*Sex + hinge+ (1|Brood/Bird), data = flights)
# computing residuals
raw_cond_accres <- compute_redres(lmer_accres)

hist(raw_cond_accres) #normality in histogram
plot_redres(lmer_accres) #heteroskedasticity
plot_resqq(lmer_accres) #normality of residuals
plot_ranef(lmer_accres) 
## Oops not so well behaving
summary(lmer_accres)
Anova(lmer_accres, type =3, test.statistic = "Chisq") # Does age affect bodyweight overall?


linearHypothesis(lmer_accres, "Age = 0") # Is the slope before the breakpoint flat? -> no
linearHypothesis(lmer_accres, "Age+hinge = 0") # Is the slope after the breakpoint flat? -> no

# Brood singular 
```
#### Plot
```{r}
pred_accres <- ggpredict(lmer_accres, terms = c("Age [all]", "hinge"))

# Create hinge variable
breakpoint <- 33.8
flights$hinge <- pmax(flights$Age - breakpoint, 0)

# Get fitted values across Age
newdat <- data.frame(
  Age = seq(min(flights$Age), max(flights$Age), by = 1),
  Sex = "F",        # or average over Sex
  Brood = mean(flights$Brood),
  hinge = pmax(seq(min(flights$Age), max(flights$Age), by = 1) - breakpoint, 0)
)

newdat$pred <- predict(lmer_accres, newdat, re.form = NA)


plot_accres_final <-
  plot_accres_mean +
  geom_line(
    data = newdat,
    aes(x = Age, y = pred),
    color = "#1f77b4", size = 1.5, inherit.aes = FALSE, #linetype = "dashed"
  ) +
  #geom_line(
  #  data = newdat,
  #  aes(x = Age, y = pred),color = "red",
  #  size = 1.2, inherit.aes = FALSE
  #) +
  geom_vline(xintercept = 33.8, linetype = "dashed", color = "red")

plot_accres_final
```
### tangential velocity
#### lmer
- Age and Hinge significant
- before and after breakpoint significant
```{r}
# Use precise breakpoint in hinge term
flights$hinge <- pmax(flights$Age - 34.3, 0)

lmer_vres <- lmer(mean_v_res ~ Age*Sex + hinge+ (1|Brood/Bird), data = flights)
# computing residuals
raw_cond_vres <- compute_redres(lmer_vres)

hist(raw_cond_vres) #normality in histogram
plot_redres(lmer_vres) #heteroskedasticity
plot_resqq(lmer_vres) #normality of residuals
plot_ranef(lmer_vres) 
## Oops not so well behaving
summary(lmer_vres)
Anova(lmer_vres, type =3, test.statistic = "Chisq") # Does age affect bodyweight overall?


linearHypothesis(lmer_vres, "Age = 0") # Is the slope before the breakpoint flat? -> no
linearHypothesis(lmer_vres, "Age+hinge = 0") # Is the slope after the breakpoint flat? -> no

# Brood singular 
```
#### Plot
```{r}
pred_vres <- ggpredict(lmer_vres, terms = c("Age [all]", "hinge"))

# Create hinge variable
breakpoint <- 34.3
flights$hinge <- pmax(flights$Age - breakpoint, 0)

# Get fitted values across Age
newdat <- data.frame(
  Age = seq(min(flights$Age), max(flights$Age), by = 1),
  Sex = "F",        # or average over Sex
  Brood = mean(flights$Brood),
  hinge = pmax(seq(min(flights$Age), max(flights$Age), by = 1) - breakpoint, 0)
)

newdat$pred <- predict(lmer_vres, newdat, re.form = NA)


plot_vres_final <-
  plot_vres_mean +
  geom_line(
    data = newdat,
    aes(x = Age, y = pred),
    color = "#1f77b4", size = 1.5, inherit.aes = FALSE, #linetype = "dashed"
  ) +
  #geom_line(
  #  data = newdat,
  #  aes(x = Age, y = pred),color = "red",
  #  size = 1.2, inherit.aes = FALSE
  #) +
  geom_vline(xintercept = 34.3, linetype = "dashed", color = "red")

plot_vres_final
```
### turn rate
#### lmer
- Age and Hinge significant
- before and after breakpoint significant
```{r}
# Use precise breakpoint in hinge term
flights$hinge <- pmax(flights$Age - 41.7, 0)

lmer_tr <- lmer(mean_turn_rate ~ Age*Sex + hinge+ (1|Brood/Bird), data = flights)
# computing residuals
raw_cond_tr <- compute_redres(lmer_tr)

hist(raw_cond_tr) #normality in histogram
plot_redres(lmer_tr) #heteroskedasticity
plot_resqq(lmer_tr) #normality of residuals
plot_ranef(lmer_tr) 
## Oops not so well behaving
summary(lmer_tr)
Anova(lmer_tr, type =3, test.statistic = "Chisq") # Does age affect bodyweight overall?


linearHypothesis(lmer_tr, "Age = 0") # Is the slope before the breakpoint flat? -> no
linearHypothesis(lmer_tr, "Age+hinge = 0") # Is the slope after the breakpoint flat? -> no

# Brood singular 
```
#### Plot
```{r}
pred_tr <- ggpredict(lmer_tr, terms = c("Age [all]", "hinge"))

# Create hinge variable

# Get fitted values across Age
newdat <- data.frame(
  Age = seq(min(flights$Age), max(flights$Age), by = 1),
  Sex = "F",        # or average over Sex
  Brood = mean(flights$Brood),
  hinge = pmax(seq(min(flights$Age), max(flights$Age), by = 1) - 41.7, 0)
)

newdat$pred <- predict(lmer_tr, newdat, re.form = NA)


plot_tr_final <-
  plot_tr_mean +
  geom_line(
    data = newdat,
    aes(x = Age, y = pred),
    color = "#1f77b4", size = 1.5, inherit.aes = FALSE, #linetype = "dashed"
  ) +
  #geom_line(
  #  data = newdat,
  #  aes(x = Age, y = pred),color = "red",
  #  size = 1.2, inherit.aes = FALSE
  #) +
  geom_vline(xintercept = 41.7, linetype = "dashed", color = "red")

plot_tr_final
```

### Z acceleration
- Age, Hinge, and Age*Sex significant
- Z acceleration increase 0.04 m/s^2 per day up to 48.3 days then decrease after.
#### lmer
```{r}

# Use precise breakpoint in hinge term
flights$hinge <- pmax(flights$Age - 33.7, 0)

lmer_accz <- lmer(mean_acc_z ~ Age*Sex + hinge+ (1|Brood/Bird), data = flights)
# computing residuals
raw_cond_accz <- compute_redres(lmer_accz)

hist(raw_cond_accz) #normality in histogram
plot_redres(lmer_accz) #heteroskedasticity
plot_resqq(lmer_accz) #normality of residuals
plot_ranef(lmer_accz) 
## Oops not so well behaving
summary(lmer_accz)
Anova(lmer_accz, type =3, test.statistic = "Chisq") # Does age affect bodyweight overall?


linearHypothesis(lmer_accz, "Age = 0") 
# Is the slope before the breakpoint flat? -> no

linearHypothesis(lmer_accz, "Age+hinge = 0") 
# Is the slope after the breakpoint flat? -> no

```
#### plot
```{r}
pred_vres <- ggpredict(lmer_accz, terms = c("Age [all]", "hinge"))

# Get fitted values across Age
newdat <- data.frame(
  Age = seq(min(flights$Age), max(flights$Age), by = 1),
  Sex = "F",        # or average over Sex
  Brood = mean(flights$Brood),
  hinge = pmax(seq(min(flights$Age), max(flights$Age), by = 1) - 33.7, 0)
)

newdat$pred <- predict(lmer_accz, newdat, re.form = NA)


plot_accz_final <-
  plot_accz_mean +
  geom_line(
    data = newdat,
    aes(x = Age, y = pred),
    color = "#1f77b4", size = 1.5, inherit.aes = FALSE, #linetype = "dashed"
  ) +
  #geom_line(
  #  data = newdat,
  #  aes(x = Age, y = pred),color = "red",
  #  size = 1.2, inherit.aes = FALSE
  #) +
  geom_vline(xintercept = 33.7, linetype = "dashed", color = "red")

plot_accz_final



```

### X velocity
#### lmer
- diagnostics looking really good 

```{r}

# Use precise breakpoint in hinge term
flights$hinge <- pmax(flights$Age - 59.9, 0)

lmer_vx <- lmer(mean_vx ~ Age*Sex + hinge+ (1|Brood/Bird), data = flights)
# computing residuals
raw_cond_vx <- compute_redres(lmer_vx)

hist(raw_cond_vx) #normality in histogram
plot_redres(lmer_vx) #heteroskedasticity
plot_resqq(lmer_vx) #normality of residuals
plot_ranef(lmer_vx) 
## Oops not so well behaving
summary(lmer_vx)
Anova(lmer_vx, type =3, test.statistic = "Chisq") # Does age affect bodyweight overall?


linearHypothesis(lmer_vx, "Age = 0") 
# Is the slope before the breakpoint flat? -> no

linearHypothesis(lmer_vx, "Age+hinge = 0") 
# Is the slope after the breakpoint flat? -> no

```
#### plot
```{r}
pred_vx <- ggpredict(lmer_vx, terms = c("Age [all]", "hinge"))

# Get fitted values across Age
newdat <- data.frame(
  Age = seq(min(flights$Age), max(flights$Age), by = 1),
  Sex = "F",        # or average over Sex
  Brood = mean(flights$Brood),
  hinge = pmax(seq(min(flights$Age), max(flights$Age), by = 1) - 59.9, 0)
)

newdat$pred <- predict(lmer_vx, newdat, re.form = NA)

plot_vx_final <-
  plot_vx_mean +
  geom_line(
    data = newdat,
    aes(x = Age, y = pred),
    color = "#1f77b4", size = 1.5, inherit.aes = FALSE, #linetype = "dashed"
  ) +
  #geom_line(
  #  data = newdat,
  #  aes(x = Age, y = pred),color = "red",
  #  size = 1.2, inherit.aes = FALSE
  #) +
  geom_vline(xintercept = 59.9, linetype = "dashed", color = "red")

plot_vx_final



```
### X acceleration
#### lmer
- diagnostics looking really good 
- Age, Sex, Hinge, Age:Sex, basically everything significant
- 
```{r}

# Use precise breakpoint in hinge term
flights$hinge <- pmax(flights$Age - 74.6, 0)

lmer_accx <- lmer(mean_acc_x ~ Age*Sex + hinge+ (1|Brood/Bird), data = flights)
# computing residuals
raw_cond_accx <- compute_redres(lmer_accx)

hist(raw_cond_accx) #normality in histogram
plot_redres(lmer_accx) #heteroskedasticity
plot_resqq(lmer_accx) #normality of residuals
plot_ranef(lmer_accx) 
## Oops not so well behaving
summary(lmer_accx)
Anova(lmer_accx, type =3, test.statistic = "Chisq") # Does age affect bodyweight overall?


linearHypothesis(lmer_accx, "Age = 0") 
# Is the slope before the breakpoint flat? -> no

linearHypothesis(lmer_accx, "Age+hinge = 0") 
# Is the slope after the breakpoint flat? -> no

```

#### plot
```{r}
pred_accx <- ggpredict(lmer_accx, terms = c("Age [all]", "hinge"))

# Get fitted values across Age
newdat <- data.frame(
  Age = seq(min(flights$Age), max(flights$Age), by = 1),
  Sex = "F",        # or average over Sex
  Brood = mean(flights$Brood),
  hinge = pmax(seq(min(flights$Age), max(flights$Age), by = 1) - 74.6, 0)
)

newdat$pred <- predict(lmer_accx, newdat, re.form = NA)

plot_accx_final <-
  plot_accx_mean +
  geom_line(
    data = newdat,
    aes(x = Age, y = pred),
    color = "#1f77b4", size = 1.5, inherit.aes = FALSE, #linetype = "dashed"
  ) +
  #geom_line(
  #  data = newdat,
  #  aes(x = Age, y = pred),color = "red",
  #  size = 1.2, inherit.aes = FALSE
  #) +
  geom_vline(xintercept = 74.6, linetype = "dashed", color = "red")

plot_accx_final



```
### downstroke - n.s.
#### lmer
- diagnostics looking not too good. Brood singular  
- Just Age significant. 
- head displacement increases until 79.9, then almost plateau
- after breakpoint NOT SIGNIFICANT!
```{r}

# Use precise breakpoint in hinge term
flights$hinge <- pmax(flights$Age - 79.9, 0)

lmer_downstroke <- lmer(downstroke_amplitude_m ~ Age*Sex + hinge+ (1|Brood/Bird), data = flights)
# computing residuals
raw_cond_downstroke <- compute_redres(lmer_downstroke)

hist(raw_cond_downstroke) #normality in histogram
plot_redres(lmer_downstroke) #heteroskedasticity
plot_resqq(lmer_downstroke) #normality of residuals
plot_ranef(lmer_downstroke) 
## Oops not so well behaving
summary(lmer_downstroke)
Anova(lmer_downstroke, type =3, test.statistic = "Chisq") # Does age affect bodyweight overall?


linearHypothesis(lmer_downstroke, "Age = 0") 
# Is the slope before the breakpoint flat? -> no

linearHypothesis(lmer_downstroke, "Age+hinge = 0") 
# Is the slope after the breakpoint flat? -> no
# Brood singular
```
#### plot
```{r}
pred_downstroke <- ggpredict(lmer_downstroke, terms = c("Age [all]", "hinge"))

# Get fitted values across Age
newdat <- data.frame(
  Age = seq(min(flights$Age), max(flights$Age), by = 1),
  Sex = "F",        # or average over Sex
  Brood = mean(flights$Brood),
  hinge = pmax(seq(min(flights$Age), max(flights$Age), by = 1) - 79.9, 0)
)

newdat$pred <- predict(lmer_downstroke, newdat, re.form = NA)

plot_downstroke_final <-
  plot_downstroke_m +
  geom_line(
    data = newdat,
    aes(x = Age, y = pred),
    color = "#1f77b4", size = 1.5, inherit.aes = FALSE, #linetype = "dashed"
  ) +
  #geom_line(
  #  data = newdat,
  #  aes(x = Age, y = pred),color = "red",
  #  size = 1.2, inherit.aes = FALSE
  #) +
  geom_vline(xintercept = 79.9, linetype = "dashed", color = "red")

plot_downstroke_final



```
### wingbeat frequency - n.s.
- Age, hinge, and Age:Sex significant
- wingbeat frequency decreases 0.03Hz/day and at 42.9 day increases 0.03Hz/day
#### lmer
```{r}
# Use precise breakpoint in hinge term
flights$hinge <- pmax(flights$Age - 42.9, 0)

lmer_wb <- lmer(wingbeat_frequency ~ Age*Sex + hinge+ (1|Brood/Bird), data = flights)
# computing residuals
raw_cond_wb <- compute_redres(lmer_wb)

hist(raw_cond_wb) #normality in histogram
plot_redres(lmer_wb) #heteroskedasticity
plot_resqq(lmer_wb) #normality of residuals
plot_ranef(lmer_wb) 
## Oops not so well behaving
summary(lmer_wb)
Anova(lmer_wb, type =3, test.statistic = "Chisq") # Does age affect bodyweight overall?


linearHypothesis(lmer_wb, "Age = 0") # Is the slope before the breakpoint flat? -> no
linearHypothesis(lmer_wb, "Age+hinge = 0") # Is the slope after the breakpoint flat? -> no

# Brood singular 
```
#### plot
```{r}
pred_vres <- ggpredict(lmer_wb, terms = c("Age [all]", "hinge"))

# Get fitted values across Age
newdat <- data.frame(
  Age = seq(min(flights$Age), max(flights$Age), by = 1),
  Sex = "F",        # or average over Sex
  Brood = mean(flights$Brood),
  hinge = pmax(seq(min(flights$Age), max(flights$Age), by = 1) - breakpoint, 0)
)

newdat$pred <- predict(lmer_wb, newdat, re.form = NA)


plot_wb_final <-
  plot_wb_mean +
  geom_line(
    data = newdat,
    aes(x = Age, y = pred),
    color = "#1f77b4", size = 1.5, inherit.aes = FALSE, #linetype = "dashed"
  ) +
  #geom_line(
  #  data = newdat,
  #  aes(x = Age, y = pred),color = "red",
  #  size = 1.2, inherit.aes = FALSE
  #) +
  geom_vline(xintercept = 42.9, linetype = "dashed", color = "red")

plot_wb_final
```

### Final plots
```{r}

plots_trajectories_significants <-
  ggarrange(plot_vx_final, plot_vy_final, plot_vz_final, 
          plot_accx_final, plot_accy_final, plot_accz_final,
          plot_tr_final, plot_vres_final, plot_accres_final,
          
          ncol = 3, nrow=3
          )

```
## wing area
- nothing significant
```{r}

#breakpoint not significant
#wing_sorted$hinge <- pmax(wing_sorted$Age - 34.3, 0)

#m model
lmer_wa <- lmer(wingarea ~ Age*Sex + Brood + (1|Bird), data = wing_sorted)

# computing residuals
raw_cond_wa <- compute_redres(lmer_wa)

hist(raw_cond_wa) #normality in histogram
plot_redres(lmer_wa) #heteroskedasticity
plot_resqq(lmer_wa) #normality of residuals
plot_ranef(lmer_wa) 
# very good fit as lmer


#testing 
summary(lmer_wa)
Anova(lmer_wa, type =3, test.statistic = "Chisq")

#hypothesis testing
linearHypothesis(lmer_wa, "Age = 0") # Is the slope before the breakpoint flat? -> no
#linearHypothesis(lmer_wa, "Age+hinge = 0") # Is the slope after the breakpoint flat? -> no

```
