---
title: "zebra-force"
author: "Kyu Min Huh"
date: "2025-12-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Aims 

In this markdown, I aim to see the age effect on the leg force at take-off. 
Aims/Steps:

1. show morphological trend with age
- Graphs: a. body mass x age, b. wing area x age, c. wing chord length x age, d. tarsus length x age. 

2. See if there's any visible trend in raw data with age
- Graphs: a. max resultant force x age, with each individual marked by point and whiskers b. horizontal force at the time of max resultant force x age c. vertical force at the time of max resultant force x age d. impulse x age e. impulse ratio x age
f. take-off angle x age 

# Libraries
```{r libraries}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(ggeffects)

library(rptR)
library(ggpubr)
library(segmented)

library(nlme)
library(lme4)
library(lmerTest)
library(emmeans)
library(PerformanceAnalytics)
library(MuMIn)
library(lmtest)
library(lmerTest)
library(AICcmodavg)
library(devtools)
library(redres)
#library(robustlmm)
library(car)
library(ggeffects)
library(gvlma)
library(performance)
library(sjPlot)
library(multcomp)
library(EnvStats)
library(vegan)
```

# I. Raw data plots (mean+-SD)
## load the dataset 
### force_data
```{r load the datasets}

# force data
force_data <- read.delim(file = "C:\\Users\\jesu4036\\Documents\\flight-ontogeny\\data\\takeoff\\force_final.csv", sep = ",") %>% distinct()
#edited greenR-orangeL age40 Takeoff 3, 5 and pinkR-blueL 40 takeoff 1 to Takeoff 1 and 6. 

nas_filled <- read.delim(file = "C:\\Users\\jesu4036\\Documents\\flight-ontogeny\\data\\takeoff\\nas_filled.csv", sep = ",") %>% distinct()
nas_filled <- nas_filled[,-1]

dim(force_data)

force_data <- force_data %>%
  mutate(angle = as.numeric(angle)) %>%
  mutate(Fx_maxFtot = as.numeric(Fx_maxFtot)) %>%
  mutate(Fy_maxFtot = as.numeric(Fy_maxFtot)) %>%
  mutate(Fz_maxFtot = as.numeric(Fz_maxFtot)) 

nas <- force_data[is.na(force_data$Fx_maxFtot),]
nas

force_data <- force_data %>%
  filter(!FlightID %in% nas$FlightID)

dim(force_data)
```

```{r}
force_data <- rbind(force_data, nas_filled) # adding the added rows. 

dim(force_data)

# check if the nas are fixed ---> should return zero rows. 
force_data[is.na(force_data$Fx_maxFtot),]

force_data <- force_data %>%
  mutate(Fx_maxFtot_bw = abs(Fx_maxFtot)/bodyweight) %>%
  mutate(Fy_maxFtot_bw = abs(Fy_maxFtot)/bodyweight) %>%
  mutate(Fz_maxFtot_bw = abs(Fz_maxFtot)/bodyweight)
```

```{r}
# changing the unit to mN
# more suitable for birds that weigh in grams not kilograms
#force_data$max_Ftot = force_data$max_Ftot*1000

# newton to gram conversion
force_data$bodyweight_g = force_data$bodyweight * 101.9716

# make relative max res force
force_data$max_Ftot_bw = force_data$max_Ftot/force_data$bodyweight

# other relative force as well 





# 'Age': change 81->80 and 61->60 

force_data$Age[force_data$Age == 81] <- 80
force_data$Age[force_data$Age == 61] <- 60

#outlier that did not have flight data
force_data <- force_data %>%
  filter(!FlightID %in% c( 3702006, 2604005, 1404001, 2902805))
#3702006 outlier (0.8) in impulse that didn't have flight data
#2604005, 1404001, 2902805 => outliers in takeoff angle that don't have flight data and seemed to have taken off opposite directions. 
```

### wing area data
```{r}
wing_data <- read.delim(file = "C:\\Users\\jesu4036\\Documents\\flight-ontogeny\\data\\morphology\\wingdata_names_edited.csv", sep = ",") %>% distinct()

scale_weight <- read.delim(file = "C:\\Users\\jesu4036\\Documents\\flight-ontogeny\\data\\morphology\\bodyweight_measured.csv", sep = ",")


wing_data_sorted <- wing_data %>%
  # sort by Bird (alphabetical) and Age (increasing)
  arrange(Bird, Age) %>%
  # keep only birds with at least 3 age records
  group_by(Bird) %>%
  filter(n_distinct(Age) >= 3) %>%
  mutate(wingarea = wingarea/10000) %>% #converting mm^2 -> m^2
  ungroup() %>%
  left_join(scale_weight, by = c("Bird", "Age")) %>%
  mutate(bodyweight_measured = as.numeric(bodyweight_measured)) %>%
  mutate(bodyweight_measured = bodyweight_measured * 0.00980665) %>%
  mutate(wing_loading = wingarea/bodyweight_measured) %>%
  dplyr::select('Bird','Age','wingspan.calc','wingarea','bodyweight_measured','wing_loading')

dim(wing_data_sorted)

wing_data_final <- wing_data_sorted %>%
  group_by(Bird, Age) %>%
  summarise(
    wingspan = mean(wingspan.calc, na.rm = TRUE),
    wingarea = mean(wingarea, na.rm = TRUE),
    scale_weight = mean(bodyweight_measured, na.rm = TRUE),
    wing_loading = mean(wing_loading, na.rm = TRUE),
    .groups = "drop"
  ) %>% drop_na()

dim(wing_data_final) #192

```

### sex and brood measured bodyweight data 
```{r}
sex_data <- read.delim(file = "C:\\Users\\jesu4036\\Documents\\flight-ontogeny\\data\\morphology\\birds_sex_brood.csv", sep = ",") %>% distinct()

force_data <- merge(force_data, sex_data, by = "Bird")
wing_data_final <- merge(wing_data_final, sex_data, by = "Bird")


```

### flight trajectory data
```{r}
flights <- read.delim(file = "C:\\Users\\jesu4036\\Documents\\flight-ontogeny\\data\\takeoff\\flight_outcome_full.csv", sep = ",") %>% left_join(sex_data, by = "Bird") 

raw_flight <- read.delim(file = "C:\\Users\\jesu4036\\Documents\\flight-ontogeny\\data\\takeoff\\clean_filtered_flights.csv", sep =",")
colnames(flights)
colnames(raw_flight)
```
## one flight
```{r}
#cyanR-blackL, Age 36, trial 1
#test_flight1 <- raw_flight %>%
 #filter(FlightID == 2303601.0)

test_flight <- raw_flight %>%
  filter(FlightID == 2303601.0)
```
## RAW PLOTS
```{r}

library(ggplot2)
library(patchwork)
df <- test_flight

# Position plots
p1 <- ggplot(df, aes(cumul_time, X)) +
  geom_line(color = "blue") +
  geom_point(color = "darkblue") +
  ggtitle("Position X")

p2 <- ggplot(df, aes(cumul_time, Y)) +
  geom_line(color = "green") +
  geom_point(color = "darkgreen") +
  ggtitle("Position Y")

p3 <- ggplot(df, aes(cumul_time, Z)) +
  geom_line(color = "red") +
  geom_point(color = "darkred") +
  ggtitle("Position Z")

# Velocity plots
p4 <- ggplot(df, aes(cumul_time, inst_V_X)) +
  geom_line(color = "blue") +
  geom_point(color = "darkblue") +
  ggtitle("Velocity X")

p5 <- ggplot(df, aes(cumul_time, inst_V_Y)) +
  geom_line(color = "green") +
  geom_point(color = "darkgreen") +
  ggtitle("Velocity Y")

p6 <- ggplot(df, aes(cumul_time, inst_V_Z)) +
  geom_line(color = "red") +
  geom_point(color = "darkred") +
  ggtitle("Velocity Z")

# Acceleration plots
p7 <- ggplot(df, aes(cumul_time, inst_acc_X)) +
  geom_line(color = "blue") +
  geom_point(color = "darkblue") +
  ggtitle("Acceleration X")

p8 <- ggplot(df, aes(cumul_time, inst_acc_Y)) +
  geom_line(color = "green") +
  geom_point(color = "darkgreen") +
  ggtitle("Acceleration Y")

p9 <- ggplot(df, aes(cumul_time, inst_acc_Z)) +
  geom_line(color = "red") +
  geom_point(color = "darkred") +
  ggtitle("Acceleration Z")


# New 5 plots
p10 <- ggplot(df, aes(cumul_time, inst_V_res)) +
  geom_line(color="purple") + geom_point(color="purple") + ggtitle("Velocity Resultant")

p11 <- ggplot(df, aes(cumul_time, inst_acc_tan)) +
  geom_line(color="orange") + geom_point(color="darkorange") + ggtitle("Tangential acceleration")

p12 <- ggplot(df, aes(cumul_time, inst_acc_mag)) +
  geom_line(color="orange") + geom_point(color="darkorange") + ggtitle("Acceleration magnitude")


p13 <- ggplot(df, aes(cumul_time, Turn_Rate_degrees)) +
  geom_line(color="brown") + geom_point(color="brown") + ggtitle("Turn Rate (degrees)")

p14 <- ggplot(df, aes(cumul_time, Centripetal_acc)) +
  geom_line(color="pink") + geom_point(color="deeppink") + ggtitle("Centripetal Acceleration")

# Combine all plots
final_plot2 <- (p1 | p2 | p3) /
              (p4 | p5 | p6) /
              (p7 | p8 | p9) /
              (p10 | p11 | p12)/
              (p13| p14)

final_plot2




```
### function plot_raw_base
```{r}

plot_raw_base <- function(variable, df, title){
  
  
  plot <- ggplot(df, aes(x = Age, y = {{variable}}, color = Bird)) +
    geom_point(position = position_jitter(width = 1), alpha = 0.3) +
    scale_x_continuous(breaks = unique(df$Age)) +
    labs(x = "Age", y = title, title = "") +
    theme(legend.position = "none")+
    guides(alpha = "none", color = "none") +
    theme_minimal()
  
  return(plot)
  
}


```
### function plot_age_var 
```{r}
plot_age_var <- function(variable, df, title) {
  
  # Summary by individual
  summary_df <- df %>%
    group_by(Bird, Age) %>%
    summarise(
      mean = mean({{variable}}, na.rm = TRUE),
      sd = sd({{variable}}, na.rm = TRUE),
      n = n(),
      .groups = "drop"
    )
  
  # Summary by age
  trend_df <- df %>%
    group_by(Age) %>%
    summarise(
      mean = mean({{variable}}, na.rm = TRUE),
      sd = sd({{variable}}, na.rm = TRUE),
      n = n(),
      .groups = "drop"
    )
  
  # Plot
  plot <- ggplot(df, aes(x = Age, y = {{variable}}, color = Bird)) +
    geom_point(position = position_jitter(width = 1), alpha = 0.3) +
    geom_errorbar(
      data = summary_df,
      aes(x = Age, ymin = mean - sd, ymax = mean + sd, color = Bird, alpha = 0.3),
      width = 2, size = 0.8,
      inherit.aes = FALSE
    ) +
    geom_point(
      data = summary_df,
      aes(y = mean, x = Age, color = Bird, alpha = 0.9),
      size = 2, shape = 18,
      inherit.aes = FALSE
    ) +
    geom_line(
      data = summary_df,
      aes(x = Age, y = mean, color = Bird, group = Bird, alpha = 0.8),
      size = 0.7,
      inherit.aes = FALSE
    ) +
    geom_ribbon(
      data = trend_df,
      aes(x = Age, ymin = mean - sd, ymax = mean + sd),
      fill = "grey70", alpha = 0.5,
      inherit.aes = FALSE
    ) +
    geom_line(
      data = trend_df,
      aes(x = Age, y = mean),
      ,color = "grey50", alpha = 0.8, size = 1.2,
      inherit.aes = FALSE
    ) +
    scale_x_continuous(breaks = unique(df$Age)) +
    labs(x = "Age", y = title, title = "") +
    guides(alpha = "none", color = "none") +
    theme(legend.position = "none")+
    theme_minimal()
  
  return(plot)
}
```

### function plot_del_var
```{r}


plot_del_var <- function(variable, df, title, drop_no_baseline = TRUE) {
  # Requires: dplyr, ggplot2
  # variable: unquoted column name (tidy-eval), e.g., BW or Mass
  # df: data.frame/tibble with columns Bird, Age, and the variable
  # title: y-axis title base text
  # baseline_age: age used as the zero point (default 20)
  # drop_no_baseline: if TRUE, drop birds without observations at baseline_age
  
  library(dplyr)
  library(ggplot2)
  
  # Get per-bird baseline at the specified age
  baseline_df <- df %>%
    dplyr::filter(Age == 20) %>%
    dplyr::group_by(Bird) %>%
    dplyr::summarise(
      baseline = mean({{ variable }}, na.rm = TRUE),
      .groups = "drop"
    )
  
  # Join baseline back and compute delta for every row
  df_delta <- df %>%
    dplyr::left_join(baseline_df, by = "Bird") %>%
    dplyr::mutate(delta = {{ variable }} - baseline)
  
  # Optionally drop birds without a baseline
  if (isTRUE(drop_no_baseline)) {
    df_delta <- df_delta %>% dplyr::filter(!is.na(baseline))
  }
  
  # Summary by individual (on delta)
  summary_df <- df_delta %>%
    dplyr::group_by(Bird, Age) %>%
    dplyr::summarise(
      mean = mean(delta, na.rm = TRUE),
      sd   = sd(delta,   na.rm = TRUE),
      n    = dplyr::n(),
      .groups = "drop"
    )
  
  # Summary by age (overall trend on delta)
  trend_df <- df_delta %>%
    dplyr::group_by(Age) %>%
    dplyr::summarise(
      mean = mean(delta, na.rm = TRUE),
      sd   = sd(delta,   na.rm = TRUE),
      n    = dplyr::n(),
      .groups = "drop"
    )
  
  # Plot (all on delta scale)
  p <- ggplot(df_delta, aes(x = Age, y = delta, color = Bird)) +
    # raw points (delta)
    geom_point(position = position_jitter(width = 1), alpha = 0.3) +
    
    # within-bird mean ± sd
    geom_errorbar(
      data = summary_df,
      aes(x = Age, ymin = mean - sd, ymax = mean + sd, color = Bird),
      width = 2, size = 0.8, alpha = 0.3, inherit.aes = FALSE
    ) +
    geom_point(
      data = summary_df,
      aes(y = mean, x = Age, color = Bird),
      size = 2, shape = 18, alpha = 0.9, inherit.aes = FALSE
    ) +
    geom_line(
      data = summary_df,
      aes(x = Age, y = mean, color = Bird, group = Bird),
      size = 0.7, alpha = 0.8, inherit.aes = FALSE
    ) +
    
    # overall trend ribbon (mean ± sd on delta)
    geom_ribbon(
      data = trend_df,
      aes(x = Age, ymin = mean - sd, ymax = mean + sd),
      fill = "grey70", alpha = 0.5, inherit.aes = FALSE
    ) +
    geom_line(
      data = trend_df,
      aes(x = Age, y = mean),
      color = "grey50", alpha = 0.8, size = 1.2, inherit.aes = FALSE
    ) +
    
    # zero reference line (baseline)
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
    
    scale_x_continuous(breaks = sort(unique(df$Age))) +
    labs(
      x = "Age",
      y = title,
      title = ""
    ) +
    guides(alpha = "none", color = "none") +
    theme_minimal() +
    theme(legend.position = "none")
  
  return(p)
}




```



### OUTLIERS detection
#### Force
```{r}
test_Ftot <- rosnerTest(force_data$max_Ftot_bw, k= 10)
test_Ftot$all.stats #no outliers

test_Fx <- rosnerTest(force_data$Fx_maxFtot_bw, k = 10)
test_Fx$all.stats #no outliers

test_Fy <- rosnerTest(force_data$Fy_maxFtot_bw, k = 10)
test_Fy$all.stats

test_Fz <- rosnerTest(force_data$Fz_maxFtot_bw, k = 10)
test_Fz$all.stats
```

```{r}
test_impulse <- rosnerTest(force_data$impulse, k = 10)
test_impulse$all.stats

test_duration <- rosnerTest(force_data$duration, k = 20)
test_duration$all.stats

test_angle <- rosnerTest(force_data$angle, k = 10)
test_angle$all.stats
```

```{r}

test_impulse_ratio <- rosnerTest(force_data$impulse, k = 10)
test_impulse_ratio$all.stats

test_left_impulse <- rosnerTest(force_data$left_impulse, k= 10)
test_left_impulse$all.stats

test_right_impulse <- rosnerTest(force_data$right_impulse, k = 10)
test_right_impulse$all.stats

test_left_duration <- rosnerTest(force_data$left_duration, k = 20)
test_left_duration$all.stats

test_right_duration <- rosnerTest(force_data$right_duration, k = 10)
test_right_duration$all.stats

```
```{r}

out_impulse <- test_impulse$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)


out_duration <- test_duration$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)


out_angle <- test_angle$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)


out_impulse_ratio <- test_impulse_ratio$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)


out_left_impulse <- test_left_impulse$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)


out_left_duration <- test_left_duration$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)


out_right_duration <- test_right_duration$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)


outliers_force <-rbind(out_impulse, out_duration, out_angle, out_impulse_ratio, out_left_impulse, out_left_duration, out_right_duration)

dim(outliers_force) 
 #41 outliers

```
```{r}
dim(force_data) #1003

force_clean <- force_data[-c(outliers_force$Obs.Num),]

dim(force_clean) #975
```

#### Trajectory
```{r}
test_vx <- rosnerTest(flights$mean_vx, k= 10)
test_vx$all.stats #no outliers

test_vy <- rosnerTest(flights$mean_vy, k = 10)
test_vy$all.stats #no outliers

test_vz <- rosnerTest(flights$mean_vz, k = 10)
test_vz$all.stats

test_vres <- rosnerTest(flights$mean_v_res, k = 10)
test_vres$all.stats
```

```{r}
test_accx <- rosnerTest(flights$mean_acc_x, k = 10)
test_accx$all.stats

test_accy <- rosnerTest(flights$mean_acc_y, k = 10)
test_accy$all.stats

test_accz <- rosnerTest(flights$mean_acc_z, k = 10)
test_accz$all.stats

test_accres <- rosnerTest(flights$mean_acc_tan, k = 10) #tangential acceleration
test_accres$all.stats
```
#### tortuosity & wingbeat
```{r}
test_tr <- rosnerTest(flights$mean_turn_rate, k = 20)
test_tr$all.stats

test_centri <- rosnerTest(flights$mean_centri_acc, k = 10)
test_centri$all.stats

test_wingbeat <- rosnerTest(flights$wingbeat_frequency, k = 10)
test_wingbeat$all.stats

test_upstroke <- rosnerTest(flights$upstroke_amplitude_m, k = 30)
test_upstroke$all.stats

test_downstroke <- rosnerTest(flights$downstroke_amplitude_m, k = 30)
test_downstroke$all.stats
```
```{r}

out_vres <- test_vres$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)


out_accx <- test_accx$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)


out_accy <- test_accy$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)


out_accz <- test_accz$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)


out_accres <- test_accres$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)


out_tr <- test_tr$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)


out_centri <- test_centri$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)

out_wingbeat <- test_wingbeat$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)

out_upstroke <- test_upstroke$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)

out_downstroke <- test_downstroke$all.stats %>%
  filter(Outlier == "TRUE") %>%
  dplyr::select(Obs.Num)


outliers_flights_speed <-rbind(out_vres, out_accx, out_accy, out_accz, out_accres) %>% distinct()

outliers_flights_tort <- rbind(out_tr, out_centri) %>% distinct()

outliers_flights_wing <- rbind(out_wingbeat, out_upstroke, out_downstroke) %>% distinct()

dim(outliers_flights_speed) # 19
dim(outliers_flights_tort) # 16
dim(outliers_flights_wing) # 36
```
```{r}
dim(flights) #937
#speed and acceleration variables

flights_speed  <- flights[-c(outliers_flights_speed$Obs.Num),]

# tortuosity variables 

flights_tort <- flights[-c(outliers_flights_tort$Obs.Num),]

# wingbeat and head displacement variables


flights_wing <- flights[-c(outliers_flights_wing$Obs.Num),]

dim(flights_speed) #918
dim(flights_tort) #921
dim(flights_wing) #901

```
## A. Morphology
### a. wing area
```{r}
plot_wingarea <- plot_age_var(wingarea, wing_data_final, expression(wing~area ~(m^2)))
plot_wingarea

plot_wingarea_delta <- plot_del_var(wingarea, wing_data_final, expression(Delta~wing~area ~(m^2)))
plot_wingarea_delta

plot_wingarea_base <- plot_raw_base(wingarea, wing_data_final, expression(wing~area ~(m^2)))
plot_wingarea_base


```



### b. calculated wingspan
```{r}
plot_wingspan <- plot_age_var(wingspan, wing_data_final, expression(wingspan~(mm)))
plot_wingspan

plot_wingspan_delta <- plot_del_var(wingspan, wing_data_final, expression(Delta~wingspan~(mm)))
plot_wingspan_delta

plot_wingspan_base <- plot_raw_base(wingspan, wing_data_final, expression(Delta~wingspan~(mm)))
plot_wingspan_base
```

### c. perching weight
```{r}
plot_perch_g <- plot_age_var(bodyweight_g, force_data, "perching weight(g)")
plot_perch_g

plot_perch_g_delta <- plot_del_var(bodyweight_g, force_data, "perching weight(g)")
plot_perch_g_delta

plot_perch_g_base <- plot_raw_base(bodyweight_g, force_data, "perching weight(g)")
plot_perch_g_base

```
### d. scale weight
```{r}
plot_scale_g <- plot_age_var(scale_weight*101.97, wing_data_final, expression(scale~mass~(g)))
plot_scale_g

plot_scale_g_delta <- plot_del_var(scale_weight*101.97, wing_data_final, expression(Delta~scale~mass~(g)))
plot_scale_g_delta

plot_scale_g_base <- plot_raw_base(scale_weight*101.97, wing_data_final, expression(Delta~scale~mass~(g)))
plot_scale_g_base
```
#### scaled mass vs. perching weight
```{r}
ggarrange(plot_perch_g, plot_scale_g, plot_perch_g_delta, plot_scale_g_delta, nrow = 2, ncol = 2)

```
### e. wing loading 
```{r}
plot_wingloading <- plot_age_var(wing_loading, wing_data_final, expression(wing~loading~(N/m^2)))
plot_wingloading

plot_wingloading_delta <- plot_del_var(wing_loading, wing_data_final, expression(Delta~wing~loading~(N/m^2)))
plot_wingloading_delta

plot_wingloading_base <- plot_raw_base(wing_loading, wing_data_final, expression(Delta~wing~loading~(N/m^2)))
plot_wingloading_base
```

### plots_morphology
#### raw
```{r}

plots_morphology <- ggarrange(plot_scale_g, plot_wingspan, plot_wingarea,  plot_wingloading, nrow = 2, ncol =2)

plots_morphology

```
#### delta
```{r}

plots_morphology_delta <- ggarrange(plot_scale_g_delta,plot_wingspan_delta, plot_wingarea_delta, plot_wingloading_delta, nrow = 2, ncol =2)

plots_morphology_delta


```
## B. Performance

### LEG FORCE
#### a. max resultant force
##### absolute
```{r}

plot_maxFtot_abs <- plot_age_var(max_Ftot, force_clean, expression(max~F~(N)))
plot_maxFtot_abs

plot_maxFtot_abs_delta <- plot_del_var(max_Ftot, force_clean, expression(max~F~(N)))
plot_maxFtot_abs_delta

plot_maxFtot_abs_base <- plot_raw_base(max_Ftot, force_clean, expression(max~F~(N)))
plot_maxFtot_abs_base

```
##### relative
```{r}

plot_maxFtot <- plot_age_var(max_Ftot_bw, force_clean, expression(max~F~"/"~bw~(N)))
plot_maxFtot

plot_maxFtot_delta <- plot_del_var(max_Ftot_bw, force_clean, expression(max~F~"/"~bw~(N)))
plot_maxFtot_delta

plot_maxFtot_base <- plot_raw_base(max_Ftot_bw, force_clean, expression(max~F~"/"~bw~(N)))
plot_maxFtot_base

```
#### c. horizontal force (Fy)
##### absolute
```{r}

plot_maxFy_abs <- plot_age_var(Fy_maxFtot, force_clean, expression(max~Fy~(N)))
plot_maxFy_abs

plot_maxFy_abs_delta <- plot_del_var(Fy_maxFtot, force_clean, expression(max~Fy~(N)))
plot_maxFy_abs_delta

plot_maxFy_abs_base <- plot_raw_base(Fy_maxFtot, force_clean, expression(max~Fy~(N)))
plot_maxFy_abs_base


```
##### relative
```{r}
plot_maxFy <- plot_age_var(Fy_maxFtot_bw, force_clean, expression(max~Fy~"/"~bw~(N)))
plot_maxFy

plot_maxFy_delta <- plot_del_var(Fy_maxFtot_bw, force_clean, expression(max~Fy~"/"~bw~(N)))
plot_maxFy_delta

plot_maxFy_base <- plot_raw_base(Fy_maxFtot_bw, force_clean, expression(max~Fy~"/"~bw~(N)))
plot_maxFy_base
```

#### d. vertical force (Fz)
##### absolute
```{r}
plot_maxFz_abs <- plot_age_var(Fz_maxFtot, force_clean, expression(max~Fz~(N)))
plot_maxFz_abs

plot_maxFz_abs_delta <- plot_del_var(Fz_maxFtot, force_clean, expression(max~Fz~(N)))
plot_maxFz_abs_delta

plot_maxFz_abs_base <- plot_raw_base(Fz_maxFtot, force_clean, expression(max~Fz~(N)))
plot_maxFz_abs_base
```

##### relative
```{r}
plot_maxFz <- plot_age_var(Fz_maxFtot_bw, force_clean, expression(max~Fz~"/"~bw~(N)))
plot_maxFz

plot_maxFz_delta <- plot_del_var(Fz_maxFtot_bw, force_clean, expression(max~Fz~"/"~bw~(N)))
plot_maxFz_delta

plot_maxFz_base <- plot_raw_base(Fz_maxFtot_bw, force_clean, expression(max~Fz~"/"~bw~(N)))
plot_maxFz_base


```
#### e. longitudinal force
##### absolute
```{r}
plot_maxFx_abs <- plot_age_var(Fx_maxFtot, force_clean, expression(max~Fx~(N)))
plot_maxFx_abs

plot_maxFx_abs_delta <- plot_del_var(Fx_maxFtot, force_clean, expression(max~Fx~(N)))
plot_maxFx_abs_delta

plot_maxFx_abs_base <- plot_raw_base(Fx_maxFtot, force_clean, expression(max~Fx~(N)))
plot_maxFx_abs_base
```
##### relative
```{r}
plot_maxFx <- plot_age_var(Fx_maxFtot_bw, force_clean, expression(max~Fx~"/"~bw~(N)))
plot_maxFx

plot_maxFx_delta <- plot_del_var(Fx_maxFtot_bw, force_clean, expression(max~Fx~"/"~bw~(N)))
plot_maxFx_delta

plot_maxFx_base <- plot_raw_base(Fx_maxFtot_bw, force_clean, expression(max~Fx~"/"~bw~(N)))
plot_maxFx_base
```
#### f. takeoff angle 
```{r}
plot_angle <- plot_age_var(angle, force_clean, expression(take-off~angle~("°")))
plot_angle

plot_angle_delta <- plot_del_var(angle, force_clean, expression(take-off~angle~("°")))
plot_angle_delta

plot_angle_base <- plot_raw_base(angle, force_clean, expression(take-off~angle~("°")))
plot_angle_base
```


### LEG IMPULSE
##### impulse 
```{r}

plot_impulse <- plot_age_var(impulse, force_clean, expression(impulse~(mN*s)))
plot_impulse

plot_impulse_delta <- plot_del_var(impulse, force_clean, expression(impulse~(mN*s)))
plot_impulse_delta

plot_impulse_base <- plot_raw_base(impulse, force_clean, expression(impulse~(mN*s)))
plot_impulse_base

```

##### duration
```{r}
plot_duration <- plot_age_var(duration, force_clean, expression(duration~(ms)))
plot_duration

plot_duration_delta <- plot_del_var(duration, force_clean, expression(duration~(ms)))
plot_duration_delta

plot_duration_base <- plot_raw_base(duration, force_clean, expression(duration~(ms)))
plot_duration_base
```

##### impulse ratio
```{r}
plot_impulse_ratio <- plot_age_var(impulse_ratio, force_clean, expression(impulse~ratio))
plot_impulse_ratio

plot_impulse_ratio_delta <- plot_del_var(impulse_ratio, force_clean, expression(impulse~ratio))
plot_impulse_ratio_delta

plot_impulse_ratio_base <- plot_raw_base(impulse_ratio, force_clean, expression(impulse~ratio))
plot_impulse_ratio_base
```


##### left impulse
```{r}
plot_left_impulse <- plot_age_var(left_impulse*1000, force_clean, expression(left~impulse~(ms))) 
plot_left_impulse

plot_left_impulse_delta <- plot_del_var(left_impulse*1000, force_clean, expression(left~impulse~(ms))) 
plot_left_impulse_delta

plot_left_impulse_base <- plot_raw_base(left_impulse*1000, force_clean, expression(left~impulse~(ms)))
plot_left_impulse_base
```

##### right impulse
```{r}
plot_right_impulse <- plot_age_var(right_impulse*1000, force_clean, expression(right~impulse~(ms))) 
plot_right_impulse

plot_right_impulse_delta <- plot_del_var(right_impulse*1000, force_clean, expression(right~impulse~(ms))) 
plot_right_impulse_delta

plot_right_impulse_base <- plot_raw_base(right_impulse*1000, force_clean, expression(right~impulse~(ms)))
plot_right_impulse_base
```

##### left duration
```{r}
plot_left_duration <- plot_age_var(left_duration*1000, force_clean, expression(left~duration~(ms)))
plot_left_duration

plot_left_duration_delta <- plot_del_var(left_duration*1000, force_clean, expression(left~duration~(ms))) 
plot_left_duration_delta

plot_left_duration_base <- plot_raw_base(left_duration*1000, force_clean, expression(left~duration~(ms)))
plot_left_duration_base
```

##### right duration
```{r}
plot_right_duration <- plot_age_var(right_duration*1000, force_clean, expression(right~duration~(ms)))
plot_right_duration

plot_right_duration_delta <- plot_del_var(right_duration*1000, force_clean, expression(right~duration~(ms))) 
plot_right_duration_delta

plot_left_duration_base <- plot_raw_base(right_duration*1000, force_clean, expression(right~duration~(ms)))
plot_left_duration_base
```

### TORTUOSITY
#### turn rate
##### mean
```{r}
plot_tr <- plot_age_var(mean_turn_rate, flights_tort, expression(turn~rate~"(°/s)"))
plot_tr

plot_tr_delta <- plot_del_var(mean_turn_rate, flights_tort, expression(turn~rate~"(°/s)"))
plot_tr_delta

plot_tr_base <- plot_raw_base(mean_turn_rate, flights_tort, expression(turn~rate~"(°/s)"))
plot_tr_base

```

##### max
```{r}

plot_tr_max <- plot_age_var(max_turn_rate, flights_tort, "max turn rate (°/s)") +ylim(-1250, 5000)
plot_tr_max
```

#### centripetal acceleration
##### mean
```{r}
ytitle <- expression(a[centripetal]~(m/s^2))


plot_ctracc <- plot_age_var(mean_centri_acc, flights_tort, ytitle) #+ylim(0, 30)
plot_ctracc

plot_ctracc_delta <- plot_del_var(mean_centri_acc, flights_tort, ytitle) #+ylim(0, 30)
plot_ctracc_delta

plot_ctracc_base <- plot_raw_base(mean_centri_acc, flights_tort, ytitle)
plot_ctracc_base

```

##### max
```{r}

plot_ctracc_max <- plot_age_var(max_centri_acc, flights, "max centripetal acceleration (m/s^2)") +ylim(-50, 150)
plot_ctracc_max

```
### FLIGHT SPEED
#### velocity magnitude
##### mean
```{r}
plot_vres <- plot_age_var(mean_v_res, flights_speed, "|v| (m/s)")
plot_vres

plot_vres_delta <- plot_del_var(mean_v_res, flights_speed, "|v| (m/s)")
plot_vres_delta

plot_vres_base <- plot_raw_base(mean_v_res, flights_speed, "|v| (m/s)")
plot_vres_base

```
##### max
```{r}
plot_vres_max <- plot_age_var(variable = max_v_res, df = flights, title = "|v| (m/s)")
plot_vres_max

```

#### tangential acceleration
##### mean
```{r}
ytitle <- expression(a[tan]~(m/s^2))
plot_tacc <- plot_age_var(mean_acc_tan, flights_speed, ytitle)
plot_tacc

plot_tacc_delta <- plot_del_var(mean_acc_tan, flights_speed, ytitle)
plot_tacc_delta

plot_tacc_base <- plot_raw_base(mean_acc_tan, flights_speed, ytitle)
plot_tacc_base


```
##### max
```{r}
plot_tacc_max <- plot_age_var(max_acc_tan, flights_speed, ytitle) #+ylim(-100, 150)
plot_tacc_max

```
#### X vel 
##### mean 
```{r}
ytitle <- expression(v[x]~(m/s))
plot_vx <- plot_age_var(mean_vx, flights_speed, ytitle)
plot_vx

plot_vx_delta <- plot_del_var(mean_vx, flights_speed, ytitle)
plot_vx_delta 

plot_vx_base <- plot_raw_base(mean_vx, flights_speed, ytitle)
plot_vx_base
```

##### max 
```{r}
plot_vx_max <- plot_age_var(max_vx, flights_speed, "max X velocity (m/s)")
plot_vx_max
```
#### X acc
##### mean
```{r}
ytitle <- expression(acc[x]~(m/s^2))
plot_accx <- plot_age_var(mean_acc_x, flights_speed, ytitle) 
plot_accx

plot_accx_delta <- plot_del_var(mean_acc_x, flights_speed, ytitle) 
plot_accx_delta

plot_accx_base <- plot_raw_base(mean_acc_x, flights_speed, ytitle) 
plot_accx_base

```
##### max 
```{r}
plot_accx_max <- plot_age_var(max_acc_x, flights, "max X acceleration (m/s^2)") 
plot_accx_max

```
#### Y vel 
##### mean 
```{r}
ytitle <- expression(v[y]~(m/s))
plot_vy <- plot_age_var(mean_vy, flights_speed, ytitle) 
plot_vy
plot_vy_delta <- plot_del_var(mean_vy, flights_speed, ytitle) 
plot_vy_delta 
plot_vy_base <- plot_raw_base(mean_vy, flights_speed, ytitle) 
plot_vy_base
```
##### max 
```{r}
plot_vy_max <- plot_age_var(max_vy, flights_speed, "max Y velocity (m/s)")
plot_vy_max #+ylim(0, 6)
```
#### Y acc
##### mean 
```{r}
ytitle <- expression(acc[y]~(m/s^2))
plot_accy <- plot_age_var(mean_acc_y, flights_speed, ytitle) 
plot_accy

plot_accy_delta <- plot_del_var(mean_acc_y, flights_speed, ytitle) 
plot_accy_delta

plot_accy_base <- plot_raw_base(mean_acc_y, flights_speed, ytitle) 
plot_accy_base
```
##### max 
```{r}
plot_accy_max <- plot_age_var(mean_acc_y, flights, "max Y acceleration (m/s^2)") 
plot_accy_max
```
#### Z vel 
##### mean 
```{r}
ytitle <- expression(v[z]~(m/s))
plot_vz <- plot_age_var(mean_vz, flights_speed, ytitle)
plot_vz

plot_vz_delta <- plot_del_var(mean_vz, flights_speed, ytitle)
plot_vz_delta 
plot_vz_base <- plot_raw_base(mean_vz, flights_speed, ytitle)
plot_vz_base
```
##### max 
```{r}
plot_vz_max <- plot_age_var(max_vz, flights, "max Z velocity (m/s)")
plot_vz_max

```
#### Z acc
##### mean
```{r}
ytitle <- expression(acc[z]~(m/s^2))
plot_accz <- plot_age_var(mean_acc_z, flights_speed, ytitle)
plot_accz

plot_accz_delta <- plot_del_var(mean_acc_z, flights_speed, ytitle)
plot_accz_delta

plot_accz_base <- plot_raw_base(mean_acc_z, flights_speed, ytitle)
plot_accz_base

```
##### max 
```{r}
plot_accz_max <- plot_age_var(max_acc_z, flights, "max Z acceleration (m/s^2)") 
plot_accz_max
```

### FLIGHT STABILITY
#### wingbeat frequency
```{r}
plot_wb <- plot_age_var(wingbeat_frequency, flights_wing,  expression(wingbeat~frequency~(Hz)))
plot_wb

plot_wb_delta <- plot_del_var(wingbeat_frequency, flights_wing, expression(Delta~wingbeat~frequency~(Hz)))
plot_wb_delta

plot_wb_base <- plot_raw_base(wingbeat_frequency, flights_wing, expression(wingbeat~frequency~(Hz)))
plot_wb_base
```


#### upstroke
##### acc
```{r}
plot_upstroke_acc <- plot_age_var(upstroke_amplitude_acc, flights, expression(upstroke~acceleration~(m/s^2)))
plot_upstroke_acc
```
##### mm
```{r}
plot_upstroke_mm <- plot_age_var(1000*upstroke_amplitude_m, flights_wing, expression(upstroke~head~displacement~(mm)))
plot_upstroke_mm

plot_upstroke_mm_delta <- plot_del_var(1000*upstroke_amplitude_m, flights_wing, expression(Delta~upstroke~head~displacement~(mm)))
plot_upstroke_mm_delta

plot_upstroke_mm_base <- plot_raw_base(1000*upstroke_amplitude_m, flights_wing, expression(upstroke~head~displacement~(mm)))
plot_upstroke_mm_base
```
#### downstroke
##### acc
```{r}
plot_downstroke_acc <- plot_age_var(downstroke_amplitude_acc, flights, expression(downstroke~head~displacement~(mm)))
plot_downstroke_acc
```

```{r}

plot_downstroke_mm <- plot_age_var(1000*downstroke_amplitude_m, flights_wing,  expression(downstroke~head~displacement~(mm)))
plot_downstroke_mm


plot_downstroke_mm_delta <- plot_del_var(1000*downstroke_amplitude_m, flights_wing, expression(Delta~downstroke~head~displacement~(mm)))
plot_downstroke_mm_delta

plot_downstroke_mm_base <- plot_raw_base(1000*downstroke_amplitude_m, flights_wing, expression(downstroke~head~displacement~(mm)))
plot_downstroke_mm_base
```

# II. Model analysis 
## function detect_breakpoint()
```{r}

detect_breakpoint <- function(variable, df){
  
# Capture the symbol name passed (e.g., wing_loading)
  response <- deparse(substitute(variable))
  
  # Build a formula: response ~ Age*Sex + Brood
  form <- reformulate(termlabels = c("Age*Sex", "Brood"), response = response)
  
  # Fit linear model
  m_lin <- lm(form, data = df, na.action = na.omit)
  
  # Fit segmented model
  m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 28))  # initial guess at 36
  #summary(m_seg) # Age not significant. Larger brood, larger wing area
  plot(m_seg)  # quick diagnostic plot
  
  # Extract estimated breakpoint and slopes
  breakpoint <- m_seg$psi["psi1.Age", "Est."] #27.8
  breakpoint_se <- m_seg$psi["psi1.Age", "St.Err"]
  slope_early <- slope(m_seg)$Age[1, "Est."]
  slope_late  <- slope(m_seg)$Age[2, "Est."]
  model_summary <- summary(m_seg)
  
  # return both the model object and the numeric estimates
  out <- list(
    model_summary = model_summary,
    estimates = c(
      breakpoint = breakpoint,
      breakpoint_se = breakpoint_se,
      slope_early = slope_early, 
      slope_late = slope_late))
  
    return(out)
}

```

## function compare_models_aic()
```{r}
# Function to compare models using AIC, ΔAIC, and Akaike weights
compare_models_aic <- function(models) {
  # Validate input
  if (!is.list(models) || length(models) < 2) {
    stop("Please provide a list of at least two fitted model objects.")
  }
  
  # Extract model names
  model_names <- names(models)
  if (is.null(model_names) || any(model_names == "")) {
    model_names <- paste0("Model_", seq_along(models))
  }
  
  # Compute AIC for each model
  aic_values <- sapply(models, function(m) {
    tryCatch(AIC(m), error = function(e) NA)
  })
  
  # Check for NA values
  if (any(is.na(aic_values))) {
    stop("One or more models could not have their AIC computed.")
  }
  
  # Calculate ΔAIC
  delta_aic <- aic_values - min(aic_values)
  
  # Calculate Akaike weights
  rel_likelihood <- exp(-0.5 * delta_aic)
  akaike_weights <- rel_likelihood / sum(rel_likelihood)
  
  # Create results data frame
  results <- data.frame(
    Model = model_names,
    AIC = round(aic_values, 3),
    Delta_AIC = round(delta_aic, 3),
    Akaike_Weight = round(akaike_weights, 4),
    stringsAsFactors = FALSE
  )
  
  # Order by AIC
  results <- results[order(results$AIC), ]
  rownames(results) <- NULL
  
  return(results)
}
```

## function compare_models_pmin()
```{r}

compare_models_pmin <- function(variable, breakpoint, df){
  
  # 1) Hinge coding: 0 before breakpoint, positive after
  # using pmin() as hinge control with pmax() highly correlates with Age. 
  df$hinge <- pmin(df$Age - breakpoint, 0)

  # 2) Capture response name for formulas
  response <- deparse(substitute(variable))

  # Helpers
  form_lm   <- function(terms) reformulate(terms, response = response)
  # For lmer, we must build the full formula string because of random-effects syntax
  form_lmer <- function(rhs) as.formula(paste(response, "~", rhs))

  ## ----- Linear models (fixed effects) -----
  m1 <- lm(form_lm(c("Age", "Sex", "hinge", "Brood", "Bird")), data = df)
  m2 <- lm(form_lm(c("Age", "Sex", "hinge", "Bird")), data = df)
  m3 <- lm(form_lm(c("Age*Sex", "hinge", "Bird")), data = df)
  m4 <- lm(form_lm(c("Age*Sex", "hinge", "Brood", "Bird")), data = df)

  ## ----- Linear mixed models (random effects) -----
  # NOTE: If comparing mixed models by AIC across different fixed-effect structures,
  # consider REML = FALSE; uncomment the REML = FALSE versions if your
  # compare_models_aic relies on ML fits.
  m5  <- lme4::lmer(form_lmer("Age + Sex + Brood + hinge + (1|Bird)"), data = df)         # , REML = FALSE
  m6  <- lme4::lmer(form_lmer("Age + Sex + hinge + (1|Bird)"), data = df,)                  # , REML = FALSE
  m7  <- lme4::lmer(form_lmer("Age*Sex + Brood + hinge + (1|Bird)"), data = df)            # , REML = FALSE
  m8  <- lme4::lmer(form_lmer("Age*Sex + hinge + (1|Bird)"), data = df)                    # , REML = FALSE
  m9  <- lme4::lmer(form_lmer("Age*Sex + hinge + (1|Brood/Bird)"), data = df)              # , REML = FALSE
  m10 <- lme4::lmer(form_lmer("Age + Sex + hinge + (1|Brood/Bird)"), data = df)            # , REML = FALSE

  ## ----- Compare models -----
  all_models <- compare_models_aic(list(
    "m1-lm no interaction"                  = m1,
    "m2-lm no interaction, no brood"        = m2,
    "m3-lm interaction, no brood"           = m3,
    "m4-lm interaction"                     = m4,
    "m5-(1|Bird), no interaction"           = m5,
    "m6-(1|Bird), no interaction, no brood" = m6,
    "m7-(1|Bird), interaction"              = m7,
    "m8-(1|Bird), interaction, no brood"    = m8,
    "m9-(1|Brood/Bird), interaction"        = m9,
    "m10-(1|Brood/Bird), no interaction"    = m10
  ))

  just_lmer <- compare_models_aic(list(
    "m5-(1|Bird), no interaction"           = m5,
    "m6-(1|Bird), no interaction, no brood" = m6,
    "m7-(1|Bird), interaction"              = m7,
    "m8-(1|Bird), interaction, no brood"    = m8,
    "m9-(1|Brood/Bird), interaction"        = m9,
    "m10-(1|Brood/Bird), no interaction"    = m10
  ))
  
  bonus <- performance::compare_performance(m1, m2, m3, m4, m5, m6, m7, m8, m9, m10, rank = TRUE)
  
  # Return comparisons AND the fitted models
  fits <- list(m1=m1, m2=m2, m3=m3, m4=m4, m5=m5, m6=m6, m7=m7, m8=m8, m9=m9, m10=m10)


  return(list(all_models = all_models, just_lmer = just_lmer, bonus = bonus, fits))
}



```
## function compare_models_pmax()
```{r}

compare_models_pmax <- function(variable, breakpoint, df){
  
  # 1) Hinge coding: 0 before breakpoint, positive after
  # using pmin() as hinge control with pmax() highly correlates with Age. 
  df$hinge <- pmax(df$Age - breakpoint, 0)

  # 2) Capture response name for formulas
  response <- deparse(substitute(variable))

  # Helpers
  form_lm   <- function(terms) reformulate(terms, response = response)
  # For lmer, we must build the full formula string because of random-effects syntax
  form_lmer <- function(rhs) as.formula(paste(response, "~", rhs))

  ## ----- Linear models (fixed effects) -----
  m1 <- lm(form_lm(c("Age", "Sex", "hinge", "Brood", "Bird")), data = df)
  m2 <- lm(form_lm(c("Age", "Sex", "hinge", "Bird")), data = df)
  m3 <- lm(form_lm(c("Age*Sex", "hinge", "Bird")), data = df)
  m4 <- lm(form_lm(c("Age*Sex", "hinge", "Brood", "Bird")), data = df)

  ## ----- Linear mixed models (random effects) -----
  # NOTE: If comparing mixed models by AIC across different fixed-effect structures,
  # consider REML = FALSE; uncomment the REML = FALSE versions if your
  # compare_models_aic relies on ML fits.
  m5  <- lme4::lmer(form_lmer("Age + Sex + Brood + hinge + (1|Bird)"), data = df)         # , REML = FALSE
  m6  <- lme4::lmer(form_lmer("Age + Sex + hinge + (1|Bird)"), data = df,)                  # , REML = FALSE
  m7  <- lme4::lmer(form_lmer("Age*Sex + Brood + hinge + (1|Bird)"), data = df)            # , REML = FALSE
  m8  <- lme4::lmer(form_lmer("Age*Sex + hinge + (1|Bird)"), data = df)                    # , REML = FALSE
  m9  <- lme4::lmer(form_lmer("Age*Sex + hinge + (1|Brood/Bird)"), data = df)              # , REML = FALSE
  m10 <- lme4::lmer(form_lmer("Age + Sex + hinge + (1|Brood/Bird)"), data = df)            # , REML = FALSE

  ## ----- Compare models -----
  all_models <- compare_models_aic(list(
    "m1-lm no interaction"                  = m1,
    "m2-lm no interaction, no brood"        = m2,
    "m3-lm interaction, no brood"           = m3,
    "m4-lm interaction"                     = m4,
    "m5-(1|Bird), no interaction"           = m5,
    "m6-(1|Bird), no interaction, no brood" = m6,
    "m7-(1|Bird), interaction"              = m7,
    "m8-(1|Bird), interaction, no brood"    = m8,
    "m9-(1|Brood/Bird), interaction"        = m9,
    "m10-(1|Brood/Bird), no interaction"    = m10
  ))

  just_lmer <- compare_models_aic(list(
    "m5-(1|Bird), no interaction"           = m5,
    "m6-(1|Bird), no interaction, no brood" = m6,
    "m7-(1|Bird), interaction"              = m7,
    "m8-(1|Bird), interaction, no brood"    = m8,
    "m9-(1|Brood/Bird), interaction"        = m9,
    "m10-(1|Brood/Bird), no interaction"    = m10
  ))
  
  bonus <- performance::compare_performance(m1, m2, m3, m4, m5, m6, m7, m8, m9, m10, rank = TRUE)
  
  # Return comparisons AND the fitted models
  fits <- list(m1=m1, m2=m2, m3=m3, m4=m4, m5=m5, m6=m6, m7=m7, m8=m8, m9=m9, m10=m10)


  return(list(all_models = all_models, just_lmer = just_lmer, bonus = bonus, fits))
}




```
## function check_assumptions()
```{r}

check_assumptions <- function(model){
  
  # check collinearity
  check_collinearity(model)
  
  # Computes the default residuals (raw conditional)
  raw_cond_model <- redres::compute_redres(model)
  
  hist(raw_cond_model, main = "Normality of residuals") #normality in histogram
  p_hetero <- plot_redres(model) #heteroskedasticity
  p_resqq <- plot_resqq(model) #residual QQ
  p_ranef <- plot_ranef(model) #random effects QQ / density 
  
  print(p_hetero)
  print(p_resqq)
  print(p_ranef)
  
  invisible(NULL)
}
```
## function plot_breakpoint
```{r}

plot_breakpoint <- function(breakpoint, breakpoint_se, base_plot){
  
  yr <- ggplot_build(base_plot)$layout$panel_params[[1]]$y.range
  y_top <- max(yr)
  
  new_plot <- base_plot + 
    ggplot2::geom_vline(xintercept = breakpoint, linetype = "dashed", color = "red") +
    ggplot2::annotate(
        "segment",
        x = breakpoint - breakpoint_se,
        xend = breakpoint + breakpoint_se,
        y    = y_top,
        yend = y_top,
        color = "#E2AFDE",
        linewidth = 3,
        alpha = 0.7) +
    ggplot2::labs(subtitle = sprintf("breakpoint = %.1f (± %.1f SE)", breakpoint, breakpoint_se))
  
  return(new_plot)
  
}


```


## function plot_final
```{r}

plot_final <- function(breakpoint, breakpoint_se,
                           model,            # lmer model
                           df,               # data used for model
                           overlay = FALSE,
                           base_plot,        # ggplot object
                           include_brood = TRUE,
                           brood_level = "Box5",
                           sex = "M"
                           ) {
  df <- as.data.frame(df)
  # 1) Hinge coding
  df$hinge <- pmin(df$Age - breakpoint, 0)

  # 2) Age sequence
  age_seq <- seq(min(df$Age, na.rm = TRUE), max(df$Age, na.rm = TRUE), by = 1)

  # 3) Ensure factors
  if ("Sex" %in% names(df) && !is.factor(df$Sex)) df$Sex <- factor(df$Sex)
  if ("Brood" %in% names(df) && !is.factor(df$Brood)) df$Brood <- factor(df$Brood)

  # 4) Base newdat
  newdat_out <- data.frame(
    Age = age_seq,
    Sex = factor(sex, levels = levels(df$Sex)),
    hinge = pmin(age_seq - breakpoint, 0)
  )

  # 5) Add Brood if requested
  if (include_brood) {
    if (!("Brood" %in% names(df))) {
      stop("'Brood' column not found in df.")
    }
    if (!(brood_level %in% levels(df$Brood))) {
      warning(sprintf("brood_level='%s' not found; using reference '%s'.",
                      brood_level, levels(df$Brood)[1]))
      brood_level <- levels(df$Brood)[1]
    }
    newdat_out$Brood <- factor(brood_level, levels = levels(df$Brood))
  }

  # 6) Predict (population-level)
  newdat_out$pred <- predict(model, newdat_out, re.form = NA)

  # 7) Compose plot
  yr <- ggplot_build(base_plot)$layout$panel_params[[1]]$y.range
  y_top <- max(yr)
  
  
  if(overlay == TRUE){

  p <- base_plot +
    
    ggplot2::geom_line(
      data = newdat_out,
      ggplot2::aes(x = Age, y = pred),
      color = "#1f77b4", size = 1.5, inherit.aes = FALSE
    ) +
    ggplot2::geom_vline(xintercept = breakpoint, linetype = "dashed", color = "red") +
    ggplot2::annotate(
        "segment",
        x = breakpoint - breakpoint_se,
        xend = breakpoint + breakpoint_se,
        y    = y_top,
        yend = y_top,
        color = "#E2AFDE",
        linewidth = 3,
        alpha = 0.7) +
    ggplot2::labs(subtitle = sprintf("breakpoint = %.1f (± %.1f SE)", breakpoint, breakpoint_se))
  }
  else{
    
  p <- ggplot2::geom_line( # predicted means
      data = newdat_out,
      ggplot2::aes(x = Age, y = pred),
      color = "#1f77b4", size = 1.5, inherit.aes = FALSE
    ) +
      geom_ribbon( # edit this part to overlay 95% CI (+-1 SD) for predicted means
      data = newdat_out,
      aes(x = Age, ymin = mean - sd, ymax = mean + sd),
      fill = "grey70", alpha = 0.5,
      inherit.aes = FALSE) + 
    # ggplot2 CI levels 
    
    ggplot2::geom_vline(xintercept = breakpoint, linetype = "dashed", color = "red") +
    ggplot2::annotate(
        "segment",
        x = breakpoint - breakpoint_se,
        xend = breakpoint + breakpoint_se,
        y    = y_top,
        yend = y_top,
        color = "#E2AFDE",
        linewidth = 3,
        alpha = 0.7) +
    ggplot2::labs(subtitle = sprintf("breakpoint = %.1f (± %.1f SE)", breakpoint, breakpoint_se))  
    
    
    
  }

  return(list(plot = p, newdat = newdat_out))
}


```
## function plot_final2
- for functions that don't show breakpoint SEs
- the only difference is y_top <- max(yr)*0.95
```{r}

plot_final2 <- function(breakpoint, breakpoint_se,
                           model,            # lmer model
                           df,               # data used for model
                           base_plot,        # ggplot object
                           include_brood = TRUE,
                           brood_level = "Box5",
                           sex = "M"
                           ) {
  df <- as.data.frame(df)
  # 1) Hinge coding
  df$hinge <- pmin(df$Age - breakpoint, 0)

  # 2) Age sequence
  age_seq <- seq(min(df$Age, na.rm = TRUE), max(df$Age, na.rm = TRUE), by = 1)

  # 3) Ensure factors
  if ("Sex" %in% names(df) && !is.factor(df$Sex)) df$Sex <- factor(df$Sex)
  if ("Brood" %in% names(df) && !is.factor(df$Brood)) df$Brood <- factor(df$Brood)

  # 4) Base newdat
  newdat_out <- data.frame(
    Age = age_seq,
    Sex = factor(sex, levels = levels(df$Sex)),
    hinge = pmin(age_seq - breakpoint, 0)
  )

  # 5) Add Brood if requested
  if (include_brood) {
    if (!("Brood" %in% names(df))) {
      stop("'Brood' column not found in df.")
    }
    if (!(brood_level %in% levels(df$Brood))) {
      warning(sprintf("brood_level='%s' not found; using reference '%s'.",
                      brood_level, levels(df$Brood)[1]))
      brood_level <- levels(df$Brood)[1]
    }
    newdat_out$Brood <- factor(brood_level, levels = levels(df$Brood))
  }

  # 6) Predict (population-level)
  newdat_out$pred <- predict(model, newdat_out, re.form = NA)

  # 7) Compose plot
  yr <- ggplot_build(base_plot)$layout$panel_params[[1]]$y.range
  y_top <- max(yr)*0.95

  p <- base_plot +
    
    ggplot2::geom_line(
      data = newdat_out,
      ggplot2::aes(x = Age, y = pred),
      color = "#1f77b4", size = 1.5, inherit.aes = FALSE
    ) +
    ggplot2::geom_vline(xintercept = breakpoint, linetype = "dashed", color = "red") +
    ggplot2::annotate(
        "segment",
        x = breakpoint - breakpoint_se,
        xend = breakpoint + breakpoint_se,
        y    = y_top,
        yend = y_top,
        color = "#E2AFDE",
        linewidth = 3,
        alpha = 0.7) +
    ggplot2::labs(subtitle = sprintf("breakpoint = %.1f (± %.1f SE)", breakpoint, breakpoint_se))
  

  return(list(plot = p, newdat = newdat_out))
}


```
## function plot_final3
- for functions that don't show breakpoint SEs
- the only difference is y_top <- max(yr)*0.90
```{r}

plot_final3 <- function(breakpoint, breakpoint_se,
                           model,            # lmer model
                           df,               # data used for model
                           base_plot,        # ggplot object
                           include_brood = TRUE,
                           brood_level = "Box5",
                           sex = "M"
                           ) {
  df <- as.data.frame(df)
  # 1) Hinge coding
  df$hinge <- pmin(df$Age - breakpoint, 0)

  # 2) Age sequence
  age_seq <- seq(min(df$Age, na.rm = TRUE), max(df$Age, na.rm = TRUE), by = 1)

  # 3) Ensure factors
  if ("Sex" %in% names(df) && !is.factor(df$Sex)) df$Sex <- factor(df$Sex)
  if ("Brood" %in% names(df) && !is.factor(df$Brood)) df$Brood <- factor(df$Brood)

  # 4) Base newdat
  newdat_out <- data.frame(
    Age = age_seq,
    Sex = factor(sex, levels = levels(df$Sex)),
    hinge = pmin(age_seq - breakpoint, 0)
  )

  # 5) Add Brood if requested
  if (include_brood) {
    if (!("Brood" %in% names(df))) {
      stop("'Brood' column not found in df.")
    }
    if (!(brood_level %in% levels(df$Brood))) {
      warning(sprintf("brood_level='%s' not found; using reference '%s'.",
                      brood_level, levels(df$Brood)[1]))
      brood_level <- levels(df$Brood)[1]
    }
    newdat_out$Brood <- factor(brood_level, levels = levels(df$Brood))
  }

  # 6) Predict (population-level)
  newdat_out$pred <- predict(model, newdat_out, re.form = NA)

  # 7) Compose plot
  yr <- ggplot_build(base_plot)$layout$panel_params[[1]]$y.range
  y_top <- max(yr)*0.94

  p <- base_plot +
    
    ggplot2::geom_line(
      data = newdat_out,
      ggplot2::aes(x = Age, y = pred),
      color = "#1f77b4", size = 1.5, inherit.aes = FALSE
    ) +
    ggplot2::geom_vline(xintercept = breakpoint, linetype = "dashed", color = "red") +
    ggplot2::annotate(
        "segment",
        x = breakpoint - breakpoint_se,
        xend = breakpoint + breakpoint_se,
        y    = y_top,
        yend = y_top,
        color = "#E2AFDE",
        linewidth = 3,
        alpha = 0.7) +
    ggplot2::labs(subtitle = sprintf("breakpoint = %.1f (± %.1f SE)", breakpoint, breakpoint_se))
  

  return(list(plot = p, newdat = newdat_out))
}


```
## A. Significant
### MORPHOLOGY
#### bodyweight - 28.1
##### breakpoint detection
```{r}
out <- detect_breakpoint(scale_weight, wing_data_final)
print(out)
breakpoint_sw <- round(out$estimates["breakpoint"], 1)
breakpoint_se_sw <- round(out$estimates["breakpoint_se"], 1)

```
```{r}
plot_mass_raw <- plot_breakpoint(breakpoint_sw, breakpoint_se_sw, plot_scale_g)

plot_mass_delta <- plot_breakpoint(breakpoint_sw, breakpoint_se_sw, plot_scale_g_delta)

plot_mass_raw
plot_mass_delta
```
##### aic comparison
```{r}
wing_data_final$scale_g <- wing_data_final$scale_weight*101.97

model_results <- compare_models_pmin(scale_g, breakpoint_sw, wing_data_final)

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_sw <- model_results[[4]]$m5
m_sw
```

##### model assumption checks
```{r}
check_collinearity(m_sw) #checking collinearity for the best linear model

# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(m_sw)
  
hist(raw_cond_model, main = "Normality of residuals") #normality in histogram
plot_redres(m_sw) #heteroskedasticity
plot_resqq(m_sw) #residual QQ
plot_ranef(m_sw)
```

##### test
```{r}
summary(m_sw)
tab_model(m_sw)
anova_sw <- Anova(m_sw, type =3, test.statistic = "F") %>% as.data.frame() %>%
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

# #before slope = 0?
# linearHypothesis(m_sw, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
# 
# #after slope = 0?
# linearHypothesis(m_sw, "Age = 0", test = "F")  # Is the slope after the breakpoint flat? -> no
# 
# #same slope?
# linearHypothesis(m_sw, "Age = Age+hinge", test = "F") #Is the slope before and after the same? -> n
# 
# #before breakpoint slope estimate
# deltaMethod(m_sw, "Age + hinge") 
# 
# #after breakpoint slope estimate
# deltaMethod(m_sw, "Age") 



```

##### plot
```{r}

# output <- plot_final(breakpoint_sw, breakpoint_se_sw, 
#                      m_sw, 
#                      wing_data_final,
#                      plot_scale_g,
#                      include_brood = TRUE,
#                      brood_level = "Box5",
#                      sex = "M"
#                      )
# plot_scale_g_final <- output[1]
# plot_scale_g_final

```


#### wingspan - 28
##### breakpoint detection
```{r}
out <- detect_breakpoint(wingspan, wing_data_final)
print(out)
breakpoint_ws <- round(out$estimates["breakpoint"], 1)
breakpoint_se_ws <- round(out$estimates["breakpoint_se"], 1)
```
```{r}

plot_wingspan_raw <- plot_breakpoint(breakpoint_ws, breakpoint_se_ws, plot_wingspan)

plot_wingspan_delta <- plot_breakpoint(breakpoint_ws, breakpoint_se_ws, plot_wingspan_delta)

plot_wingspan_raw
plot_wingspan_delta

```
##### aic comparison
```{r}

model_results <- compare_models_pmin(wingspan, breakpoint_ws, wing_data_final)

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_wingspan <- model_results[[4]]$m5
m_wingspan


```

##### model assumption check
```{r}
check_collinearity(m_wingspan)
  
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(m_wingspan)
  
hist(raw_cond_model, main = "Normality of residuals") #normality in histogram
plot_redres(m_wingspan) #heteroskedasticity
plot_resqq(m_wingspan) #residual QQ
plot_ranef(m_wingspan)
```
##### test
```{r}
# for residual variance / R^2
tab_model(m_wingspan)


# Wald F test with KR approximation & FDR
anova_wingspan <- Anova(m_wingspan, type =3, test.statistic = "F") %>% as.data.frame() %>%
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_wingspan

# # Hypothesis testing
# linearHypothesis(m_wingspan, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
# linearHypothesis(m_wingspan, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
# linearHypothesis(m_wingspan, "Age = Age+hinge", test = "F") #Is the slope before and after the same? -> n
# 
# # slope estimates, SE, CI 
# 
# #before slope
# print(deltaMethod(m_wingspan, "Age + hinge"))
# #after slope
# print(deltaMethod(m_wingspan, "Age"))


```
##### plot
```{r}

# output <- plot_final(breakpoint_ws, breakpoint_se_ws, 
#                      m_wingspan, 
#                      wing_data_final,
#                      plot_wingspan,
#                      include_brood = TRUE,
#                      brood_level = "Box5",
#                      sex = "M"
#                      )
# plot_wingspan_final <- output[1]
# plot_wingspan_final
# 
# wingspan_newdat_out <- output[2]

```

#### wing area - 28.5
##### breakpoint detection
```{r}
out <- detect_breakpoint(wingarea, wing_data_final)
print(out)
breakpoint_wa <- round(out$estimates["breakpoint"], 1)
breakpoint_se_wa <- round(out$estimates["breakpoint_se"], 1)
```
```{r}

plot_wingarea_raw <- plot_breakpoint(breakpoint_wa, breakpoint_se_wa, plot_wingarea)

plot_wingarea_delta <- plot_breakpoint(breakpoint_wa, breakpoint_se_wa, plot_wingarea_delta)

plot_wingarea_raw
plot_wingarea_delta


```
##### aic comparison
```{r}

model_results <- compare_models_pmin(wingarea, breakpoint_wa, wing_data_final)

# shows the results of model comparison
model_results[1:3]
```
```{r}
# saves the model to test 
m_wingarea <- model_results[[4]]$m6
m_wingarea


```

##### model assumption check
```{r}

check_collinearity(m_wingarea)
  
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(m_wingarea)
  
hist(raw_cond_model, main = "Normality of residuals") #normality in histogram
plot_redres(m_wingarea) #heteroskedasticity
plot_resqq(m_wingarea) #residual QQ
plot_ranef(m_wingarea)
```
##### test
```{r}


# for residual variance / R^2
tab_model(m_wingarea)

# Wald F test with KR approximation & FDR
anova_wingarea <- Anova(m_wingarea, type =3, test.statistic = "F") %>% as.data.frame() %>%
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_wingarea

# # Hypothesis testing
# linearHypothesis(m_wingarea, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
# linearHypothesis(m_wingarea, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
# linearHypothesis(m_wingspan, "Age = Age+hinge", test = "F")
# 
# # slope estimates, SE, CI 
# 
# #before slope
# print(deltaMethod(m_wingspan, "Age + hinge"))
# #after slope
# print(deltaMethod(m_wingspan, "Age"))

```
##### plot
```{r}

# output <- plot_final(breakpoint_wa, breakpoint_se_wa, 
#                      m_wingarea, 
#                      wing_data_final,
#                      plot_wingarea,
#                      include_brood = FALSE,
#                      brood_level = "Box5",
#                      sex = "M"
#                      )
# plot_wingarea_final <- output[1]
# plot_wingarea_final

```
#### wing loading
##### breakpoint duration
```{r}

m_lin <- lm(wing_loading ~ Age*Sex + Brood, data = wing_data_final) #CHANGE HERE
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_wl <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
breakpoint_se_wl <- m_seg$psi["psi1.Age", "St.Err"] # breakpoint se.


breakpoint_wl #32.9
breakpoint_se_wl


```
```{r}
plot_wl_raw <- plot_breakpoint(breakpoint_wl, breakpoint_se_wl, plot_wingloading)

plot_wl_delta <- plot_breakpoint(breakpoint_wl, breakpoint_se_wl, plot_wingloading_delta)

plot_wl_raw
plot_wl_delta


```
##### aic comparison
```{r}

model_results <- compare_models_pmax(wing_loading, breakpoint_wl, wing_data_final) #CHANGE

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_wl <- model_results[[4]]$m10 # in theory could
m_wl

```
##### model assumptions
```{r}

model <- m_wl
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)



```
##### test
```{r}

model <- m_wl

tab_model(model)
anova_wl <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_wl # CHANGE THIS!!

# linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
# linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
# linearHypothesis(model, "Age = Age+hinge", test = "F")
# 
# #before slope
# print(deltaMethod(model, "Age + hinge"))
# #after slope
# print(deltaMethod(model, "Age"))

```
##### plot
```{r}

# output <- plot_final(breakpoint_wl, breakpoint_se_wl, #CHANGE HERE
#                      m_wl, # CHANGE HERE
#                      wing_data_final, # CHANGE HERE
#                      plot_wingloading, # CHANGE HERE
#                      include_brood = FALSE,
#                      brood_level = "Box3",
#                      sex = "M"
#                      )
# plot_wl_final <- output[1] #CHANGE HERE
# plot_wl_final # CHANGE HERE


```
### Raw & Delta plots
```{r}
library(patchwork)
morphology_plots <- (plot_mass_raw | plot_wingspan_raw)/ (plot_wingarea_raw | plot_wl_raw)

morphology_delta_plots <- (plot_mass_delta | plot_wingspan_delta)/ (plot_wingarea_delta | plot_wl_delta)

morphology_plots
morphology_delta_plots
```
### LEG


#### max Fz-> x(paper)/bw -40
##### breakpoint detection
```{r}
out <- detect_breakpoint(Fz_maxFtot_bw, force_clean)
print(out)
breakpoint_Fx <- round(out$estimates["breakpoint"], 1)
breakpoint_se_Fx <- round(out$estimates["breakpoint_se"], 1)
```
##### aic comparison
```{r}

model_results <- compare_models_pmin(Fz_maxFtot_bw, breakpoint_Fx, force_clean)

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
lm_Fx <- model_results[[4]]$m1
m_Fx <- model_results[[4]]$m6
#m_Fx_brood <- model_results[[4]]$m10


```

##### model assumption check
```{r}

check_collinearity(lm_Fx)
check_collinearity(m_Fx)



  
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(m_Fx)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(m_Fx) #heteroskedasticity
plot_resqq(m_Fx) #residual QQ
plot_ranef(m_Fx)

```
##### test
```{r}

#before slope
# print(deltaMethod(m_Fx, "Age + hinge"))
# #after slope
# print(deltaMethod(m_Fx, "Age"))
# 
# #hypothesis test
# linearHypothesis(m_Fx, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
# linearHypothesis(m_Fx, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
# linearHypothesis(m_Fx, "Age = Age+hinge", test = "F")

#term test
anova_Fx <- Anova(m_Fx, type =3, test.statistic = "F") %>% as.data.frame() %>%
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_Fx

# Random effects 
# tab_model(m_Fx)



# slope estimates, SE, CI 



```
##### plot
```{r}

# output <- plot_final(breakpoint_Fx, breakpoint_se_Fx, 
#                      m_Fx, 
#                      force_clean,
#                      plot_age_Fx_bw,
#                      include_brood = FALSE,
#                      brood_level = "Box5",
#                      sex = "M"
#                      )
# plot_Fx_final <- output[1]
# plot_Fx_final

```
#### max Fy / bw - 33.8
##### breakpoint detection
```{r}
out <- detect_breakpoint(Fy_maxFtot_bw, force_clean)
print(out)
breakpoint_Fy <- round(out$estimates["breakpoint"], 1)
breakpoint_se_Fy <- round(out$estimates["breakpoint_se"], 1)
```
##### aic comparison
```{r}

model_results <- compare_models_pmin(Fy_maxFtot_bw, breakpoint_Fy, force_clean)

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_Fy <- model_results[[4]]$m6
m_Fy


```

##### model assumption check
```{r}
check_collinearity(m_Fy)
  
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(m_Fy)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(m_Fy) #heteroskedasticity
plot_resqq(m_Fy) #residual QQ
plot_ranef(m_Fy)
```
##### test
```{r}

tab_model(m_Fy)
anova_Fy <- Anova(m_Fy, type =3, test.statistic = "F") %>% as.data.frame() %>%
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_Fy

# linearHypothesis(m_Fy, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
# linearHypothesis(m_Fy, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
# linearHypothesis(m_Fy, "Age = Age+hinge", test = "F")
# 
# #before slope
# print(deltaMethod(m_Fy, "Age + hinge"))
# #after slope
# print(deltaMethod(m_Fy, "Age"))

```
##### plot
```{r}



# output <- plot_final2(breakpoint_Fy, breakpoint_se_Fy, 
#                      m_Fy, 
#                      force_clean,
#                      plot_age_Fy_bw,
#                      include_brood = FALSE,
#                      brood_level = "Box5",
#                      sex = "M"
#                      ) 
# 
# 
# plot_Fy_final <- output[1]
# plot_Fy_final

```
#### max Fz / bw (vertical force, Fx in data)
##### breakpoint detection
```{r}
out <- detect_breakpoint(Fx_maxFtot_bw, force_clean)
print(out)
breakpoint_Fz <- round(out$estimates["breakpoint"], 1)
breakpoint_se_Fz <- round(out$estimates["breakpoint_se"], 1)
```
##### aic comparison
```{r}

model_results <- compare_models_pmax(Fx_maxFtot_bw, breakpoint_Fz, force_clean)

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_Fz <- model_results[[4]]$m10
m_Fz


```

##### model assumption check
```{r}
check_collinearity(m_Fz)
  
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(m_Fz)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(m_Fz) #heteroskedasticity
plot_resqq(m_Fz) #residual QQ
plot_ranef(m_Fz)
```
##### test
```{r}

tab_model(m_Fz)
anova_Fz <- Anova(m_Fz, type =3, test.statistic = "F") %>% as.data.frame() %>%
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_Fz

# linearHypothesis(m_Fz, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
# linearHypothesis(m_Fz, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
# linearHypothesis(m_Fz, "Age = Age+hinge", test = "F")
# 
# #before slope
# print(deltaMethod(m_Fz, "Age + hinge"))
# #after slope
# print(deltaMethod(m_Fz, "Age"))

```
##### plot
```{r}

# output <- plot_final(breakpoint_Fz, breakpoint_se_Fz, 
#                      m_Fz, 
#                      force_clean,
#                      plot_age_Fz_bw,
#                      include_brood = FALSE,
#                      brood_level = "Box5",
#                      sex = "F"
#                      )
# plot_Fz_final <- output[1]
# plot_Fz_final

```
#### max Ftot /bw - 37.1
##### breakpoint detection
```{r}
out <- detect_breakpoint(max_Ftot_bw, force_clean)
print(out)
breakpoint_Ftot <- round(out$estimates["breakpoint"], 1)
breakpoint_se_Ftot <- round(out$estimates["breakpoint_se"], 1)
```
##### aic comparison
```{r}

model_results <- compare_models_pmin(max_Ftot_bw, breakpoint_Ftot, force_clean)

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_Ftot <- model_results[[4]]$m6
m_Ftot


```

##### model assumption check
```{r}
check_collinearity(m_Ftot)
  
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(m_Ftot)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(m_Ftot) #heteroskedasticity
plot_resqq(m_Ftot) #residual QQ
plot_ranef(m_Ftot)
```
##### test
```{r}
model <- m_Ftot

#before slope
# print(deltaMethod(model, "Age + hinge"))
# #after slope
# print(deltaMethod(model, "Age"))


# linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
# linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
# linearHypothesis(model, "Age = Age+hinge", test = "F")

anova_Ftot <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>%
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_Ftot

# tab_model(model)


```
##### plot
```{r}

# output <- plot_final(breakpoint_Ftot, breakpoint_se_Ftot, 
#                      m_Ftot, 
#                      force_data,
#                      plot_raw_Ftot,
#                      include_brood = FALSE,
#                      brood_level = "Box5",
#                      sex = "M"
#                      )
# plot_Ftot_final <- output[1]
# plot_Ftot_final

```

#### take-off angle - 34.01
##### breakpoint detection
```{r}
out <- detect_breakpoint(angle, force_clean)
print(out)
breakpoint_angle <- round(out$estimates["breakpoint"], 1)
breakpoint_se_angle <- round(out$estimates["breakpoint_se"], 1)
```
##### aic comparison
```{r}

model_results <- compare_models_pmin(angle, breakpoint_angle, force_clean)

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_angle <- model_results[[4]]$m7
m_angle


```

##### model assumption check
```{r}

check_collinearity(m_angle)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(m_angle)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(m_angle) #heteroskedasticity
plot_resqq(m_angle) #residual QQ
plot_ranef(m_angle)
```
##### test
```{r}
model <- m_angle

tab_model(model)
anova_angle <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>%
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_angle

# linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
# linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
# linearHypothesis(model, "Age = Age+hinge", test = "F")
# 
# #before slope
# print(deltaMethod(model, "Age + hinge"))
# #after slope
# print(deltaMethod(model, "Age"))
```
##### plot
```{r}

# output_M <- plot_final(breakpoint_angle, breakpoint_se_angle, 
#                      m_angle, 
#                      force_clean,
#                      plot_age_angle,
#                      include_brood = TRUE,
#                      brood_level = "Box5",
#                      sex = "M"
#                      )
# output_F <- plot_final(breakpoint_angle, breakpoint_se_angle, 
#                      m_angle, 
#                      force_clean,
#                      plot_age_angle,
#                      include_brood = TRUE,
#                      brood_level = "Box5",
#                      sex = "F"
#                      )
# 
# 
# plot_angle_M <- output_M[1]
# plot_angle_F <- output_F[1]
# plot_angle_M
# plot_angle_F
```
##### plot with brood & Sex
- salmon (F) and blue (M) are in the same brood and green (M) in a different brood. 
```{r}

# # 1) Hinge coding
# force_data$hinge <- pmin(force_data$Age - breakpoint_angle, 0)
# 
# # 2) Age sequence
# age_seq <- seq(min(force_data$Age, na.rm = TRUE), max(force_data$Age, na.rm = TRUE), by = 1)
# 
# # 4) Base newdat
# newdat_F_wpbr <- data.frame(
#     Age = age_seq,
#     Sex = "F",
#     Brood = "WhitePinkR-BlueR",
#     hinge = pmin(age_seq - breakpoint_angle, 0)
#   )
# 
# newdat_M_wpbr <- data.frame(
#     Age = age_seq,
#     Sex = "M",
#     Brood = "WhitePinkR-BlueR",
#     hinge = pmin(age_seq - breakpoint_angle, 0)
#   )
# 
# newdat_M_Box11 <- data.frame(
#   Age = age_seq, 
#   Sex = "M",
#   Brood = "Box11",
#   hinge = pmin(age_seq - breakpoint_angle, 0)
# )
# 
# # 6) Predict (population-level)
# newdat_F_wpbr$pred <- predict(m_angle, newdat_F_wpbr, re.form = NA)
# newdat_M_wpbr$pred <- predict(m_angle, newdat_M_wpbr, re.form = NA)
# newdat_M_Box11$pred <- predict(m_angle, newdat_M_Box11, re.form = NA)
# 
# 
# # 7) Compose plot
# yr <- ggplot_build(plot_age_angle)$layout$panel_params[[1]]$y.range
# y_top <- max(yr)
# 
# p <- plot_age_angle +
#     # prediction line no.1 
#     ggplot2::geom_line(
#       data = newdat_F_wpbr,
#       ggplot2::aes(x = Age, y = pred),
#       color = "#E08D79", size = 1.5, inherit.aes = FALSE
#     ) +
#   
#    # prediction line no.2 
#     ggplot2::geom_line(
#       data = newdat_M_wpbr,
#       ggplot2::aes(x = Age, y = pred),
#       color = "#1f77b4", size = 1.5, inherit.aes = FALSE
#     ) +
#    # prediction line no. 3
#     ggplot2::geom_line(
#       data = newdat_M_Box11,
#       ggplot2::aes(x = Age, y = pred),
#       color = "#4D9078", size = 1.5, inherit.aes = FALSE
#     ) +
#   
#     ggplot2::geom_vline(xintercept = breakpoint_angle, linetype = "dashed", color = "red") +
#     ggplot2::annotate(
#         "segment",
#         x = breakpoint_angle - breakpoint_se_angle,
#         xend = breakpoint_angle + breakpoint_se_angle,
#         y    = y_top,
#         yend = y_top,
#         color = "#E2AFDE",
#         linewidth = 3,
#         alpha = 0.7) +
#     ggplot2::labs(subtitle = sprintf("breakpoint = %.1f (± %.1f SE)", breakpoint_angle, breakpoint_se_angle))
#   
# p

```

#### impulse -30.2
##### breakpoint detection
```{r}
#getting rid of 3702006 because there is no connected flight data and seems like there were three push attempts 

out <- detect_breakpoint(impulse*1000, force_clean)
print(out)
breakpoint_impulse <- round(out$estimates["breakpoint"], 1)
breakpoint_se_impulse <- round(out$estimates["breakpoint_se"], 1)
```
##### aic comparison
```{r}

model_results <- compare_models_pmin(impulse*1000, breakpoint_impulse, force_clean)

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_impulse <- model_results[[4]]$m5
m_impulse


```
##### model assumption check
```{r}

check_collinearity(m_impulse)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(m_impulse)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(m_impulse) #heteroskedasticity
plot_resqq(m_impulse) #residual QQ
plot_ranef(m_impulse)
```
##### test
```{r}
model <- m_impulse

tab_model(model)
anova_impulse <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>%
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_impulse

# linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
# linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
# linearHypothesis(model, "Age = Age+hinge", test = "F")
# 
# #before slope
# print(deltaMethod(model, "Age + hinge"))
# #after slope
# print(deltaMethod(model, "Age"))



```

##### plot
```{r}

# output <- plot_final3(breakpoint_impulse, breakpoint_se_impulse, 
#                      m_impulse, 
#                      force_clean,
#                      plot_age_impulse,
#                      include_brood = TRUE,
#                      brood_level = "Box3",
#                      sex = "M"
#                      )
# plot_impulse_final <- output[1]
# plot_impulse_final


```

#### left impulse -32
##### breakpoint detection
```{r}
m_lin <- lm(left_impulse*1000 ~ Age*Sex + Brood, data = force_clean)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_left_impulse <- m_seg$psi["psi1.Age", "Est."]
breakpoint_se_left_impulse <- m_seg$psi["psi1.Age", "St.Err"]
slope_early_left_impulse <- slope(m_seg)$Age[1, "Est."]
slope_late_left_impulse  <- slope(m_seg)$Age[2, "Est."]

breakpoint_left_impulse
slope_early_left_impulse
slope_late_left_impulse
```
##### aic comparison
```{r}

model_results <- compare_models_pmin(left_impulse*1000, breakpoint_left_impulse, force_clean)

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_left_impulse <- model_results[[4]]$m5
m_left_impulse


```
##### model assumptions
```{r}
check_collinearity(m_left_impulse)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(m_left_impulse)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(m_left_impulse) #heteroskedasticity
plot_resqq(m_left_impulse) #residual QQ
plot_ranef(m_left_impulse)


```
##### test
```{r}

model <- m_left_impulse

#tab_model(model)
anova_left_impulse <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>%
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_left_impulse

# linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
# linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
# linearHypothesis(model, "Age = Age+hinge", test = "F")
# 
# #before slope
# print(deltaMethod(model, "Age + hinge"))
# #after slope
# print(deltaMethod(model, "Age"))


```
##### plot
```{r}

# output <- plot_final(breakpoint_left_impulse, breakpoint_se_left_impulse, 
#                      m_left_impulse, 
#                      force_clean,
#                      plot_left_impulse,
#                      include_brood = TRUE,
#                      brood_level = "Box10",
#                      sex = "M"
#                      )
# plot_left_impulse_final <- output[1]
# plot_left_impulse_final


```
#### right impulse - 24.9
##### breakpoint detection
```{r}
m_lin <- lm(right_impulse*1000 ~ Age*Sex + Brood, data = force_clean)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_right_impulse <- m_seg$psi["psi1.Age", "Est."] #36.5. # male max resultant force generally stronger than females.
breakpoint_se_right_impulse <- m_seg$psi["psi1.Age", "St.Err"]
slope_early <- slope(m_seg)$Age[1, "Est."]
slope_late  <- slope(m_seg)$Age[2, "Est."]

breakpoint_right_impulse
breakpoint_se_right_impulse
```
##### aic comparison
```{r}

model_results <- compare_models_pmin(right_impulse*1000, breakpoint_right_impulse, force_clean)

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_right_impulse <- model_results[[4]]$m5
m_right_impulse
```
##### model assumptions
```{r}
model <- m_right_impulse
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)
```
##### test
```{r}

model <- m_right_impulse

#tab_model(model)
anova_right_impulse <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>%
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_right_impulse

# linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
# linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
# linearHypothesis(model, "Age = Age+hinge", test = "F")
# 
# #before slope
# print(deltaMethod(model, "Age + hinge"))
# #after slope
# print(deltaMethod(model, "Age"))
```
##### plot
```{r}

# output <- plot_final(breakpoint_right_impulse, breakpoint_se_right_impulse, 
#                      m_right_impulse, 
#                      force_clean,
#                      plot_right_impulse,
#                      include_brood = TRUE,
#                      brood_level = "Box11",
#                      sex = "M"
#                      )
# plot_right_impulse_final <- output[1]
# plot_right_impulse_final


```
#### impulse ratio 
##### breakpoint detection
```{r}

m_lin <- lm(impulse_ratio ~ Age*Sex + Brood, data = force_clean) #CHANGE HERE
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_impulse_ratio <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
breakpoint_se_impulse_ratio <- m_seg$psi["psi1.Age", "St.Err"] # breakpoint se.


breakpoint_impulse_ratio #32.9
breakpoint_se_impulse_ratio


```
##### aic comparison
```{r}

model_results <- compare_models_pmin(impulse_ratio, breakpoint_impulse_ratio, force_clean) #CHANGE

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_impulse_ratio <- model_results[[4]]$m10 # in theory could
m_impulse_ratio

```
##### model assumptions
```{r}

model <- m_impulse_ratio
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)



```
##### test
```{r}

model <- m_impulse_ratio

#tab_model(model)
anova_impulse_ratio <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_impulse_ratio # CHANGE THIS!!

# linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
# linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
# linearHypothesis(model, "Age = Age+hinge", test = "F")
# 
# #before slope
# print(deltaMethod(model, "Age + hinge"))
# #after slope
# print(deltaMethod(model, "Age"))

```
##### plot
```{r}

# output <- plot_final(breakpoint_impulse_ratio, breakpoint_se_impulse_ratio, #CHANGE HERE
#                      m_impulse_ratio, # CHANGE HERE
#                      force_clean, # CHANGE HERE
#                      plot_age_impulse_ratio, # CHANGE HERE
#                      include_brood = TRUE,
#                      brood_level = "Box3",
#                      sex = "M"
#                      )
# plot_impulse_ratio_final <- output[1] #CHANGE HERE
# plot_impulse_ratio_final # CHANGE HERE


```
#### duration - 39.8
##### breakpoint detection
```{r}
m_lin <- lm(100*duration ~ Age*Sex + Brood, data = force_clean)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_duration <- m_seg$psi["psi1.Age", "Est."] #36.5. # male max resultant force generally stronger than females.
breakpoint_se_duration <- m_seg$psi["psi1.Age", "St.Err"] #36.5. # male max resultant force generally stronger than females.


breakpoint_duration
breakpoint_se_duration
```
##### aic comparison
```{r}

model_results <- compare_models_pmin(1000*duration, breakpoint_duration, force_clean)

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_duration <- model_results[[4]]$m5 # in theory could
m_duration


```
##### model assumptions
```{r}

model <- m_duration
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)



```
##### test
```{r}

model <- m_duration

#tab_model(model)
anova_duration <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>%
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_duration

# linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
# linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
# linearHypothesis(model, "Age = Age+hinge", test = "F")
# 
# #before slope
# print(deltaMethod(model, "Age + hinge"))
# #after slope
# print(deltaMethod(model, "Age"))

```
##### plot
```{r}

# output <- plot_final(breakpoint_duration, breakpoint_se_duration, 
#                      m_duration, 
#                      force_clean,
#                      plot_age_duration,
#                      include_brood = TRUE,
#                      brood_level = "orangeR-pinkR",
#                      sex = "M"
#                      )
# plot_duration_final <- output[1]
# plot_duration_final

```


#### left duration - 40
##### breakpoint detection
```{r}
m_lin <- lm(1000*left_duration ~ Age*Sex + Brood, data = force_clean)
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_left_duration <- m_seg$psi["psi1.Age", "Est."] #36.5. # male max resultant force generally stronger than females.
breakpoint_se_left_duration <- m_seg$psi["psi1.Age", "St.Err"]
slope_early <- slope(m_seg)$Age[1, "Est."]
slope_late  <- slope(m_seg)$Age[2, "Est."]

breakpoint_left_duration
breakpoint_se_left_duration
slope_early
slope_late
```
##### aic comparison
```{r}

model_results <- compare_models_pmin(1000*left_duration, breakpoint_left_duration, force_clean)

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_left_duration <- model_results[[4]]$m5 # in theory could
m_left_duration
```
##### model assumption
```{r}

model <- m_left_duration
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)
```
##### test
```{r}

model <- m_left_duration

#tab_model(model)
anova_left_duration <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_left_duration # CHANGE THIS!!

# linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
# linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
# linearHypothesis(model, "Age = Age+hinge", test = "F")
# 
# #before slope
# print(deltaMethod(model, "Age + hinge"))
# #after slope
# print(deltaMethod(model, "Age"))

```
##### plot
```{r}

# output <- plot_final(breakpoint_left_duration, breakpoint_se_left_duration, #CHANGE HERE
#                      m_left_duration, # CHANGE HERE
#                      force_clean,
#                      plot_left_duration, # CHANGE HERE
#                      include_brood = TRUE,
#                      brood_level = "Box11",
#                      sex = "M"
#                      )
# plot_left_duration_final <- output[1] #CHANGE HERE
# plot_left_duration_final # CHANGE HERE

```
#### right duration -33.1
##### breakpoint detection
```{r}
m_lin <- lm(1000*right_duration ~ Age*Sex + Brood, data = force_clean) # CHANGE HERE
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_right_duration <- m_seg$psi["psi1.Age", "Est."] 
breakpoint_se_right_duration <- m_seg$psi["psi1.Age", "St.Err"] 

slope_early <- slope(m_seg)$Age[1, "Est."]
slope_late  <- slope(m_seg)$Age[2, "Est."]

breakpoint_right_duration
breakpoint_se_right_duration
slope_early
slope_late
```
##### aic comparison
```{r}

model_results <- compare_models_pmin(1000*right_duration, breakpoint_right_duration, force_clean)

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_right_duration <- model_results[[4]]$m5 # in theory could
m_right_duration

```

##### model assumptions
```{r}
model <- m_right_duration
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)

```

##### test
```{r}
model <- m_right_duration

#tab_model(model)
anova_right_duration <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_right_duration # CHANGE THIS!!

# linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
# linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
# linearHypothesis(model, "Age = Age+hinge", test = "F")
# 
# #before slope
# print(deltaMethod(model, "Age + hinge"))
# #after slope
# print(deltaMethod(model, "Age"))

```

##### plot
```{r}
# output <- plot_final(breakpoint_right_duration, breakpoint_se_right_duration, #CHANGE HERE
#                      m_right_duration, # CHANGE HERE
#                      force_clean,
#                      plot_right_duration, # CHANGE HERE
#                      include_brood = TRUE,
#                      brood_level = "Box5",
#                      sex = "M"
#                      )
# plot_right_duration_final <- output[1] #CHANGE HERE
# plot_right_duration_final # CHANGE HERE

```




### TRAJECTORY
#### Z velocity - 32.7
##### breakpoint duration
```{r}

m_lin <- lm(mean_vz ~ Age*Sex + Brood, data = flights_speed) #CHANGE HERE
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_vz <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
breakpoint_se_vz <- m_seg$psi["psi1.Age", "St.Err"] # breakpoint se.


breakpoint_vz #32.9
breakpoint_se_vz


```
##### aic comparison
```{r}

model_results <- compare_models_pmin(mean_vz, breakpoint_vz, flights_speed) #CHANGE

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_vz <- model_results[[4]]$m10 # in theory could
m_vz

```
##### model assumptions
```{r}

model <- m_vz
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)



```
##### test
```{r}

model <- m_vz

#tab_model(model)
anova_vz <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_vz # CHANGE THIS!!

# linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
# linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
# linearHypothesis(model, "Age = Age+hinge", test = "F")
# 
# #before slope
# print(deltaMethod(model, "Age + hinge"))
# #after slope
# print(deltaMethod(model, "Age"))

```
##### plot
```{r}

# output <- plot_final(breakpoint_vz, breakpoint_se_vz, #CHANGE HERE
#                      m_vz, # CHANGE HERE
#                      flights_speed, # CHANGE HERE
#                      plot_vz_mean, # CHANGE HERE
#                      include_brood = FALSE,
#                      brood_level = "Box5",
#                      sex = "M"
#                      )
# plot_vz_final <- output[1] #CHANGE HERE
# plot_vz_final # CHANGE HERE


```
#### Y velocity - 33.0
##### breakpoint duration
```{r}

m_lin <- lm(mean_vy ~ Age*Sex + Brood, data = flights_speed) #CHANGE HERE
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_vy<- m_seg$psi["psi1.Age", "Est."] 
breakpoint_se_vy <- m_seg$psi["psi1.Age", "St.Err"] 


breakpoint_vy #32.9
breakpoint_se_vy


```
##### aic comparison
```{r}

model_results <- compare_models_pmin(mean_vy, breakpoint_vy, flights_speed) #CHANGE

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_vy <- model_results[[4]]$m10 # in theory could
m_vy

```
##### model assumptions
```{r}

model <- m_vy
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)



```
##### test
```{r}

model <- m_vy

#tab_model(model)
anova_vy <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_vy # CHANGE THIS!!

# linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
# linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
# linearHypothesis(model, "Age = Age+hinge", test = "F")
# 
# #before slope
# print(deltaMethod(model, "Age + hinge"))
# #after slope
# print(deltaMethod(model, "Age"))

```
##### plot
```{r}

# output <- plot_final(breakpoint_vy, breakpoint_se_vy, #CHANGE HERE
#                      m_vy, # CHANGE HERE
#                      flights_speed, # CHANGE HERE
#                      plot_vy_mean, # CHANGE HERE
#                      include_brood = FALSE,
#                      brood_level = "Box5",
#                      sex = "M"
#                      )
# plot_vy_final <- output[1] #CHANGE HERE
# plot_vy_final # CHANGE HERE


```
#### X velocity - 60
##### breakpoint duration
```{r}

m_lin <- lm(mean_vx ~ Age*Sex + Brood, data = flights_speed) #CHANGE HERE
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_vx <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
breakpoint_se_vx <- m_seg$psi["psi1.Age", "St.Err"] # breakpoint se.


breakpoint_vx #32.9
breakpoint_se_vx


```
##### aic comparison
```{r}

model_results <- compare_models_pmin(mean_vx, breakpoint_vx, flights_speed) #CHANGE

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_vx <- model_results[[4]]$m10 # in theory could
m_vx

```
##### model assumptions
```{r}

model <- m_vx
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)



```
##### test
```{r}

model <- m_vx

#tab_model(model)
anova_vx <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_vx # CHANGE THIS!!

# linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
# linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
# linearHypothesis(model, "Age = Age+hinge", test = "F")
# 
# #before slope
# print(deltaMethod(model, "Age + hinge"))
# #after slope
# print(deltaMethod(model, "Age"))

```
##### plot
```{r}

# output <- plot_final(breakpoint_vx, breakpoint_se_vx, #CHANGE HERE
#                      m_vx, # CHANGE HERE
#                      flights_speed, # CHANGE HERE
#                      plot_vx_mean, # CHANGE HERE
#                      include_brood = FALSE,
#                      brood_level = "Box5",
#                      sex = "M"
#                      )
# plot_vx_final <- output[1] #CHANGE HERE
# plot_vx_final # CHANGE HERE


```
#### X acceleration -75.1
##### breakpoint duration
```{r}

m_lin <- lm(mean_acc_x ~ Age*Sex + Brood, data = flights_speed) #CHANGE HERE
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_accx <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
breakpoint_se_accx <- m_seg$psi["psi1.Age", "St.Err"] # breakpoint se.


breakpoint_accx #32.9
breakpoint_se_accx


```
##### aic comparison
```{r}

model_results <- compare_models_pmin(mean_acc_x, breakpoint_accx, flights_speed) #CHANGE

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_accx <- model_results[[4]]$m8 # in theory could
m_accx

```
##### model assumptions
```{r}

model <- m_accx
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)



```
##### test
```{r}

model <- m_accx

#tab_model(model)
anova_accx <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_vx # CHANGE THIS!!

# linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
# linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
# linearHypothesis(model, "Age = Age+hinge", test = "F")
# 
# #before slope
# print(deltaMethod(model, "Age + hinge"))
# #after slope
# print(deltaMethod(model, "Age"))

```
##### plot
```{r}

# output <- plot_final(breakpoint_accx, breakpoint_se_accx, #CHANGE HERE
#                      m_accx, # CHANGE HERE
#                      flights_speed, # CHANGE HERE
#                      plot_accx_mean, # CHANGE HERE
#                      include_brood = FALSE,
#                      brood_level = "Box5",
#                      sex = "M"
#                      )
# plot_accx_final <- output[1] #CHANGE HERE
# plot_accx_final # CHANGE HERE


```

#### Y acceleration -33.7
##### breakpoint duration
```{r}

m_lin <- lm(mean_acc_y ~ Age*Sex + Brood, data = flights_speed) #CHANGE HERE
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_accy <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
breakpoint_se_accy <- m_seg$psi["psi1.Age", "St.Err"] # breakpoint se.


breakpoint_accy #32.9
breakpoint_se_accy


```
##### aic comparison
```{r}

model_results <- compare_models_pmin(mean_acc_y, breakpoint_accy, flights_speed) #CHANGE

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_accy <- model_results[[4]]$m6 # in theory could
m_accy

```
##### model assumptions
```{r}

model <- m_accy
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)



```
##### test
```{r}

model <- m_accy

#tab_model(model)
anova_accy <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_accy # CHANGE THIS!!

# linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
# linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
# linearHypothesis(model, "Age = Age+hinge", test = "F")
# 
# #before slope
# print(deltaMethod(model, "Age + hinge"))
# #after slope
# print(deltaMethod(model, "Age"))

```
##### plot
```{r}

# output <- plot_final(breakpoint_accy, breakpoint_se_accy, #CHANGE HERE
#                      m_accy, # CHANGE HERE
#                      flights_speed, # CHANGE HERE
#                      plot_accy_mean, # CHANGE HERE
#                      include_brood = FALSE,
#                      brood_level = "Box5",
#                      sex = "M"
#                      )
# plot_accy_final <- output[1] #CHANGE HERE
# plot_accy_final # CHANGE HERE


```
#### Z acceleration -34.2
##### breakpoint duration
```{r}

m_lin <- lm(mean_acc_z ~ Age*Sex + Brood, data = flights_speed) #CHANGE HERE
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_accz <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
breakpoint_se_accz <- m_seg$psi["psi1.Age", "St.Err"] # breakpoint se.


breakpoint_accz #32.9
breakpoint_se_accz


```
##### aic comparison
```{r}

model_results <- compare_models_pmin(mean_acc_z, breakpoint_accz, flights_speed) #CHANGE

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_accz <- model_results[[4]]$m6 # in theory could
m_accz

```
##### model assumptions
```{r}

model <- m_accz
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)



```
##### test
```{r}

model <- m_accz

#tab_model(model)
anova_accz <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_accz# CHANGE THIS!!

# linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
# linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
# linearHypothesis(model, "Age = Age+hinge", test = "F")
# 
# #before slope
# print(deltaMethod(model, "Age + hinge"))
# #after slope
# print(deltaMethod(model, "Age"))

```
##### plot
```{r}

# output <- plot_final(breakpoint_accz, breakpoint_se_accz, #CHANGE HERE
#                      m_accz, # CHANGE HERE
#                      flights_speed, # CHANGE HERE
#                      plot_accz_mean, # CHANGE HERE
#                      include_brood = FALSE,
#                      brood_level = "Box5",
#                      sex = "M"
#                      )
# plot_accz_final <- output[1] #CHANGE HERE
# plot_accz_final # CHANGE HERE


```
#### tangential acceleration - 33.9
##### breakpoint duration
```{r}

m_lin <- lm(mean_acc_tan ~ Age*Sex + Brood, data = flights_speed) #CHANGE HERE
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_tacc <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
breakpoint_se_tacc <- m_seg$psi["psi1.Age", "St.Err"] # breakpoint se.


breakpoint_tacc #32.9
breakpoint_se_tacc


```
##### aic comparison
```{r}

model_results <- compare_models_pmin(mean_acc_tan, breakpoint_tacc, flights_speed) #CHANGE

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_tacc <- model_results[[4]]$m10 # in theory could
m_tacc

```
##### model assumptions
```{r}

model <- m_tacc
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)



```
##### test
```{r}

model <- m_tacc

#tab_model(model)
anova_tacc <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_tacc # CHANGE THIS!!

# linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
# linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
# linearHypothesis(model, "Age = Age+hinge", test = "F")
# 
# #before slope
# print(deltaMethod(model, "Age + hinge"))
# #after slope
# print(deltaMethod(model, "Age"))

```
##### plot
```{r}

# output <- plot_final(breakpoint_tacc, breakpoint_se_tacc, #CHANGE HERE
#                      m_tacc, # CHANGE HERE
#                      flights_speed, # CHANGE HERE
#                      plot_tacc_mean, # CHANGE HERE
#                      include_brood = FALSE,
#                      brood_level = "Box5",
#                      sex = "M"
#                      )
# plot_tacc_final <- output[1] #CHANGE HERE
# plot_tacc_final # CHANGE HERE


```

#### velocity magnitude - 33.5
##### breakpoint duration
```{r}

m_lin <- lm(mean_v_res ~ Age*Sex + Brood, data = flights_speed) #CHANGE HERE
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_vres <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
breakpoint_se_vres <- m_seg$psi["psi1.Age", "St.Err"] # breakpoint se.


breakpoint_vres #32.9
breakpoint_se_vres


```
##### aic comparison
```{r}

model_results <- compare_models_pmin(mean_v_res, breakpoint_vres, flights_speed) #CHANGE

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_vres <- model_results[[4]]$m10 # in theory could
m_vres

```
##### model assumptions
```{r}

model <- m_vres
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)



```
##### test
```{r}

model <- m_vres

#tab_model(model)
anova_vres <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_vres # CHANGE THIS!!

# linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
# linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
# linearHypothesis(model, "Age = Age+hinge", test = "F")
# 
# #before slope
# print(deltaMethod(model, "Age + hinge"))
# #after slope
# print(deltaMethod(model, "Age"))

```
##### plot
```{r}

# output <- plot_final(breakpoint_vres, breakpoint_se_vres, #CHANGE HERE
#                      m_vres, # CHANGE HERE
#                      flights_speed, # CHANGE HERE
#                      plot_vres_mean, # CHANGE HERE
#                      include_brood = FALSE,
#                      brood_level = "Box5",
#                      sex = "M"
#                      )
# plot_vres_final <- output[1] #CHANGE HERE
# plot_vres_final # CHANGE HERE


```

#### turn rate - 40.0
##### breakpoint duration
```{r}

m_lin <- lm(mean_turn_rate ~ Age*Sex + Brood, data = flights_tort) #CHANGE HERE
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_tr <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
breakpoint_se_tr <- m_seg$psi["psi1.Age", "St.Err"] # breakpoint se.


breakpoint_tr #32.9
breakpoint_se_tr


```
##### aic comparison
```{r}

model_results <- compare_models_pmin(mean_turn_rate, breakpoint_tr, flights_speed) #CHANGE

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_tr <- model_results[[4]]$m7 # in theory could
m_tr

```
##### model assumptions
```{r}

model <- m_tr
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)



```
##### test
```{r}

model <- m_tr

#tab_model(model)
anova_tr <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_tr # CHANGE THIS!!

# linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
# linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
# linearHypothesis(model, "Age = Age+hinge", test = "F")
# 
# #before slope
# print(deltaMethod(model, "Age + hinge"))
# #after slope
# print(deltaMethod(model, "Age"))

```
##### plot
```{r}

# output <- plot_final(breakpoint_tr, breakpoint_se_tr, #CHANGE HERE
#                      m_tr, # CHANGE HERE
#                      flights_tort, # CHANGE HERE
#                      plot_tr_mean, # CHANGE HERE
#                      include_brood = TRUE,
#                      brood_level = "Box3",
#                      sex = "M"
#                      )
# plot_tr_final <- output[1] #CHANGE HERE
# plot_tr_final # CHANGE HERE


```






















#### mean centripetal acceleration
##### breakpoint duration
```{r}

m_lin <- lm(mean_centri_acc ~ Age*Sex + Brood, data = flights_tort) #CHANGE HERE
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_centri <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
breakpoint_se_centri <- m_seg$psi["psi1.Age", "St.Err"] # breakpoint se.


breakpoint_centri #32.9
breakpoint_se_centri


```
##### aic comparison
```{r}

model_results <- compare_models_pmin(mean_centri_acc, breakpoint_centri, flights_tort) #CHANGE

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_centri <- model_results[[4]]$m5 # in theory could
m_centri

```
##### model assumptions
```{r}

model <- m_centri
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)



```
##### test
```{r}

model <- m_centri

#tab_model(model)
anova_centri <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_centri # CHANGE THIS!!

# linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
# linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
# linearHypothesis(model, "Age = Age+hinge", test = "F")
# 
# #before slope
# print(deltaMethod(model, "Age + hinge"))
# #after slope
# print(deltaMethod(model, "Age"))

```
##### plot
```{r}

# output <- plot_final2(breakpoint_centri, breakpoint_se_centri, #CHANGE HERE
#                      m_centri, # CHANGE HERE
#                      flights_speed, # CHANGE HERE
#                      plot_ctracc_mean, # CHANGE HERE
#                      include_brood = TRUE,
#                      brood_level = "Box11",
#                      sex = "M"
#                      )
# plot_centri_final <- output[1] #CHANGE HERE
# plot_centri_final # CHANGE HERE


```
#### wingbeat frequency 
##### breakpoint detection = 40.0
```{r}

m_lin <- lm(wingbeat_frequency ~ Age*Sex + Brood, data = flights_wing) #CHANGE HERE
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_wb <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
breakpoint_se_wb <- m_seg$psi["psi1.Age", "St.Err"] # breakpoint se.


breakpoint_wb #32.9
breakpoint_se_wb


```
##### aic comparison
```{r}

model_results <- compare_models_pmax(wingbeat_frequency, breakpoint_wb, flights_wing) #CHANGE

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_wb <- model_results[[4]]$m5 # in theory could
m_wb

```
##### model assumptions
```{r}

model <- m_wb
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)



```
##### test
```{r}

model <- m_wb

#tab_model(model)
anova_wb <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_wb # CHANGE THIS!!

# linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
# linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
# linearHypothesis(model, "Age = Age+hinge", test = "F")
# 
# #before slope
# print(deltaMethod(model, "Age + hinge"))
# #after slope
# print(deltaMethod(model, "Age"))

```
##### plot
```{r}

# output <- plot_final(breakpoint_wb, breakpoint_se_wb, #CHANGE HERE
#                      m_wb, # CHANGE HERE
#                      flights_wing, # CHANGE HERE
#                      plot_wb_mean, # CHANGE HERE
#                      include_brood = TRUE,
#                      brood_level = "Box5",
#                      sex = "M"
#                      )
# plot_wb_final <- output[1] #CHANGE HERE
# plot_wb_final # CHANGE HERE


```
#### downstroke
##### breakpoint detection = 40.0
```{r}

m_lin <- lm(1000*downstroke_amplitude_m ~ Age*Sex + Brood, data = flights_wing) #CHANGE HERE
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_downstroke <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
breakpoint_se_downstroke <- m_seg$psi["psi1.Age", "St.Err"] # breakpoint se.


breakpoint_downstroke #32.9
breakpoint_se_downstroke


```
##### aic comparison
```{r}

model_results <- compare_models_pmin(1000*downstroke_amplitude_m, breakpoint_downstroke, flights_wing) #CHANGE

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_downstroke <- model_results[[4]]$m5 # in theory could
m_downstroke

```
##### model assumptions
```{r}

model <- m_downstroke
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)



```
##### test
```{r}

model <- m_downstroke

#tab_model(model)
anova_downstroke <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_downstroke # CHANGE THIS!!

# linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
# linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
# linearHypothesis(model, "Age = Age+hinge", test = "F")
# 
# #before slope
# print(deltaMethod(model, "Age + hinge"))
# #after slope
# print(deltaMethod(model, "Age"))

```
##### plot
```{r}

# output <- plot_final(breakpoint_downstroke, breakpoint_se_downstroke, #CHANGE HERE
#                      m_downstroke, # CHANGE HERE
#                      flights_wing, # CHANGE HERE
#                      plot_downstroke_mm, # CHANGE HERE
#                      include_brood = TRUE,
#                      brood_level = "Box5",
#                      sex = "M"
#                      )
# plot_downstroke_final <- output[1] #CHANGE HERE
# plot_downstroke_final # CHANGE HERE


```
#### upstroke - 60 
##### breakpoint detection = 32.0
```{r}

m_lin <- lm(1000*upstroke_amplitude_m ~ Age*Sex + Brood, data = flights_wing) #CHANGE HERE
m_seg <- segmented(m_lin, seg.Z = ~ Age, psi = list(Age = 36))  # initial guess at 36
summary(m_seg)
plot(m_seg)  # quick diagnostic plot

# Extract estimated breakpoint and slopes
breakpoint_upstroke <- m_seg$psi["psi1.Age", "Est."] # breakpoint: 34.3. Age significant, male grows quicker. 
breakpoint_se_upstroke <- m_seg$psi["psi1.Age", "St.Err"] # breakpoint se.


breakpoint_upstroke #32.9
breakpoint_se_upstroke


```
##### aic comparison
```{r}

model_results <- compare_models_pmin(1000*upstroke_amplitude_m, breakpoint_upstroke, flights_wing) #CHANGE

# shows the results of model comparison
model_results[1:3]

# saves the model to test 
m_upstroke <- model_results[[4]]$m5 # in theory could
m_upstroke

```
##### model assumptions
```{r}

model <- m_upstroke
check_collinearity(model)
# Computes the default residuals (raw conditional)
raw_cond_model <- redres::compute_redres(model)
  
hist(raw_cond_model, main = "Norality of residuals") #normality in histogram
plot_redres(model) #heteroskedasticity
plot_resqq(model) #residual QQ
plot_ranef(model)



```
##### test
```{r}

model <- m_upstroke

#tab_model(model)
anova_upstroke <- Anova(model, type =3, test.statistic = "F") %>% as.data.frame() %>% # CHANGE THIS!!
  mutate(p_FDR = p.adjust(`Pr(>F)`, method = "fdr"))

anova_upstroke # CHANGE THIS!!

# linearHypothesis(model, "Age+hinge = 0", test = "F") # Is the slope before the breakpoint flat? -> no
# linearHypothesis(model, "Age = 0", test = "F") # Is the slope after the breakpoint flat? -> no
# linearHypothesis(model, "Age = Age+hinge", test = "F")
# 
# #before slope
# print(deltaMethod(model, "Age + hinge"))
# #after slope
# print(deltaMethod(model, "Age"))

```
##### plot
```{r}

# output <- plot_final(breakpoint_upstroke, breakpoint_se_upstroke, #CHANGE HERE
#                      m_upstroke, # CHANGE HERE
#                      flights_wing, # CHANGE HERE
#                      plot_upstroke_mm, # CHANGE HERE
#                      include_brood = TRUE,
#                      brood_level = "Box5",
#                      sex = "M"
#                      )
# plot_upstroke_final <- output[1] #CHANGE HERE
# plot_upstroke_final # CHANGE HERE


```
### MODEL OUTPUT
#### function fmt_term: F = xx.xx (Df, Df.res) (p-display)
```{r}
# Helper to format one term: F = xx.xx (Df, Df.res) (p-display)
fmt_term <- function(F, Df, Df_res, p, digits_F = 1, digits_Df = 1, p_thresh = 0.001) {
  if (anyNA(c(F, Df, Df_res, p))) return(NA_character_)
  F_str   <- format(round(F, digits_F), nsmall = digits_F, trim = TRUE)
  Df_str  <- format(Df, trim = TRUE)
  Res_str <- format(round(Df_res, digits_Df), nsmall = digits_Df, trim = TRUE)
  
  p_str <- if (is.na(p)) {
      "p = NA"
    } else if (p < p_thresh) {
      "p < 0.001"
    } else {
      paste0("p = ", format(round(p, 2), nsmall = 2))
    }

  sprintf("F<sub>%s, %s</sub> = %s, %s",  Df_str, Res_str, F_str, p_str)

  }
```

```{r}
library(knitr)

labels <- c(
  # morphology -------- 4 
  "scaled_weight", "wingspan", "wingarea", "wing loading",
  #leg force ---------- 12
  "Fx/bw (Fz in data)", "Fy/bw", "Fz/bw (Fx in dat)", "Ftot/bw", 
  "angle", "impulse", "left impulse", "right impulse", "impulse ratio", "duration", "left_duration", "right_duration",
  #trajectory --------- 8 
  "vx", "vy", "vz", "ax", "ay", "az", "|v|","atan", 
  #tortuosity --------- 2
  "turn rate", "centripetal acceleration", 
  #wings ---------------3 ---------------> 29 variables!
  "wingbeat frequency", "downstroke head displacement (mm)", "upstroke head displacement (mm)")


# anova dataframe
anova_tabs <- list(
  # morphology --------- 4
  anova_sw, anova_wingspan, anova_wingarea, anova_wl, 
  
  # leg force ---------- 12
  anova_Fx, anova_Fy, anova_Fz, anova_Ftot, anova_angle, anova_impulse, anova_left_impulse, anova_right_impulse,
  anova_impulse_ratio, anova_duration, anova_left_duration, anova_right_duration,
  
  # trajectory --------- 8 
  anova_vx, anova_vy, anova_vz, anova_accx, anova_accy, anova_accz, anova_vres, anova_tacc,
  
  # tortuosity --------- 2 -----------------> 29 variables 
  anova_tr, anova_centri,
  
  # wings -------------- 3
  anova_wb, anova_downstroke, anova_upstroke
  )

# lmer mods
lmer_mods <- list(
  # morphology --------- 4 
  m_sw, m_wingspan, m_wingarea, m_wl,
  
  # leg force ---------- 12
  m_Fx, m_Fy, m_Fz, m_Ftot, m_angle, m_impulse, m_left_impulse, m_right_impulse, m_impulse_ratio, 
  m_duration, m_left_duration, m_right_duration,
  
  # trajectory --------- 8
  m_vx, m_vy, m_vz, m_accx, m_accy, m_accz, m_vres, m_tacc,
  
  # tortuosity --------- 2
  m_tr, m_centri, 
  
  # wings -------------- 3
  m_wb, m_downstroke, m_upstroke
  )          

#stop if the number of anova tabs != number of models. It should
stopifnot(length(labels) == length(anova_tabs), length(anova_tabs) == length(lmer_mods))

#storage for all rows
out_all <- list()


for (i in seq_along(lmer_mods)){
  
  tab <- anova_tabs[[i]] # the ANOVA data.frame for term extraction
  mod <- lmer_mods[[i]] # the fitted model for summary metrics

  
# --- helper to extract a term safely from the current model ---
  get_term <- function(term, what) {
    # if the term is not present, return NA
    if (!term %in% rownames(tab)) return(NA_real_)
    val <- tab[term, ][[what]]
    if (is.null(val)) return(NA_real_)
    val
  }

  # pull values for each term from *this* model
  ageF      <- get_term("Age",   "F")
  ageDf     <- get_term("Age",   "Df")
  ageDf.res <- get_term("Age",   "Df.res")
  ageP      <- get_term("Age",   "p_FDR")

  SexF      <- get_term("Sex",   "F")
  SexDf     <- get_term("Sex",   "Df")
  SexDf.res <- get_term("Sex",   "Df.res")
  SexP      <- get_term("Sex",   "p_FDR")

  hingeF      <- get_term("hinge", "F")
  hingeDf     <- get_term("hinge", "Df")
  hingeDf.res <- get_term("hinge", "Df.res")
  hingeP      <- get_term("hinge", "p_FDR")
  
  # Brood is optional — if missing, all will be NA
  BroodF      <- get_term("Brood",   "F")
  BroodDf     <- get_term("Brood",   "Df")
  BroodDf.res <- get_term("Brood",   "Df.res")
  BroodP      <- get_term("Brood",   "p_FDR")
  
  #Optional interaction
  AgeSexF      <- get_term("Age:Sex",   "F")
  AgeSexDf     <- get_term("Age:Sex",   "Df")
  AgeSexDf.res <- get_term("Age:Sex",   "Df.res")
  AgeSexP      <- get_term("Age:Sex",   "p_FDR")
  
  # Build the single-row data frame in your requested format ---
  out_row <- data.frame(
    Age   = fmt_term(ageF,   ageDf,   ageDf.res,   ageP),
    Sex   = fmt_term(SexF,   SexDf,   SexDf.res,   SexP),
    Hinge = fmt_term(hingeF, hingeDf, hingeDf.res,  hingeP),# uses the parenthesis style
    Brood = fmt_term(BroodF, BroodDf, BroodDf.res, BroodP),
    AgeSex = fmt_term(AgeSexF, AgeSexDf, AgeSexDf.res, AgeSexP) # uses "p = xx" style
    
    )
  
  # model summary variables
  
  var <- as.data.frame(VarCorr(mod))
  
  if (dim(var)[1] > 2){
    sd_bird <- var$sdcor[1] 
    sd_brood <- var$sdcor[2]
    sd_fixed <- var$sdcor[3]
    ICC <-  (var$sdcor[1] + var$sdcor[2]) / (var$sdcor[1] + var$sdcor[2] +var$sdcor[3])  # how much % of variation is due to individual variation nested in brood?
  }else{
    sd_bird <- var$sdcor[1]
    sd_brood <- NA
    sd_fixed <- var$sdcor[2]
    ICC <- var$sdcor[1] / (var$sdcor[1] + var$sdcor[2]) # how much % of variation is due to individual variation?
  }
  
  
  N_Bird <- ngrps(mod)[1]
  Observations <- stats::nobs(mod)
  R2_marginal <- performance::r2(mod, tolerance = 0 )$R2_marginal
  R2_conditional <- performance::r2(mod, tolerance = 0 )$R2_conditional

  summary_cols <- data.frame(
    sd_fixed   = round(sd_fixed, 2),
    sd_bird    = round(sd_bird, 2),
    sd_brood = ifelse(is.na(sd_brood), NA, round(sd_brood,2 )),
    ICC           = round(ICC, 2),
    N_Bird        = N_Bird,
    Observations  = Observations,
    `Marginal R2 / Conditional R2` = paste0(round(R2_marginal,3), " / ", round(R2_conditional,3)),
    check.names = FALSE, 
    row.names = NULL
  )

  
  # Add the Variables column as the first column
  out_row <- cbind(Variables = labels[i], out_row, summary_cols)
  
  out_all[[i]] <- out_row
}

# Combine rows from all models into one data frame
out_all <- do.call(rbind, out_all)

kable(out_all, escape = FALSE)

```
### functions for plots
```{r}


# ---- dependencies
#library(ggplot2)

sanitize_filename <- function(x) {
  x <- gsub("[^A-Za-z0-9]+", "_", x)
  x <- gsub("^_+|_+$", "", x)
  tolower(x)
}

fmt_term_plain <- function(F, Df, Df_res, p, digits_F = 1, digits_Df = 1, p_thresh = 0.001) {
  if (any(is.na(c(F, Df, Df_res, p)))) return("F = NA, p = NA")
  F_str   <- format(round(F, digits_F), nsmall = digits_F, trim = TRUE)
  Df_str  <- format(Df, trim = TRUE)
  Res_str <- format(round(Df_res, digits_Df), nsmall = digits_Df, trim = TRUE)

  p_str <- if (is.na(p)) {
    "p = NA"
  } else if (p < p_thresh) {
    "p < 0.001"
  } else {
    paste0("p = ", format(round(p, 2), nsmall = 2))
  }

  sig <- if (!is.na(p) && p < 0.05) " (p < 0.05)" else ""
  sprintf("F[%s, %s] = %s, %s%s", Df_str, Res_str, F_str, p_str, sig)
}

make_model_plot_plain <- function(
  mod, label, breakpoint, breakpoint_se,
  before_string, after_string,
  before_term_txt, after_term_txt, same_term_txt
) {
  df <- model.frame(mod)
  y_name <- all.vars(formula(mod))[1]

  base_plot <- ggplot(df, aes(x = Age, y = .data[[y_name]])) +
    geom_point(alpha = 0.6) +
    theme_bw() +
    labs(title = label, x = "Age", y = label)

  p <- plot_breakpoint(breakpoint, breakpoint_se, base_plot)

  cap <- paste0(
    "Slope summaries: before = ", before_string, "; after = ", after_string, "\n",
    "Tests: before = 0? ", before_term_txt, "; ",
    "after = 0? ",  after_term_txt, "; ",
    "before = after? ", same_term_txt
  )

  p + labs(caption = cap) + theme(plot.caption = element_text(size = 9))
}

save_model_plot_plain <- function(
  mod, label, breakpoint, breakpoint_se,
  before_string, after_string,
  before_term_txt, after_term_txt, same_term_txt,
  out_dir = "plots", width = 7, height = 5, dpi = 300
) {
  if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)
  p <- make_model_plot_plain(
    mod, label, breakpoint, breakpoint_se,
    before_string, after_string,
    before_term_txt, after_term_txt, same_term_txt
  )
  fn <- file.path(out_dir, paste0("plot_model_", sanitize_filename(label), ".png"))
  ggplot2::ggsave(fn, plot = p, width = width, height = height, dpi = dpi)
  fn
}



```
### TEST OUTPUT
```{r}

library(knitr)

labels <- c(
  # morphology -------- 4 
  "scaled_weight", "wingspan", "wingarea", "wing loading",
  #leg force ---------- 12
  "Fx/bw (Fz in data)", "Fy/bw", "Fz/bw (Fx in dat)", "Ftot/bw", 
  "angle", "impulse", "left impulse", "right impulse", "impulse ratio", "duration", "left_duration", "right_duration",
  #trajectory --------- 8 
  "vx", "vy", "vz", "ax", "ay", "az", "|v|","atan", 
  #tortuosity --------- 2
  "turn rate", "centripetal acceleration", 
  #wings ---------------3 ---------------> 29 variables!
  "wingbeat frequency", "downstroke head displacement (mm)", "upstroke head displacement (mm)")


# lmer mods
lmer_mods <- list(
  # morphology --------- 4 
  m_sw, m_wingspan, m_wingarea, m_wl,
  
  # leg force ---------- 12
  m_Fx, m_Fy, m_Fz, m_Ftot, m_angle, m_impulse, m_left_impulse, m_right_impulse, m_impulse_ratio, 
  m_duration, m_left_duration, m_right_duration,
  
  # trajectory --------- 8
  m_vx, m_vy, m_vz, m_accx, m_accy, m_accz, m_vres, m_tacc,
  
  # tortuosity --------- 2
  m_tr, m_centri, 
  
  # wings -------------- 3
  m_wb, m_downstroke, m_upstroke
  )          

# breakpoints
breakpoints <- c(
  # morphology
  breakpoint_sw, breakpoint_ws, breakpoint_wa, breakpoint_wl,
  
  # leg force & impulse
  breakpoint_Fx, breakpoint_Fy, breakpoint_Fz, breakpoint_Ftot,
  breakpoint_angle, breakpoint_impulse, breakpoint_left_impulse,
  breakpoint_right_impulse, breakpoint_impulse_ratio, breakpoint_duration, breakpoint_left_duration, breakpoint_right_duration,
  
  # flight speed
  breakpoint_vx, breakpoint_vy, breakpoint_vz, breakpoint_accx, breakpoint_accy, breakpoint_accz, breakpoint_vres, breakpoint_tacc,
  
  # tortuosity
  breakpoint_tr, breakpoint_centri, 
  
  # wings
  breakpoint_wb, breakpoint_downstroke, breakpoint_upstroke)
  
breakpoints_se <- c(
  # morphology
  breakpoint_se_sw, breakpoint_se_ws, breakpoint_se_wa, breakpoint_se_wl,
  
  # leg force & impulse
  breakpoint_se_Fx, breakpoint_se_Fy, breakpoint_se_Fz, breakpoint_se_Ftot,
  breakpoint_se_angle, breakpoint_se_impulse, breakpoint_se_left_impulse,
  breakpoint_se_right_impulse, breakpoint_se_impulse_ratio, breakpoint_se_duration, breakpoint_se_left_duration, breakpoint_se_right_duration,
  
  # flight speed
  breakpoint_vx, breakpoint_vy, breakpoint_vz, breakpoint_accx, breakpoint_accy, breakpoint_accz, breakpoint_vres, breakpoint_tacc,
  
  # tortuosity
  breakpoint_tr, breakpoint_centri, 
  
  # wings
  breakpoint_wb, breakpoint_downstroke, breakpoint_upstroke)

  
breakpoints_se <- c()

out_all <- vector("list", length(lmer_mods))

plot_files <- vector("character", length(lmer_mods))


for (i in seq_along(lmer_mods)){
  
  #tab <- anova_tabs[[i]] # the ANOVA data.frame for term extraction
  mod <- lmer_mods[[i]] # the fitted model for summary metrics

  #before slope = 0?
  before_slope <- linearHypothesis(mod, "Age+hinge = 0", test = "F") %>% as.data.frame()# Is the slope before the breakpoint flat? -> no
  
  #after slope = 0?
  after_slope <- linearHypothesis(mod, "Age = 0", test = "F") %>% as.data.frame() # Is the slope after the breakpoint flat? -> no
  
  #same slope?
  same_slope <- linearHypothesis(mod, "Age = Age+hinge", test = "F") %>% as.data.frame()#Is the slope before and after the same? -> n
  
  #before breakpoint slope estimate
  before_slope_estimate <- deltaMethod(mod, "Age + hinge") %>% as.data.frame()
  
  #after breakpoint slope estimate
  after_slope_estimate <- deltaMethod(mod, "Age") %>% as.data.frame()
  
  before_slope_mean <-  before_slope_estimate$Estimate 
  before_slope_se <-  before_slope_estimate$SE
  before_slope_lower <- before_slope_estimate$`2.5 %`
  before_slope_upper <- before_slope_estimate$`97.5 %`
  
  before_slope_F <- before_slope$F[2]
  before_slope_Df <- before_slope$Df[2]
  before_slope_ResDf <- before_slope$Res.Df[2]
  before_slope_P <- before_slope$`Pr(>F)`[2]
  
  after_slope_mean <-  after_slope_estimate$Estimate 
  after_slope_se <-  after_slope_estimate$SE
  after_slope_lower <- after_slope_estimate$`2.5 %`
  after_slope_upper <- after_slope_estimate$`97.5 %`
  
  after_slope_F <- after_slope$F[2]
  after_slope_Df <- after_slope$Df[2]
  after_slope_ResDf <- after_slope$Res.Df[2]
  after_slope_P <- after_slope$`Pr(>F)`[2]
  
  same_slope_F <- same_slope$F[2]
  same_slope_Df <- same_slope$Df[2]
  same_slope_ResDf <- same_slope$Res.Df[2]
  same_slope_P <- same_slope$`Pr(>F)`[2]
  
  
  before_string <- sprintf(
    "%.2f ± %.2f [%.2f, %.2f]",
    before_slope_mean,
    before_slope_se,
    before_slope_lower,
    before_slope_upper)
  
  after_string <- sprintf(
    "%.2f ± %.2f [%.2f, %.2f]",
    after_slope_mean,
    after_slope_se,
    after_slope_lower,
    after_slope_upper)
  
  
  # --- NEW: markdown-formatted test strings with bold p<0.05 ---
 
  before_term_txt <- fmt_term_plain(before_slope_F, before_slope_Df, before_slope_ResDf, before_slope_P)
  after_term_txt  <- fmt_term_plain(after_slope_F,  after_slope_Df,  after_slope_ResDf,  after_slope_P)
  same_term_txt   <- fmt_term_plain(same_slope_F,   same_slope_Df,   same_slope_ResDf,   same_slope_P)


  
  
  # test summary variables

  test_cols <- data.frame(
    "slope before Mean ± SE [CI]" = before_string,
    "slope before = 0?" = fmt_term(before_slope_F, before_slope_Df, before_slope_ResDf, before_slope_P),
    "slope after Mean ± SE [CI]" = after_string,
    "slope after = 0?" = fmt_term(after_slope_F, after_slope_Df, after_slope_ResDf, after_slope_P),
    "slope before = after?" = fmt_term(same_slope_F, same_slope_Df, same_slope_ResDf, same_slope_P),
    check.names = FALSE, 
    row.names = NULL
  )

  
  # Add the Variables column as the first column
  out_row <- cbind(Variables = labels[i], test_cols)
  
  out_all[[i]] <- out_row
  
  
  bp <- breakpoints[i]      # <-- supply your per-model breakpoint
  bp_se <- breakpoints_se[i]   # <-- supply your per-model breakpoint SE

  plot_files[i] <- save_model_plot_plain(
    mod            = mod,
    label          = labels[i],
    breakpoint     = bp,
    breakpoint_se  = bp_se,
    before_string  = before_string,
    after_string   = after_string,
    before_term_txt = before_term_txt,
    after_term_txt  = after_term_txt,
    same_term_txt   = same_term_txt,
    out_dir        = "plots"    # change if you prefer a different folder
  )
}


# Combine rows from all models into one data frame
out_all_test <- do.call(rbind, out_all)

knitr::kable(out_all_test, escape = FALSE)

plot_files[1]
```
### function model_plot
- raw data, breakpoint (mean and se), predicted lines from linear Hypothesis
```{r}



```
### GRAPH OUTPUT

#### morphology
```{r}
library(patchwork)
morphology_plots <- (plot_scale_g_final$plot | plot_wl_final$plot ) /
              (plot_wingarea_final$plot | plot_wingspan_final$plot) 

morphology_plots
```
#### leg force 
```{r}
leg_force_plots <- (plot_Fy_final$plot | plot_Ftot_final$plot) / (plot_Fx_final$plot | plot_Fz_final$plot | plot_angle_M$plot)
leg_force_plots
```

#### impulse and duration
```{r}
# impulse and duration

impulse_plots <- 
  (plot_impulse_final$plot | plot_impulse_ratio_final$plot | plot_duration_final$plot ) / 
  (plot_left_duration_final$plot | plot_right_duration_final$plot | plot_left_impulse_final$plot | plot_right_impulse_final$plot)                                           

impulse_plots
```

#### trajectory
```{r}

trajectory_plots <- (plot_vx_final$plot | plot_vy_final$plot | plot_vz_final$plot ) /
                              (plot_vres_final$plot | plot_tacc_final) /
                (plot_accx_final$plot | plot_accy_final$plot | plot_accz_final$plot)


trajectory_plots
```

#### tortuosity & wings
```{r}

wavy_plots <- (plot_tr_final$plot | plot_centri_final$plot )/( plot_wb_final$plot | plot_downstroke_final$plot | plot_upstroke_final$plot)

wavy_plots

```



### sample size errors
```{r}

flights_wb_na <- flights_wing %>% filter(is.na(wingbeat_frequency))
flights_wb_na

flights_upstroke_na <- flights_wing %>% filter(is.na(upstroke_amplitude_m))
flights_upstroke_na

flights_downstroke_na <- flights_wing %>% filter(is.na(downstroke_amplitude_m))
flights_downstroke_na

wings_nas <- rbind(flights_wb_na, flights_upstroke_na, flights_downstroke_na) %>% distinct()
wings_nas
```

### Breakpoint pairwise test
```{r}


# Welch pairwise tests from summary stats (means, SE/SD, n) for k variables
# Provide either 'ses' (standard errors) or 'sds' (standard deviations).
# If both provided, 'sds' is used and 'ses' ignored.
welch_pairs_from_summary <- function(means, ns, ses = NULL, sds = NULL,
                                     var_names = NULL, adjust = "BH",
                                     conf_level = 0.95) {
  stopifnot(is.numeric(means), is.numeric(ns))
  k <- length(means)
  if (length(ns) != k) stop("ns must have same length as means.")
  if (is.null(sds) && is.null(ses)) stop("Provide either 'sds' or 'ses'.")
  if (!is.null(sds)) {
    if (length(sds) != k) stop("'sds' must have same length as means.")
  } else {
    if (length(ses) != k) stop("'ses' must have same length as means.")
    sds <- ses * sqrt(ns)  # convert SE to SD
  }
  if (is.null(var_names)) var_names <- paste0("V", seq_len(k))
  if (length(var_names) != k) stop("'var_names' must match length of means.")
  
  # helper for one Welch comparison
  welch_two <- function(m1, s1, n1, m2, s2, n2, conf_level = 0.95) {
    se_diff <- sqrt(s1^2/n1 + s2^2/n2)
    tstat   <- (m1 - m2) / se_diff
    df      <- (s1^2/n1 + s2^2/n2)^2 /
               ((s1^2/n1)^2/(n1-1) + (s2^2/n2)^2/(n2-1))
    p       <- 2 * pt(abs(tstat), df = df, lower.tail = FALSE)
    alpha   <- 1 - conf_level
    tcrit   <- qt(1 - alpha/2, df = df)
    diff    <- (m1 - m2)
    ci_low  <- diff - tcrit * se_diff
    ci_high <- diff + tcrit * se_diff
    list(diff = diff, se_diff = se_diff, t = tstat, df = df, p = p,
         ci_low = ci_low, ci_high = ci_high)
  }
  
  # iterate pairs
  out <- list()
  pmat <- matrix(NA_real_, k, k, dimnames = list(var_names, var_names))
  idx <- 1
  for (i in 1:(k-1)) {
    for (j in (i+1):k) {
      w <- welch_two(means[i], sds[i], ns[i], means[j], sds[j], ns[j],
                     conf_level = conf_level)
      out[[idx]] <- data.frame(
        var1    = var_names[i],
        var2    = var_names[j],
        n1      = ns[i],
        n2      = ns[j],
        mean1   = means[i],
        mean2   = means[j],
        sd1     = sds[i],
        sd2     = sds[j],
        diff    = w$diff,
        se_diff = w$se_diff,
        t       = w$t,
        df      = w$df,
        p       = w$p,
        ci_low  = w$ci_low,
        ci_high = w$ci_high,
        stringsAsFactors = FALSE
      )
      pmat[i, j] <- w$p
      pmat[j, i] <- w$p
      idx <- idx + 1
    }
  }
  res <- do.call(rbind, out)
  res$p_adj <- p.adjust(res$p, method = adjust)
  res <- res[order(res$p_adj, res$var1, res$var2), ]
  rownames(res) <- NULL
  list(table = res, p_matrix = pmat)
}


```
### pairwise test table 
#### all variables vs. all variables
```{r}
# Suppose you have 8 variables:
var_names <- c("mass", "wingarea", "wingspan", "wing loading",
               
               "Fx", "Fy", "Fz", "F_mag", "angle",
               
               "impulse", "impulse ratio", "impulse duration",
               
               "left duration", "right duration", "left impulse", "right impulse",
               
               "vx", "vy", "vz", "v_mag", "atan", "accx", "accy", "accz",
               
               "turn rate", "centri", "wingbeatf", "downstroke", "upstroke")

# Vectors of length 8:
means <- c(breakpoint_sw, breakpoint_wa, breakpoint_ws, breakpoint_wl,
           
           breakpoint_Fx, breakpoint_Fy, breakpoint_Fz, breakpoint_Ftot, breakpoint_angle, 
           
           breakpoint_impulse, breakpoint_impulse_ratio, breakpoint_duration,
           
           breakpoint_left_duration, breakpoint_right_duration, breakpoint_left_impulse, breakpoint_right_impulse, 
           breakpoint_vx, breakpoint_vy, breakpoint_vz, breakpoint_vres, breakpoint_tacc, breakpoint_accx, breakpoint_accy, breakpoint_accz, 
           breakpoint_tr, breakpoint_centri, breakpoint_wb, breakpoint_downstroke, breakpoint_upstroke)



# If you have SEs per variable:
ses   <- c(breakpoint_se_sw, breakpoint_se_wa, breakpoint_se_ws, breakpoint_se_wl,
           
           breakpoint_se_Fx, breakpoint_se_Fy, breakpoint_se_Fz, breakpoint_se_Ftot, breakpoint_se_angle, 
           
           breakpoint_se_impulse, breakpoint_se_impulse_ratio, breakpoint_se_duration,
           
           breakpoint_se_left_duration, breakpoint_se_right_duration, breakpoint_se_left_impulse, breakpoint_se_right_impulse, 
           breakpoint_se_vx, breakpoint_se_vy, breakpoint_se_vz, breakpoint_se_vres, breakpoint_se_tacc, breakpoint_se_accx, breakpoint_se_accy, breakpoint_se_accz, 
           breakpoint_se_tr, breakpoint_se_centri, breakpoint_se_wb, breakpoint_se_downstroke, breakpoint_se_upstroke)

#number of sampels
ns    <- c(184, 184, 184, 184, 
           975, 975, 975, 975, 975, 975, 975, 975, 975, 975, 975, 975,
           918, 918, 918, 918, 918, 918, 918, 918, 
           921, 921, 898, 897, 888)

# If instead you have SDs, set sds=... and omit ses
res <- welch_pairs_from_summary(means = means, ns = ns, ses = ses, var_names = var_names, adjust = "BH", conf_level = 0.95)

# Tidy table:
res$table
# Symmetric matrix of p-values:

```
#### per group
```{r}
groups <- c("morphology", "morphology", "morphology", "morphology", 
            "leg force",  "leg force",  "leg force",  "leg force",  "leg force", 
            "leg impulse", "leg impulse", "leg impulse", 
            "leg impulse", "leg impulse", "leg impulse", "leg impulse",
            "flight speed", "flight speed", "flight speed", "flight speed", 
            "flight speed", "flight speed", "flight speed", "flight speed", 
            "tortuosity", "tortuosity", "flight stability", "flight stability", "flight stability")

df_groups <- data.frame("variable" = var_names, 
                        "group" = groups, 
                        "mean" = means, 
                        "se" = ses, 
                        "n" = ns)

morphology <- subset(df_groups, group == "morphology")
leg_force <- subset(df_groups, group == "leg force")
leg_impulse <- subset(df_groups, group == "leg impulse")
flight_speed <- subset(df_groups, group == "flight speed")
tortuosity <- subset(df_groups, group == "tortuosity")
flight_stability <- subset(df_groups, group == "flight stability")



# If instead you have SDs, set sds=... and omit ses
res_morphology <- welch_pairs_from_summary(means = morphology$mean, ns = morphology$n, ses = morphology$se, var_names = morphology$variable, adjust = "BH", conf_level = 0.95)

res_force <- welch_pairs_from_summary(means = leg_force$mean, ns = leg_force$n, ses = leg_force$se, var_names = leg_force$variable, adjust = "BH", conf_level = 0.95)

res_impulse <- welch_pairs_from_summary(means = leg_impulse$mean, ns = leg_impulse$n, ses = leg_impulse$se, var_names = leg_impulse$variable, adjust = "BH", conf_level = 0.95)

res_speed <- welch_pairs_from_summary(means = flight_speed$mean, ns = flight_speed$n, ses = flight_speed$se, var_names = flight_speed$variable, adjust = "BH", conf_level = 0.95)

res_tort <- welch_pairs_from_summary(means = tortuosity$mean, ns = tortuosity$n, ses = tortuosity$se, var_names = tortuosity$variable, adjust = "BH", conf_level = 0.95)

res_stable <- welch_pairs_from_summary(means = flight_stability$mean, ns = flight_stability$n, ses = flight_stability$se, var_names = flight_stability$variable, adjust = "BH", conf_level = 0.95)
# Tidy table:

print(res_morphology)
print(res_force)
print(res_impulse)
print(res_speed)
print(res_tort)
print(res_stable)
```
#### kable formatting
##### function make_welch_kable()
```{r}


# ---- Packages ----
library(dplyr)
library(knitr)
library(kableExtra)

# ---- Helper: format p_adj per your rules ----
fmt_p <- function(p) {
  ifelse(p < 0.001, 
         "<0.001",
         ifelse(p < 0.01, 
                sprintf("%.3f", p), 
                sprintf("%.2f", p)))
}

# ---- Helper: significance stars column ----
sig_marks <- function(p) {
  dplyr::case_when(
    p < 0.001 ~ "***",
    p >= 0.001 & p < 0.01 ~ "**",
    p >= 0.01  & p < 0.05 ~ "*",
    p >= 0.05  & p < 0.10 ~ "(*)",
    TRUE ~ ""
  )
}

# ---- Function to build kable from your Welch summary df ----
# Expect columns: var1, var2, n1, n2, diff, ci_low, ci_high, t, df, p_adj
make_welch_kable <- function(df, caption = "Welch’s two-sample t-tests (formatted)") {

  out <- df %>%
    dplyr::mutate(
      # round t and df to 1 decimal
      t_fmt  = sprintf("%.1f", t),
      df_fmt = sprintf("%.1f", df),
      # format p_adj per rules
      p_fmt  = fmt_p(p_adj),
      # significance marks
      sig    = sig_marks(p_adj),
      # difference and CI (optional rounding if you wish)
      diff_fmt   = sprintf("%.2f", diff),
      ci_low_fmt = sprintf("%.2f", ci_low),
      ci_high_fmt= sprintf("%.2f", ci_high)
    ) %>%
    dplyr::transmute(
      var1,
      var2,
      n1, n2,
      diff = diff_fmt,
      `95% CI` = paste0("[", ci_low_fmt, ", ", ci_high_fmt, "]"),
      t  = t_fmt,
      df = df_fmt,
      p_adj = p_fmt,
      sig
    )


  knitr::kable(out, align = c("l","l","r","r","r","l","r","r","r","c"),
      caption = caption, booktabs = TRUE) %>%
    kableExtra::kable_styling(full_width = FALSE)
}

# ---- Example usage ----
# Suppose your results are in 'welch_df' (like the table you pasted):
# welch_df <- data.frame(
#   var1 = c("|F|","wingspan","Fz","|F|"),
#   var2 = c("right impulse","|F|","right impulse","impulse"),
#   n1   = c(975,184,975,975),
#   n2   = c(975,975,975,975),
#   diff = c(19.90878623,-17.2,54.70878623,16.1),
#   ci_low = c(14.36656263,-22.41328468,37.24842738,10.88708427),
#   ci_high= c(25.45100983,-11.98671532,72.16914508,21.31291573),
#   t    = c(7.046366303,-6.473305394,6.148495812,6.059314932),
#   df   = c(1475.546117,1140.834289,1017.568493,1222.290473),
#   p_adj= c(1.14043335901699e-09,2.88479250681053e-08,1.51836411805496e-07,1.84498366145796e-07)
# )

# Render the formatted table:
# make_welch_kable(welch_df)

```
##### morphology
```{r}

make_welch_kable(res_morphology$table)

```
##### leg force
```{r}

make_welch_kable(res_force$table)

```
##### leg impulse
```{r}
make_welch_kable(res_impulse$table)

```
##### flight speed
```{r}
make_welch_kable(res_speed$table)

```
##### tortuosity
```{r}
make_welch_kable(res_tort$table)

```
##### stability
```{r}
make_welch_kable(res_stable$table)

```

### p-value matrix
- without wing loading and left duration
```{r}

library(reshape2)

# Example long data
df <- as.data.frame(res$table) %>% 
  dplyr::select(var1, var2, p_adj) %>%
  dplyr:: filter(!var1 %in% c("left duration", "wing loading")) %>%
  dplyr:: filter(!var2 %in% c("left duration", "wing loading")) %>%
  mutate(p_adj = round(p_adj, 3))
# (Optional) melt: here it just confirms the long structure
long <- melt(df, id.vars = c("var1","var2"), measure.vars = "p_adj")

# Cast to a matrix: rows = var1, columns = var2, values = p
p_mat <- acast(long, var1 ~ var2, value.var = "value")  # 'value' is the melted column name
p_mat
#      X     Y
# A 0.10 0.25

```
#### upper triangle
```{r}
# p_mat: matrix produced by acast(var1 ~ var2)
# 1) Fill the upper triangle from the lower triangle when the upper is NA
upper_mask <- upper.tri(p_mat, diag = FALSE)
p_mat[upper_mask] <- ifelse(
  is.na(p_mat[upper_mask]),
  t(p_mat)[upper_mask],  # take the mirrored entry if available
  p_mat[upper_mask]
)

# 2) Drop the lower triangle + diagonal
p_mat[lower.tri(p_mat, diag = TRUE)] <- NA

```
#### bold the values if it is more than 0.05>
```{r}
library(knitr)
library(kableExtra)


p_fmt <- matrix(sprintf("%.2f", p_mat), nrow = nrow(p_mat),
                dimnames = dimnames(p_mat))
mask_upper <- upper.tri(p_mat, diag = FALSE)
bold_mask  <- !is.na(p_mat) & mask_upper & (p_mat > 0.05)
p_fmt[bold_mask] <- paste0("**", p_fmt[bold_mask], "**")
p_fmt[lower.tri(p_mat, diag = TRUE)] <- ""

kable(p_fmt)

```

### NMDS
```{r}


library(readxl)
library(vegan)

# 1) Read
p <- as.data.frame(p_mat)

# 2) Use first column as row names only if it’s non-numeric
if (!is.numeric(p[[1]])) {
  rownames(p) <- p[[1]]
  p <- p[ , -1, drop = FALSE]
}

# 3) To matrix
p <- as.matrix(p)
mode(p) <- "numeric"  # coerce any characters to NA

# 4) Mirror upper triangle to lower (if lower is empty/NA)
#    Copy non-NA from the opposite triangle where NA
p_lower_na <- is.na(p) & lower.tri(p)
p[p_lower_na] <- t(p)[p_lower_na]

p_upper_na <- is.na(p) & upper.tri(p)
p[p_upper_na] <- t(p)[p_upper_na]

# 5) Force symmetry and set diagonal p=1 (same breakpoints with self)
p <- (p + t(p)) / 2
diag(p) <- 1

# 6) Drop variables (rows/cols) that still have NA
keep <- rowSums(is.na(p)) == 0 & colSums(is.na(p)) == 0
if (!all(keep)) {
  p <- p[keep, keep, drop = FALSE]
}

# Final check
if (any(is.na(p))) stop("NA values remain after symmetry fix; please inspect the input matrix.")

# 7) Build dissimilarity: large p = strong association => small distance
D <- 1 - p
diag(D) <- 0

# If all distances are zero (e.g., all p=1), NMDS cannot run
if (!any(D[upper.tri(D)] > 0)) stop("All dissimilarities are zero; NMDS cannot be performed.")

# 8) NMDS
Ddist <- as.dist(D)
nmds <- metaMDS(Ddist, k = 2, autotransform = FALSE, trace = FALSE)

# 9) Plots (minimal)
plot(nmds, type = "t")
stressplot(nmds)


```

#### circle
```{r}


library(readr)
library(dplyr)
library(ggplot2)

# 0) If you don't have 'coords' yet, create it from your nmds object:
coords <- as.data.frame(vegan::scores(nmds, display = "sites"))

# 1) Turn rownames into a 'variable' column so we can join groups
coords <- coords %>%
  tibble::rownames_to_column(var = "variable")

# 2) Load the groups file (variable, group, mean, se, n ...)
df_groups <- read_csv("C:\\Users\\jesu4036\\Documents\\df_groups.csv", show_col_types = FALSE)  # has 'variable' and 'group' columns
# Join groups to the NMDS coordinates (no label cleaning, matches as-is)
df_groups_var <- df_groups %>% dplyr::select(variable, group)
coord_g <- coords %>% left_join(df_groups_var, by = "variable")

# 3) Small helper to make a circle as a data frame
circle_df <- function(cx, cy, r, n = 200) {
  t <- seq(0, 2*pi, length.out = n)
  data.frame(x = cx + r*cos(t), y = cy + r*sin(t))
}

# 4) For each group, compute a circle centered at the group centroid.
#    Radius = max distance from centroid to group members (times a small pad).
pad <- 1.05
circles <- coord_g %>%
  group_by(group) %>%
  filter(!is.na(group)) %>%
  summarise(cx = mean(NMDS1), cy = mean(NMDS2), .groups = "drop") %>%
  rowwise() %>%
  do({
    g <- .$group
    cx <- .$cx; cy <- .$cy
    dists <- coord_g %>% filter(group == g) %>%
      transmute(d = sqrt((NMDS1 - cx)^2 + (NMDS2 - cy)^2))
    r <- ifelse(nrow(dists) > 0, max(dists$d) * pad, 0)
    cbind(group = g, circle_df(cx, cy, r))
  }) %>%
  ungroup()

#ordering the labels
order <- c("morphology", "leg impulse", "leg force",
           "flight speed", "tortuosity", "flight stability")

# Ensure factor order in BOTH data frames
coord_g$group  <- factor(coord_g$group,  levels = order)
circles$group  <- factor(circles$group,  levels = order)


# 5) Plot: points + text + group circles
nmds_plot <- ggplot(coord_g, aes(NMDS1, NMDS2)) +
  # circles under the points
  geom_path(data = circles, aes(x = x, y = y, color = group), inherit.aes = FALSE, linewidth = 0.9, alpha = 0.9) +
  # group fill (optional, light translucent fill behind the circle)
  # geom_polygon(data = circles, aes(x = x, y = y, fill = group), inherit.aes = FALSE, alpha = 0.08, color = NA) +
  geom_point(aes(color = group), size = 2.5) +
  ggrepel::geom_text_repel(aes(label = variable, color = group), size = 3, max.overlaps = Inf, box.padding = 0.2, min.segment.length = 0) +
  coord_equal() +
  theme_minimal(base_size = 11) +
  labs(title = "NMDS with group circles", x = "NMDS1", y = "NMDS2", color = "Group", fill = "Group")


nmds_plot
```
#### ellipse and segment
```{r}

# --- Packages ---
library(readr)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(vegan)

# -----------------------------
# 1) Build dissimilarities and NMDS (your original minimal pipeline)
# -----------------------------

# If p_mat is already in memory (square matrix of p-values), start here:
p <- as.data.frame(p_mat)

# Use first column as row names only if it’s non-numeric
if (!is.numeric(p[[1]])) {
  rownames(p) <- p[[1]]
  p <- p[ , -1, drop = FALSE]
}

# To matrix; coerce to numeric
p <- as.matrix(p)
mode(p) <- "numeric"  # non-numeric entries become NA

# Mirror upper/lower triangles to fill NA, then symmetrize; set diagonal p=1
p_lower_na <- is.na(p) & lower.tri(p)
p[p_lower_na] <- t(p)[p_lower_na]
p_upper_na <- is.na(p) & upper.tri(p)
p[p_upper_na] <- t(p)[p_upper_na]

p <- (p + t(p)) / 2
diag(p) <- 1

# Drop any variables that still have NA
keep <- rowSums(is.na(p)) == 0 & colSums(is.na(p)) == 0
if (!all(keep)) p <- p[keep, keep, drop = FALSE]
if (any(is.na(p))) stop("NA values remain after symmetry fix; please inspect the input matrix.")

# Dissimilarity: large p = strong association => small distance
D <- 1 - p
diag(D) <- 0

# Sanity: must have nonzero dissimilarities
if (!any(D[upper.tri(D)] > 0)) stop("All dissimilarities are zero; NMDS cannot be performed.")

# NMDS
Ddist <- as.dist(D)
set.seed(42)
nmds <- metaMDS(Ddist, k = 2, autotransform = FALSE, trace = FALSE)

# -----------------------------
# 2) Prepare coordinates + join groups
# -----------------------------
coords <- as.data.frame(vegan::scores(nmds, display = "sites"))
coords <- tibble::rownames_to_column(coords, var = "variable")

# Load groups (no label cleaning; match as-is)
df_groups <- read_csv("C:\\Users\\jesu4036\\Documents\\df_groups.csv", show_col_types = FALSE)
df_groups_var <- df_groups %>% dplyr::select(variable, group)

coord_g <- coords %>% left_join(df_groups_var, by = "variable")

# -----------------------------
# 3) Fallback logic:
#    - Ellipses for groups with >=3 points
#    - Segments connecting the two points for groups with exactly 2 points
# -----------------------------
grp_n <- coord_g %>%
  filter(!is.na(group)) %>%
  count(group, name = "n_pts")

have3 <- grp_n %>% filter(n_pts >= 3) %>% pull(group)
seg_groups <- grp_n %>% filter(n_pts == 2) %>% pull(group)

# Build segments (x1,y1 -> x2,y2) for two-point groups
segments_df <- coord_g %>%
  filter(group %in% seg_groups) %>%
  group_by(group) %>%
  summarise(
    x1 = NMDS1[1], y1 = NMDS2[1],
    x2 = NMDS1[2], y2 = NMDS2[2],
    .groups = "drop"
  )

# -----------------------------
# 4) Plot: ellipses (>=3), segments (==2), points & labels
# -----------------------------
nmds_plot <- ggplot(coord_g, aes(NMDS1, NMDS2)) +
  # Ellipse outlines per group (95% normal-theory) — only for groups with >=3 points
  stat_ellipse(
    data = subset(coord_g, group %in% have3),
    aes(group = group, color = group),
    type = "norm", level = 0.68, linewidth = 0.9
  ) +
  # Fallback segments for groups with exactly 2 points
  geom_segment(
    data = segments_df,
    aes(x = x1, y = y1, xend = x2, yend = y2, color = group),
    inherit.aes = FALSE, linewidth = 1
  ) +
  # Points and labels (all groups)
  geom_point(aes(color = group), size = 2.5) +
  ggrepel::geom_text_repel(
    aes(label = variable, color = group),
    size = 3, max.overlaps = Inf,
    box.padding = 0.2, min.segment.length = 0
  ) +
  coord_equal() +
  theme_minimal(base_size = 11) +
  labs(
    title = "", #NMDS with group ellipses (fallback segments for 2-point groups)
    x = "NMDS1", y = "NMDS2", color = "Group"
  )

# Print the plot
#print(nmds_plot)
nmds_plot
# Save to file
ggsave("nmds_group_ellipses.png", nmds_plot, width = 7.5, height = 6, dpi = 200)


```

```{r}

library(vegan)



```



# group-wide comparison
## making a dataframe
```{r}
var_names <- c("mass", "wingarea", "wingspan", "wing loading",
               
               "Fx", "Fy", "Fz", "F_mag", "angle",
               
               "impulse", "impulse ratio", "impulse duration",
               
               "left duration", "right duration", "left impulse", "right impulse",
               
               "vx", "vy", "vz", "v_mag", "atan", "accx", "accy", "accz",
               
               "turn rate", "centri", "wingbeatf", "downstroke", "upstroke")

groups <- c("morphology", "morphology", "morphology", "morphology", 
            "leg force",  "leg force",  "leg force",  "leg force",  "leg force", 
            "leg impulse", "leg impulse", "leg impulse", 
            "leg impulse", "leg impulse", "leg impulse", "leg impulse",
            "flight speed", "flight speed", "flight speed", "flight speed", 
            "flight speed", "flight speed", "flight speed", "flight speed", 
            "tortuosity", "tortuosity", "flight stability", "flight stability", "flight stability")

# Vectors of length 8:
means <- c(breakpoint_sw, breakpoint_wa, breakpoint_ws, breakpoint_wl,
           
           breakpoint_Fx, breakpoint_Fy, breakpoint_Fz, breakpoint_Ftot, breakpoint_angle, 
           
           breakpoint_impulse, breakpoint_impulse_ratio, breakpoint_duration,
           
           breakpoint_left_duration, breakpoint_right_duration, breakpoint_left_impulse, breakpoint_right_impulse, 
           breakpoint_vx, breakpoint_vy, breakpoint_vz, breakpoint_vres, breakpoint_tacc, breakpoint_accx, breakpoint_accy, breakpoint_accz, 
           breakpoint_tr, breakpoint_centri, breakpoint_wb, breakpoint_downstroke, breakpoint_upstroke)



# If you have SEs per variable:
ses   <- c(breakpoint_se_sw, breakpoint_se_wa, breakpoint_se_ws, breakpoint_se_wl,
           
           breakpoint_se_Fx, breakpoint_se_Fy, breakpoint_se_Fz, breakpoint_se_Ftot, breakpoint_se_angle, 
           
           breakpoint_se_impulse, breakpoint_se_impulse_ratio, breakpoint_se_duration,
           
           breakpoint_se_left_duration, breakpoint_se_right_duration, breakpoint_se_left_impulse, breakpoint_se_right_impulse, 
           breakpoint_se_vx, breakpoint_se_vy, breakpoint_se_vz, breakpoint_se_vres, breakpoint_se_tacc, breakpoint_se_accx, breakpoint_se_accy, breakpoint_se_accz, 
           breakpoint_se_tr, breakpoint_se_centri, breakpoint_se_wb, breakpoint_se_downstroke, breakpoint_se_upstroke)

#number of sampels
ns    <- c(184, 184, 184, 184, 
           975, 975, 975, 975, 975, 975, 975, 975, 975, 975, 975, 975,
           918, 918, 918, 918, 918, 918, 918, 918, 
           921, 921, 898, 897, 888)

df_groups <- data.frame("variable" = var_names, 
                        "group" = groups, 
                        "mean" = means, 
                        "se" = ses, 
                        "n" = ns)


```

## dataset
```{r}
df_groups <- read_csv("C:\\Users\\jesu4036\\Documents\\df_groups.csv")

head(df_groups)

```

### all variables 
```{r}
library(metafor)

df_groups$group <- factor(df_groups$group)

m <- rma.mv(
  yi = mean,
  V  = se^2,
  mods = ~ 0 +group,
  method = "REML",# estimate each group’s breakpoint
  random = ~ 1 | variable, # account for differences among variables
  data = df_groups
)

summary(m)
```
#### group-pairwise comparison
```{r}
df_groups$group <- factor(df_groups$group)

# Levels and number of groups
lvls <- levels(df_groups$group)
K <- length(lvls)

# All pairwise indices
pairs <- combn(K, 2)

# Contrast rows: +1 for the first group, -1 for the second
L <- t(apply(pairs, 2, function(idx) {
  v <- numeric(K)
  v[idx[1]] <-  1
  v[idx[2]] <- -1
  v
}))

rownames(L) <- paste(lvls[pairs[1, ]], "-", lvls[pairs[2, ]])

# Wald tests for contrasts
pw <- anova(m, L = L)  # works for rma.mv

pw
```
#### plot
```{r}

# Extract group estimates and SEs
b   <- coef(m)         # named by group
Vb  <- vcov(m)
se  <- sqrt(diag(Vb))

# 95% CI (Wald, z)
z    <- qnorm(0.975)
df_est <- data.frame(
  group = sub("^group", "", names(b)),  # remove "group" prefix if present
  estimate = as.numeric(b),
  se = se,
  ci_lo = b - z * se,
  ci_hi = b + z * se,
  row.names = NULL
)


# Desired order
ord <- c("morphology", "leg impulse", "leg force",
         "flight speed", "tortuosity", "flight stability")

# Ensure 'group' is a factor with that order
df_est$group <- factor(df_est$group, levels = ord)

# Plot (unchanged)
plot_all_bps <- ggplot(df_est, aes(x = group, y = estimate)) +
  geom_hline(yintercept = 0, linetype = 2, color = "grey50") +
  geom_pointrange(aes(ymin = ci_lo, ymax = ci_hi), size = 0.6) +
  labs(x = "", y = "", title = "all breakpoints") +
  theme_minimal(base_size = 14)


#order it on the x-axis as "morphology", "leg impulse", "leg force", "flight speed", "tortuosity", "flight stability"
```
### without left duration and wing loading
```{r}

df_groups2 <- df_groups %>%
  filter(!variable %in% c("wing loading", "left duration"))
df_groups2$group <- factor(df_groups2$group)

m2 <- rma.mv(
  yi = mean,
  V  = se^2,
  mods = ~ 0 +group,
  method = "REML",# estimate each group’s breakpoint
  random = ~ 1 | variable, # account for differences among variables
  data = df_groups2
)

summary(m2)

```
#### group-pair comparison
```{r}

df_groups2$group <- factor(df_groups2$group)

# Levels and number of groups
lvls <- levels(df_groups2$group)
K <- length(lvls)

# All pairwise indices
pairs <- combn(K, 2)

# Contrast rows: +1 for the first group, -1 for the second
L <- t(apply(pairs, 2, function(idx) {
  v <- numeric(K)
  v[idx[1]] <-  1
  v[idx[2]] <- -1
  v
}))

rownames(L) <- paste(lvls[pairs[1, ]], "-", lvls[pairs[2, ]])

# Wald tests for contrasts
pw2 <- anova(m2, L = L)  # works for rma.mv

pw2

```
#### plot
```{r}

# Extract group estimates and SEs
b   <- coef(m2)         # named by group
Vb  <- vcov(m2)
se  <- sqrt(diag(Vb))

# 95% CI (Wald, z)
z    <- qnorm(0.975)
df_est <- data.frame(
  group = sub("^group", "", names(b)),  # remove "group" prefix if present
  estimate = as.numeric(b),
  se = se,
  ci_lo = b - z * se,
  ci_hi = b + z * se,
  row.names = NULL
)


# Desired order
ord <- c("morphology", "leg impulse", "leg force",
         "flight speed", "tortuosity", "flight stability")

# Ensure 'group' is a factor with that order
df_est$group <- factor(df_est$group, levels = ord)

# Plot (unchanged)
plot_only_bps <- ggplot(df_est, aes(x = group, y = estimate)) +
  geom_hline(yintercept = 0, linetype = 2, color = "grey50") +
  geom_pointrange(aes(ymin = ci_lo, ymax = ci_hi), size = 0.6) +
  labs(x = "", y = "", title = "only significant breakpoints") +
  theme_minimal(base_size = 14)

```
### LAST PLOT
```{r}
plot_last <- ggarrange(plot_all_bps, plot_only_bps, nrow = 2, ncol = 1) %>%
  annotate_figure(
    left = text_grob("Estimated breakpoint (± 95% CI)", 
                     rot = 90, size = 14)
  )

plot_last
```
```{r}

plot_only_bps2 <- ggplot(df_est, aes(x = group, y = estimate)) +
  geom_hline(yintercept = 0, linetype = 2, color = "grey50") +
  geom_pointrange(aes(ymin = ci_lo, ymax = ci_hi), size = 0.6) +
  labs(x = "", y = "Estimated breakpoint (± 95% CI)") +
  theme_minimal(base_size = 14)

plot_only_bps2
```

### plot candidates
#### forest 
```{r}


ord <- order(df_groups$mean)
labels_vec <- as.character(df_groups$variable)

# Forest plot with your labels and ordering
forest(
  x      = df_groups$mean[ord],        # yi
  vi     = (df_groups$se[ord])^2,      # sampling variances
  slab   = labels_vec[ord],              # y-axis labels
  #Breakpoint (mean ± 95% CI)
  xlab   = "",
  refline = 0,
  cex = 0.65
)


#forest(m)
```

### plot with sjPlot
```{r}

sjPlot::plot_model(m, type = "est", axis.title = "", axis.labels = labels_vec) 
```
#### installing packages - orchaRd
```{r}

pacman::p_load(devtools, tidyverse, metafor, patchwork, R.rsp, emmeans)

devtools::install_github("daniel1noble/orchaRd", force = TRUE)

```

#### loading libraries 
```{r cars}
library(metafor)
library(orchaRd)
library(RColorBrewer)
library(ggplot2)
```
#### plot with orchard
```{r}

orchaRd::orchard_plot(m, group = "variable", xlab = "Standardised mean difference",
    transfm = "none", twig.size = 0.5, trunk.size = 1) 

```
### ANOSIM
```{r}

library(dplyr)
library(tidyr)
library(vegan)

# --- Helper: align group labels to D in correct order ---
vars_in_D <- rownames(D)

# If 'coord_g' already exists (variable, group, NMDS1, NMDS2), use it; else re-read groups file
if (!exists("coord_g")) {
  df_groups <- readr::read_csv("C:\\Users\\jesu4036\\Documents\\df_groups.csv", show_col_types = FALSE)
  coord_g <- df_groups %>% dplyr::select(variable, group)
}

grp_df <- coord_g %>% dplyr::select(variable, group)
group_vec <- grp_df %>%
  right_join(tibble::tibble(variable = vars_in_D), by = "variable") %>%
  pull(group)

if (any(is.na(group_vec))) {
  stop("Some variables in D are missing a 'group' label. Check df_groups.csv and row/col names of D.")
}

Ddist <- as.dist(D)

# =========================================================
# 1) BETWEEN-GROUP ANOSIM (global + pairwise)
# =========================================================

set.seed(42)
anosim_global <- anosim(Ddist, grouping = group_vec, permutations = 999)
cat("\nGlobal ANOSIM (groups):\n")
print(anosim_global)

anosim_global_tbl <- tibble::tibble(
  method = "ANOSIM (global by groups)",
  R_stat = unname(anosim_global$statistic),
  p_value = unname(anosim_global$signif),
  permutations = attr(anosim_global, "permutations") %||% 999
)

# Pairwise ANOSIM for every pair of groups with ≥2 variables per group
grp_counts <- table(group_vec)
groups_use <- names(grp_counts[grp_counts >= 2])
pairs <- t(combn(groups_use, 2))

pairwise_results <- apply(pairs, 1, function(gp) {
  gA <- gp[1]; gB <- gp[2]
  keep <- group_vec %in% c(gA, gB)
  grp_sub <- droplevels(factor(group_vec[keep]))

  if (min(table(grp_sub)) >= 2) {
    D_sub <- as.dist(as.matrix(Ddist)[keep, keep])
    set.seed(42)
    a <- anosim(D_sub, grouping = grp_sub, permutations = 999)
    data.frame(group_A = gA,
               group_B = gB,
               R_stat = unname(a$statistic),
               p_value = unname(a$signif),
               n_A = as.integer(table(grp_sub)[levels(grp_sub)[1]]),
               n_B = as.integer(table(grp_sub)[levels(grp_sub)[2]]),
               permutations = attr(a, "permutations") %||% 999,
               stringsAsFactors = FALSE)
  } else {
    data.frame(group_A = gA, group_B = gB,
               R_stat = NA_real_, p_value = NA_real_,
               n_A = as.integer(table(grp_sub)[1]),
               n_B = as.integer(table(grp_sub)[2]),
               permutations = NA_integer_, stringsAsFactors = FALSE)
  }
})

pairwise_tbl <- dplyr::bind_rows(pairwise_results) %>%
  arrange(p_value) %>%
  mutate(p_adj_BH = p.adjust(p_value, method = "BH"))

#cat("\nPairwise ANOSIM (between groups):\n")
View(pairwise_tbl)

# =========================================================
# 2) WITHIN-GROUP “BETWEEN-VARIABLES” ANOSIM
#    -> Find clusters of variables within each group (hclust),
#       require replication (≥2 vars per cluster), then run ANOSIM.
# =========================================================

within_group_anosim <- function(Ddist, group_vec, min_k = 2, max_k_try = 5, perms = 999) {
  out <- list()
  gl <- droplevels(factor(group_vec))
  glv <- levels(gl)

  for (g in glv) {
    idx <- which(gl == g)
    if (length(idx) < 3) {
      out[[g]] <- data.frame(
        group = g, k = NA_integer_, R_stat = NA_real_, p_value = NA_real_,
        n_clusters = NA_integer_, cluster_sizes = NA_character_, note = "Fewer than 3 variables",
        stringsAsFactors = FALSE
      )
      next
    }
    # subset distance for variables in group g
    Dg <- as.dist(as.matrix(Ddist)[idx, idx])
    # hierarchical clustering of variables in this group
    hc <- hclust(Dg, method = "average")

    # try k = 2..min(max_k_try, m-1) to get ≥2 clusters with all cluster sizes ≥2
    m <- length(idx)
    k_candidates <- seq(min_k, max(2, min(max_k_try, m - 1)))
    success <- FALSE
    best_k <- NA_integer_; grp_sub <- NULL

    for (k in k_candidates) {
      cl <- cutree(hc, k = k)
      sizes <- table(cl)
      if (length(sizes) >= 2 && min(sizes) >= 2) {
        success <- TRUE
        best_k <- k
        grp_sub <- factor(cl)
        break
      }
    }

    if (!success) {
      out[[g]] <- data.frame(
        group = g, k = NA_integer_, R_stat = NA_real_, p_value = NA_real_,
        n_clusters = NA_integer_, cluster_sizes = NA_character_, note = "No k produced ≥2 clusters with size ≥2",
        stringsAsFactors = FALSE
      )
      next
    }

    set.seed(42)
    a <- anosim(Dg, grouping = grp_sub, permutations = perms)
    out[[g]] <- data.frame(
      group = g, k = best_k,
      R_stat = unname(a$statistic),
      p_value = unname(a$signif),
      n_clusters = nlevels(grp_sub),
      cluster_sizes = paste(as.integer(table(grp_sub)), collapse = ","),
      permutations = attr(a, "permutations") %||% perms,
      note = "ANOSIM between clusters of variables inside the group",
      stringsAsFactors = FALSE
    )
  }
  dplyr::bind_rows(out)
}

wg_tbl <- within_group_anosim(Ddist, group_vec, min_k = 2, max_k_try = 6, perms = 999)
cat("\nWithin-group ANOSIM (between clusters of variables):\n")
print(wg_tbl, row.names = FALSE)

# --- Optional: save results ---
# readr::write_csv(anosim_global_tbl, "anosim_global.csv")
# readr::write_csv(pairwise_tbl, "anosim_pairwise_groups.csv")
# readr::write_csv(wg_tbl, "anosim_within_group_clusters.csv")


```


